\documentclass[10pt,reqno]{amsart}

\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{array}
\usepackage{hyperref}
\usepackage{url}

\hypersetup{
  hidelinks,
  pdftitle={Companion Manuscript: A Scoped Calculus of Equality Traces},
  pdfauthor={Arthur Freitas Ramos, Ruy J. G. B. de Queiroz, Anjolina Grisi de Oliveira, and Dov M. Gabbay}
}
\allowdisplaybreaks
\emergencystretch=3em
\setlength{\abovedisplayskip}{7pt plus 2pt minus 3pt}
\setlength{\belowdisplayskip}{7pt plus 2pt minus 3pt}
\setlength{\abovedisplayshortskip}{4pt plus 2pt}
\setlength{\belowdisplayshortskip}{5pt plus 2pt minus 2pt}

% Archive note: replace or supplement the stable commit citation with the
% future Zenodo archival DOI in the submission version.
\title[Companion: A Scoped Calculus of Equality Traces]
  {Companion Manuscript:\\A Scoped Calculus of Equality Traces:
   Structural Metatheory and a Metadata-Fiber Criterion for
   \texorpdfstring{\(J\)}{J}}

\author{Arthur Freitas Ramos}
\author{Ruy J. G. B. de Queiroz}
\author{Anjolina Grisi de Oliveira}
\author{Dov M. Gabbay}
\renewcommand{\shortauthors}{Ramos et al.}

\subjclass[2020]{03B15, 03B38, 03B70, 18N45}
\keywords{dependent type theory, de Bruijn syntax, definitional equality,
substitution, syntactic quotient, equality traces, path induction}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{warning}[theorem]{Warning}

\newcommand{\Fin}{\mathsf{Fin}}
\newcommand{\Expr}{\mathsf{Expr}}
\newcommand{\Ctx}{\mathsf{Ctx}}
\newcommand{\Ext}[2]{\mathsf{ext}(#1,#2)}
\newcommand{\EmptyCtx}{\mathsf{empty}}
\newcommand{\Ren}{\mathsf{Ren}}
\newcommand{\Sub}{\mathsf{Sub}}
\newcommand{\Var}{\mathsf{var}}
\newcommand{\Sort}{\mathsf{U}}
\newcommand{\PiTy}{\mathsf{Pi}}
\newcommand{\SigmaTy}{\mathsf{Sigma}}
\newcommand{\Lam}{\mathsf{lam}}
\newcommand{\App}{\mathsf{app}}
\newcommand{\Pair}{\mathsf{pair}}
\newcommand{\Fst}{\mathsf{fst}}
\newcommand{\Snd}{\mathsf{snd}}
\newcommand{\NatTy}{\mathsf{Nat}}
\newcommand{\Zero}{\mathsf{zero}}
\newcommand{\Succ}{\mathsf{succ}}
\newcommand{\NatElim}{\mathsf{natElim}}
\newcommand{\CodeNat}{\widehat{\mathsf{Nat}}}
\newcommand{\CodePi}{\widehat{\mathsf{Pi}}}
\newcommand{\CodeSigma}{\widehat{\mathsf{Sigma}}}
\newcommand{\CodeId}{\widehat{\mathsf{Id}}}
\newcommand{\El}{\mathsf{El}}
\newcommand{\IdTy}{\mathsf{Id}}
\newcommand{\Refl}{\mathsf{refl}}
\newcommand{\EqJ}{\mathsf{eqJ}}
\newcommand{\Tr}{\mathsf{Tr}}
\newcommand{\TrStep}{\mathrel{\triangleright_{\mathrm{tr}}}}
\newcommand{\TrEq}{\mathrel{\simeq_{\mathrm{tr}}}}
\newcommand{\Frame}{\mathsf{Frame}}
\newcommand{\IProg}{\mathsf{IProg}}
\newcommand{\ev}{\mathsf{ev}}
\newcommand{\den}[1]{\mathopen{\lbrack\!\lbrack}#1\mathclose{\rbrack\!\rbrack}}
\newcommand{\defeq}{\mathrel{\equiv_{\mathrm d}}}
\newcommand{\red}{\mathrel{\longmapsto}}
\newcommand{\srw}{\mathrel{\rightsquigarrow_{\mathrm s}}}
\newcommand{\srweq}{\mathrel{\simeq_{\mathrm s}}}
\newcommand{\idren}{\mathsf{id}}
\newcommand{\wk}{\mathsf{wk}}
\newcommand{\lift}{\uparrow}
\newcommand{\Q}{\mathcal{Q}}

\begin{document}

\begin{abstract}
We define a raw, scope-indexed dependent-style calculus using de Bruijn
indices.  Its expressions include dependent products and sums,
level-indexed natural-number syntax, Tarski-style codes, identity,
reflexivity, and a two-binder raw \(J\)-constructor.  Contexts are
same-scope assignments, not well-formed telescopes.  Renaming and
simultaneous substitution are defined explicitly under one and two binders.
We prove their identity, composition, lifting, weakening, and instantiation
laws; typed substitution for the raw typing judgment; and preservation of
typing, modulo source definitional equality between displayed type
expressions, for exactly ten top-level primitive redexes.

Source definitional equality is untyped and has no conversion rule.  It is
closed under substitution, so substitution descends to a
representative-independent action on the syntactic quotient that preserves
identity and composition.
An independent source language of identity programs evaluates into
\emph{unlabelled quotient traces}: lists of equality steps over the quotient
together with an endpoint equality.  We give the target trace operations,
one-step rules, congruence, and symmetric-reflexive-transitive closure, and
prove soundness of the source groupoid and finite-frame rewrites.
For quoted expressions, inhabitation of the trace record is equivalent to
source definitional equality.  This is the endpoint exactness of the
syntactic quotient and is tautological once that quotient has been formed.

The conceptual theorem classifies equality refinements.  For
\(R_M(a,b)=\sum_{h:a=b}M(b,h)\), an unrestricted based eliminator satisfying
propositional beta exists exactly when the reflexivity metadata fiber
\(M(a,\mathsf{refl})\) is contractible.  The
proof identifies \(\sum_bR_M(a,b)\) with that fiber.  List-valued trace
metadata has distinct empty and singleton reflexive values, so intrinsic
composability does not yield an unrestricted based eliminator satisfying
propositional beta.  Motives factoring through the equality projection remain
available.  This external theorem is separate from raw
\(J\), which has only its syntactic reflexivity beta rule and no
interpretation theorem for its motive.  The carriers and motive values lie
in a fixed Grothendieck-style universe \(\mathcal U\); the collection of all
\(\mathcal U\)-valued motives and the beta-qualified elimination proposition
lie in a larger \(\mathcal V\).  No judgmental beta is claimed.  The results do not supply a model of
Martin-L\"of or homotopy type theory.
\end{abstract}

\maketitle

\begin{center}
\small
This separately buildable companion preserves the complete raw,
scope-indexed de Bruijn calculus.  The focused theory article is
\texttt{../main.tex}.
\end{center}

\section{Introduction}
\label{sec:introduction}

The conceptual headline is a metadata-fiber classification for
equality refinements.  Alongside it, we
prove structural theorems about a deliberately raw calculus, not an
interpretation theorem.  Fix \(n\in\mathbb N\).  The
expressions \(\Expr_n\) have exactly \(n\) de Bruijn variables and the
constructors listed in Section~\ref{sec:syntax}.  A context is merely a
same-scope assignment
\(\Gamma:\Fin(n)\to\Expr_n\).  There is no context-formation judgment saying
that \(\Gamma\) is a well-formed telescope.  The typing family
\(\Gamma\vdash t:A\) is raw; its definitional equality
\(t\defeq u\) is an untyped relation on expressions.  There is no conversion
rule and no typed definitional-equality judgment.

Let \(\Q_n=\Expr_n/{\defeq}\) be the syntactic quotient and write
\(\den t\) for the class of \(t\).  Over any carrier \(X\) we use the record
\[
 \Tr_X(a,b)
   =\operatorname{List}\!\left(\sum_{x,y:X}(x=y)\right)\times(a=b).
\tag{1.1}\label{eq:intro-trace-record}
\]
When \(X=\Q_n\), this is an \emph{unlabelled quotient trace}.  Its list
contains equality steps over already-quotiented endpoints.  In particular,
a source derivation of \(t\defeq u\) is not retained as a label: it is sent to
the proof-irrelevant quotient equality \(\den t=\den u\).
Throughout, \(X\simeq Y\) denotes specified mutually inverse maps, and
\(\mathsf{transport}_D(e,-)\) denotes equality transport in a family \(D\);
Section~\ref{sec:ambient} states the precise universe conventions and
propositional computation law.

\begin{theorem}[Metadata-fiber criterion, headline form]
\label{thm:intro-j-collapse}
Fix \(a:A\), let
\(M:\prod_{b:A}(a=b)\to\mathcal U\), and choose
\(m_0:M(a,\mathsf{refl}_a)\).  For
\[
 R_M(a,b)=\sum_{h:a=b}M(b,h)
\]
with reflexivity \((\mathsf{refl}_a,m_0)\), the refinement admits an
unrestricted based eliminator satisfying propositional beta if and only if
the metadata fiber \(M(a,\mathsf{refl}_a)\) is contractible.
\end{theorem}

Here \(A\) and all \(M(b,h)\) are \(\mathcal U\)-small, while the proposition
quantifying over all \(\mathcal U\)-valued motives is formed in
\(\mathcal V\).  The beta law is an equality proposition, never a
judgmental equality.

The proof combines the standard equivalence between based elimination and
contractibility of \(\sum_bR_M(a,b)\) with an explicit equivalence
\[
 \sum_{b:A}\sum_{h:a=b}M(b,h)
   \simeq M(a,\mathsf{refl}_a).
\]
The complete argument is Theorem~\ref{thm:metadata-fiber-criterion}.  Empty
and singleton reflexive trace metadata give a noncontractible fiber.

\begin{theorem}[Raw scoped structural results]
\label{thm:intro-main}
For the raw scoped calculus with \(\Pi\), \(\Sigma\), level-indexed natural
syntax, Tarski-style codes, identity, reflexivity, and raw two-binder
\(\EqJ\), the following statements hold.
\begin{enumerate}
\item Renamings \(\Ren(n,m)=\Fin(n)\to\Fin(m)\) and simultaneous
  substitutions
  \(\Sub(n,m)=\Fin(n)\to\Expr_m\) have explicit one- and two-binder lifts.
  Renaming and substitution satisfy identity, composition, weakening, and
  one- and two-variable instantiation equations.
\item A pointwise typed simultaneous substitution preserves the raw typing
  family.  Each of the ten \emph{top-level} primitive computation
  constructors preserves typing modulo \(\defeq\) between the displayed
  source and target type expressions.  There is no contextual or multi-step
  reduction theorem.
\item The untyped relation \(\defeq\) is stable under simultaneous
  substitution and under pointwise \(\defeq\)-equal substitution
  environments.  Consequently every
  \(\sigma:\Sub(n,m)\) induces a representative-independent map
  \(\Q(\sigma):\Q_n\to\Q_m\), and these maps respect substitution identity
  and composition.
\item An independent source family
  \(\IProg_n(t,u)\), generated by \(\defeq\)-atoms, reflexivity, inverse,
  composition, and a finite grammar of same-scope frames, is endpoint-trivial
  and has the expected unique structural-recursion interpretation into any
  target carrying those five operations.  Its quotient-trace instance
  evaluates into \(\Tr_{\Q_n}(\den t,\den u)\).  The named source rewrite equivalence is
  sound for the explicitly defined target trace rewrite equivalence:
  \[
      p\srweq q\quad\Longrightarrow\quad
      \ev(p)\TrEq\ev(q).
  \]
\item Raw \(\EqJ\) has its listed syntactic reflexivity beta constructor.
  This statement is not obtained from, and is not packaged with, any
  external interpretation of its motive.
\end{enumerate}
\end{theorem}

\begin{proposition}[Endpoint exactness of the syntactic quotient]
\label{prop:intro-endpoint}
For \(t,u\in\Expr_n\),
\[
 \operatorname{Nonempty}\bigl(\Tr_{\Q_n}(\den t,\den u)\bigr)
 \quad\Longleftrightarrow\quad t\defeq u.
\tag{1.2}\label{eq:intro-endpoint}
\]
\end{proposition}

This equivalence is tautological after quotient construction.  The forward
direction forgets the list and uses quotient exactness; the reverse direction
uses quotient soundness and forms a singleton trace.  It is included to
identify exactly what endpoint statement is available, not as a semantic
completeness theorem.

\begin{corollary}[Trace-metadata instance]
\label{thm:intro-obstruction}
The record \(\Tr_X\) admits an eliminator through its ambient-equality
component.  It does not admit an unrestricted based eliminator satisfying
propositional beta for arbitrary motives that can observe the list field.
\end{corollary}

The obstruction uses a well-formed singleton reflexive trace, not a malformed
or non-composable list.  Hence imposing intrinsic composability on traces
does not change the classification.  The metadata fiber must be made
contractible, or motives must be restricted through the equality projection.

\subsection*{Contribution as a metatheoretic note}

The contribution is the combination of five sharply bounded pieces:
binder-correct raw scoped syntax; source-derived preservation for ten
primitive redexes; representative-independent and compositional quotient
substitution; source/target-separated identity programs; and the
metadata-fiber criterion applied to computational trace records.  The
equivalence between a based eliminator with propositional beta and
contractibility of the based total space is standard background.  The
explicit reduction to the reflexivity metadata fiber, its trace application,
and the binder-correct raw interface are the contribution of this note.
Source identity
rewrite soundness is a homomorphic simulation, largely forced by the
structural-recursion interpretation once the target operations and their
laws are supplied.  Endpoint exactness is tautological after quotienting.
The note claims no absolute priority; its point is the combined interface and
the trace-metadata application.

\subsection*{Status of the calculus}

The terminology in Theorem~\ref{thm:intro-main} is intentional.  We study a
\emph{scoped dependent-style calculus}, not a presentation of full
Martin-L\"of type theory.  We construct only a syntactic quotient; no typed
model or initial CwF is supplied.  We prove primitive-redex preservation,
not standard contextual or multi-step subject reduction.  Source identity
programs have no typing judgment and are endpoint-trivial in the sense of
Proposition~\ref{prop:iprogram-endpoints}: an atom already requires
\(t\defeq u\).  There is no normalization, canonicity,
decidable type checking, univalence, higher inductive type, or completeness
theorem for source identity rewrites.

\subsection*{Relation to earlier computational paths}

The prior computational-path articles
\cite{deQueirozEtAl2016,RamosEtAl2017,RamosEtAl2018} and the 2026 monograph
\cite{RamosEtAl2026} study label-sensitive paths: equality reasons are
first-class labels and their rewrite calculus preserves that intentional
presentation.  The target in this paper is weaker.  A
\(\defeq\)-derivation is used only to obtain equality in \(\Q_n\); proof
irrelevance then erases which derivation was chosen.  Evaluated identity
programs are therefore trace-decorated paths over quotient-collapsed
endpoints, not carriers of the original computational reason.  This change
of target is central and is compared theorem by theorem in
Section~\ref{sec:comparison}.

\subsection*{Organization}

Section~\ref{sec:layers} separates the equality layers.
Sections~\ref{sec:syntax} and~\ref{sec:substitution} give the raw syntax and
its structural algebra.  Section~\ref{sec:judgments} presents typing,
computation, substitution, and primitive-redex preservation.  The syntactic
quotient is constructed in Section~\ref{sec:syntactic-quotient}.  Source identity programs
and their trace semantics occupy Section~\ref{sec:identity-programs}.
Section~\ref{sec:example} gives lambda, dependent-projection, and
natural-successor examples.  Section~\ref{sec:main-result} collects the raw
structural theorem, while Section~\ref{sec:j-boundary} separately proves the
external elimination obstruction.  Sections~\ref{sec:family-semantics} and
\ref{sec:comparison} give an independent fixed-universe algebra and the
comparison with prior work.
Appendices~\ref{app:grammar} and~\ref{app:rewrites} collect the raw rules.

\section{The equality layers}
\label{sec:layers}

\subsection{Ambient metatheory}
\label{sec:ambient}

Fix once and for all a Grothendieck-style universe \(\mathcal U\) of small
sets.  Every carrier, expression family, metadata fiber, and quotient used
below is an element of \(\mathcal U\).  The universe is closed under finite
lists, dependent sums and products indexed by small sets, equality
propositions, and quotients by equivalence relations.  We reason in ordinary
extensional mathematics about these sets.  Concretely:
\begin{enumerate}
\item equality has its usual eliminator and equality propositions are proof
  irrelevant:
  \(h,k:x=y\) implies \(h=k\);
\item functions are extensional, so pointwise equal functions are equal;
\item finite lists with append, reversal, map, and no-confusion
  \([]\neq z::L\);
\item every equivalence relation \(E\) on a small set \(X\) has an effective
  small quotient
  \(\eta:X\to X/E\) with
  \[
       E(x,y)\Longrightarrow\eta x=\eta y
       \quad\text{and}\quad
       (\eta x=\eta y)\Longrightarrow E(x,y).
  \tag{2.1}\label{eq:quotient-sound-exact}
  \]
\end{enumerate}
The first implication in~\eqref{eq:quotient-sound-exact} is quotient
soundness; the second is quotient exactness.

For small sets \(X,Y\), the notation \(X\simeq Y\) means specified maps
\[
 F:X\to Y,\qquad G:Y\to X
\]
together with propositional equalities
\(F(Gy)=y\) and \(G(Fx)=x\).  If \(D:X\to\mathcal U\) and \(e:x=y\), then
\(\mathsf{transport}_D(e,-):D(x)\to D(y)\) denotes equality transport.
We use only its propositional reflexivity law
\[
 \mathsf{transport}_D(\mathsf{refl}_x,z)=z.
\tag{2.2}\label{eq:transport-beta}
\]
No judgmental computation rule is part of the ambient assumptions.

When we quantify over the collection of all \(\mathcal U\)-valued motives
\(C\), that collection is regarded as living in one fixed larger universe
\(\mathcal V\).  Each individual motive value, equality proposition,
\(\IProg\)-fiber, \(R\)-fiber, and metadata fiber remains
\(\mathcal U\)-small.  Thus the phrase ``admits an unrestricted based
eliminator satisfying propositional beta'' means an eliminator into every
\(\mathcal U\)-valued motive together with that beta family; it makes no
impredicative claim that
the collection of all such motives lies in \(\mathcal U\) itself.
For \(X\in\mathcal U\) and \(c\in X\), the sets
\(\mathsf{Contr}_c(X)\) and \(\mathsf{isContr}(X)\) are
\(\mathcal U\)-small.  By contrast, the set of all
\(\mathcal U\)-valued motives on \(X\), the data of an eliminator over all of
them, and the proposition asserting existence of such data lie in
\(\mathcal V\).  Each individual propositional beta equation lies in the
small equality proposition of a fiber \(C(x)\in\mathcal U\).

These assumptions have sharply delimited uses.  Function extensionality
turns the pointwise lift and environment calculations of
Sections~\ref{sec:substitution} and~\ref{sec:syntactic-quotient} into
equalities of functions.  Quotient soundness defines quotient substitution
and the trace attached to a source atom.  Quotient exactness is used only in
the map from trace inhabitation back to source \(\defeq\) in endpoint
exactness.  Proof irrelevance erases the
choice of source \(\defeq\)-derivation after quotient soundness and makes the
external equality-component eliminator insensitive to equality proofs.
List no-confusion proves that the empty and singleton traces differ.
Lean is one realization of this metatheory for reproducibility; it is not the
definition of the mathematics.

\subsection{The dependency layers}

The word ``equality'' occurs at several levels.  The direction of dependence
matters: none of the target trace relations is used to generate a source
typing or definitional-equality derivation.

\begin{table}[ht]
\centering
\small
\renewcommand{\arraystretch}{1.22}
\begin{tabular}{>{\raggedright\arraybackslash}p{0.20\textwidth}
                >{\centering\arraybackslash}p{0.19\textwidth}
                >{\raggedright\arraybackslash}p{0.48\textwidth}}
\toprule
Layer & Notation & Information and role\\
\midrule
Metatheoretic equality
  & \(x=y\)
  & Proof-irrelevant equality used to form quotient classes and the endpoint
    field of a trace record.\\
Primitive source computation
  & \(t\red u\)
  & Exactly one of ten top-level constructors.  There is no contextual or
    multi-step closure.\\
Source definitional equality
  & \(t\defeq u\)
  & Equivalence and enumerated compatible closure of primitive computation.
    It is untyped and defines the syntactic quotient.\\
Syntactic-quotient equality
  & \(\den t=\den u\)
  & Ambient equality in \(\Q_n=\Expr_n/{\defeq}\).  Quotient exactness
    identifies this precisely with \(t\defeq u\).\\
Source identity program
  & \(p:\IProg_n(t,u)\)
  & Explicit syntax generated by source atoms, unit, inverse, composition,
    and finite same-scope frames.\\
Unlabelled quotient trace
  & \(T:\Tr_{\Q_n}(\den t,\den u)\)
  & A list of quotient-equality steps plus an endpoint equality.  Source
    \(\defeq\)-derivation labels have been erased.\\
Source identity rewrite
  & \(p\srweq q\)
  & Symmetric reflexive-transitive closure of the named source program laws.\\
Target trace rewrite
  & \(T\TrEq U\)
  & Explicit symmetric-reflexive-transitive trace equivalence.  Soundness maps the
    preceding row to this one; no reflection theorem is asserted.\\
\bottomrule
\end{tabular}
\caption{The layered equality discipline.}
\label{tab:equality-layers}
\end{table}

There are two essential arrows.  The first is quotient soundness
\[
       (t\defeq u)\longmapsto(\den t=\den u).
\]
The second evaluates source identity syntax:
\[
       \ev:\IProg_n(t,u)\longrightarrow
       \Tr_{\Q_n}(\den t,\den u).
\]
The first arrow erases source evidence: quotient soundness returns only an
ambient equality proof, and proof irrelevance identifies different proofs of
the same quotient equality.  The second arrow consequently cannot preserve a
\(\defeq\)-derivation as a label.

\begin{remark}[Scope of the no-target-leakage statement]
The source constructors in Definition~\ref{def:identity-programs} consume
only source \(\defeq\)-derivations and source frames.  The statement applies
to that family alone (named \texttt{RawMLTT.IdentityExpr} in the
accompanying software).  The broader software repository also contains
convenience syntax that starts from already constructed target trace records; that
syntax is excluded from the calculus and from
Theorem~\ref{thm:raw-structural}.
\end{remark}

\section{An independently scoped source calculus}
\label{sec:syntax}

\subsection{Scopes and raw expressions}

Following de Bruijn's nameless representation
\cite{deBruijn1972}, let
\(\Fin(n)=\{0,\ldots,n-1\}\), with \(0\) the most recently bound
variable.  The family \(\Expr_n\) is defined inductively by
\begin{align*}
t,u,A,B,M ::= {}&
  \Var(i) \mid \Sort_\ell
  \mid \PiTy(A,B) \mid \SigmaTy(A,B)
  \mid \Lam(t) \mid \App(t,u)\\
 &\mid \Pair(t,u)\mid\Fst(t)\mid\Snd(t)
  \mid \NatTy_\ell\mid\Zero_\ell\mid\Succ_\ell(t)\\
 &\mid \NatElim_\ell(M,z,s,t)
  \mid \CodeNat_\ell\mid\CodePi(A,B)
  \mid\CodeSigma(A,B)\mid\CodeId(A,t,u)\\
 &\mid \El(A)\mid\IdTy(A,t,u)\mid\Refl(t)
  \mid \EqJ(M,r,b,p).
\end{align*}
Here \(i\in\Fin(n)\) and \(\ell\in\mathbb{N}\).  The displayed grammar
suppresses scope indices, which are part of the inductive definition:
\[
\begin{array}{c|c}
\text{constructor} & \text{scopes of arguments}\\ \hline
\PiTy,\SigmaTy,\CodePi,\CodeSigma
  & \Expr_n\times\Expr_{n+1}\to\Expr_n\\
\Lam & \Expr_{n+1}\to\Expr_n\\
\NatElim_\ell
  & \Expr_{n+1}\times\Expr_n\times\Expr_{n+2}
      \times\Expr_n\to\Expr_n\\
\EqJ
  & \Expr_{n+2}\times\Expr_n\times\Expr_n\times\Expr_n
      \to\Expr_n.
\end{array}
\]
All other arguments remain in scope \(n\).  Thus the codomains of
\(\PiTy\), \(\SigmaTy\), and their codes bind one variable; the successor
branch of \(\NatElim\) binds a predecessor and an induction hypothesis; and
the motive of \(\EqJ\) binds an endpoint and an identity proof.

A raw context of length \(n\) is a function
\[
                       \Gamma:\Fin(n)\longrightarrow\Expr_n.
\]
This is a \emph{same-scope assignment}: every declaration is already written
in the full scope \(n\).  There is no judgment
\(\vdash\Gamma\;\mathsf{ctx}\), no inductively generated telescope, and no
theorem relating these assignments to well-formed MLTT contexts.  The raw
typing constructors use whichever declarations they require.  The empty
assignment is
\(\EmptyCtx_0:\Fin(0)\to\Expr_0\).  Extension is the explicit operation
\[
 \Ext{\Gamma}{A}\in\Ctx(n+1)
 \quad(\Gamma\in\Ctx(n),\ A\in\Expr_n)
\]
that weakens every declaration:
\[
\begin{split}
\Ext{\Gamma}{A}(0)&=A[\wk_n]_{\mathrm r},\\
\Ext{\Gamma}{A}(i+1)&=\Gamma(i)[\wk_n]_{\mathrm r}.
\end{split}
\tag{3.1}\label{eq:context-extension}
\]

\subsection{Levels, natural numbers, and codes}

Levels are natural numbers internal to the grammar only as parameters.
The rules use
\[
              \Gamma\vdash\Sort_\ell:\Sort_{\ell+1},
   \qquad
              \Gamma\vdash\NatTy_\ell:\Sort_\ell.
\]
Natural numbers, zero, successor, and their recursor all carry the same
level parameter.  This indexing is not cosmetic: it makes
\[
                 \El(\CodeNat_\ell)\red\NatTy_\ell
\]
type preserving at every level.  There is no cumulativity or resizing rule.

The code constructors are Tarski-style in the limited sense relevant here:
codes are explicit syntax and are decoded by \(\El\).  A code at level
\(\ell\) has type \(\Sort_\ell\), and its decoding also has type
\(\Sort_\ell\).  Product and sum codes bind over the decoding of their domain
code.  The calculus does not purport to give a full hierarchy of inductive
universes or a universe-polymorphic elaboration algorithm.

\subsection{\texorpdfstring{The raw two-binder \(J\)-constructor}
{The raw two-binder J-constructor}}

Suppose \(A,a\in\Expr_n\).  Define the proof declaration in endpoint scope by
\[
 \mathsf{IdProof}_n(A,a)
   \coloneqq
   \IdTy\bigl(A[\wk_n]_{\mathrm r},
              a[\wk_n]_{\mathrm r},\Var(0)\bigr)
   \in\Expr_{n+1}.
\tag{3.2}\label{eq:idproof-definition}
\]
The endpoint is the new variable of \(\Ext{\Gamma}{A}\).  Extending this
assignment once more gives the explicit motive assignment
\[
 \Ext{\Ext{\Gamma}{A}}{\mathsf{IdProof}_n(A,a)}
   \in\Ctx(n+2).
\tag{3.3}\label{eq:motive-context}
\]
Because the latest variable is numbered \(0\), a motive
\(M\in\Expr_{n+2}\) sees the proof at index \(0\), the endpoint at index
\(1\), and the ambient variables from index \(2\) onward.  The raw term
\(\EqJ(M,r,b,p)\) instantiates those two variables by \(p\) and \(b\).
Its only computation constructor is the syntactic rule
\[
                 \EqJ(M,r,a,\Refl(a))\red r.
\]
Nothing in the raw syntax says that \(M\) factors through an ambient equality
component, and this paper proves no such interpretation.  The independent
external eliminator of Section~\ref{sec:j-boundary} is therefore a separate
construction, not a semantics for this raw motive.

\section{Renaming and simultaneous substitution}
\label{sec:substitution}

\subsection{Renamings and binder lift}

Set
\[
                  \Ren(n,m)\coloneqq\Fin(n)\to\Fin(m).
\]
We write \(\rho:n\rightsquigarrow_{\mathrm r}m\) only when
\(\rho\in\Ren(n,m)\).  Identity and composition are
\[
  \idren^{\Ren}_n(i)=i,\qquad
  (\rho;_{\Ren}\tau)(i)=\tau(\rho(i)),
\]
where \(\rho\in\Ren(n,m)\) and \(\tau\in\Ren(m,k)\).
Weakening \(\wk_n\in\Ren(n,n+1)\) is \(\wk_n(i)=i+1\).  The binder lift
\(\rho^\lift\in\Ren(n+1,m+1)\) is
\[
        \rho^\lift(0)=0,\qquad
        \rho^\lift(i+1)=\rho(i)+1.
\tag{4.1}\label{eq:ren-lift}
\]
Renaming \(t[\rho]_{\mathrm r}\) is structural recursion, using one lift
under one binder and two lifts under the \(\NatElim\) successor branch and
the \(\EqJ\) motive.

\begin{proposition}[Renaming algebra]
\label{prop:renaming}
For every expression and every composable pair of renamings,
\[
\begin{aligned}
t[\idren^{\Ren}_n]_{\mathrm r}&=t,\\
t[\rho]_{\mathrm r}[\tau]_{\mathrm r}
  &=t[\rho;_{\Ren}\tau]_{\mathrm r},\\
(\idren^{\Ren}_n)^\lift&=\idren^{\Ren}_{n+1},\\
(\rho;_{\Ren}\tau)^\lift&=\rho^\lift;_{\Ren}\tau^\lift,\\
\wk_n;_{\Ren}\rho^\lift&=\rho;_{\Ren}\wk_m.
\end{aligned}
\]
\end{proposition}

\begin{proof}
Prove the three equations between renamings first.  At index \(0\), both
\((\idren^{\Ren}_n)^\lift\) and
\((\rho;_{\Ren}\tau)^\lift\) return \(0\).  At \(i+1\),
\[
 (\idren^{\Ren}_n)^\lift(i+1)=i+1,
\qquad
 (\rho;_{\Ren}\tau)^\lift(i+1)
   =\tau(\rho(i))+1
   =(\rho^\lift;_{\Ren}\tau^\lift)(i+1).
\]
For \(i:\Fin(n)\),
\[
 (\wk_n;_{\Ren}\rho^\lift)(i)=\rho(i)+1
   =(\rho;_{\Ren}\wk_m)(i).
\]
Function extensionality gives the three renaming equalities.

Now induct structurally on \(t\), proving identity and composition
simultaneously.  The variable case is the corresponding equation for
\(\rho\); \(\Sort_\ell,\NatTy_\ell,\Zero_\ell,\CodeNat_\ell\) are fixed.
For a one-binder constructor, for example,
\[
\begin{aligned}
\PiTy(A,B)[\rho]_{\mathrm r}[\tau]_{\mathrm r}
 &=\PiTy(A[\rho]_{\mathrm r}[\tau]_{\mathrm r},
          B[\rho^\lift]_{\mathrm r}[\tau^\lift]_{\mathrm r})\\
 &=\PiTy(A[\rho;_{\Ren}\tau]_{\mathrm r},
          B[(\rho;_{\Ren}\tau)^\lift]_{\mathrm r}),
\end{aligned}
\]
using the two induction hypotheses and lift composition.
The same calculation covers \(\SigmaTy,\CodePi,\CodeSigma\); \(\Lam\)
uses only the bound induction hypothesis.  Application, pairing,
projections, successor, decoding, identity, code identity, and reflexivity
use the induction hypotheses componentwise.

For the two-binder constructors the recursive equations display all lifts:
\[
\begin{aligned}
&\NatElim_\ell(M,z,s,k)[\rho]_{\mathrm r}[\tau]_{\mathrm r}\\
&=\NatElim_\ell(
 M[\rho^\lift]_{\mathrm r}[\tau^\lift]_{\mathrm r},
 z[\rho]_{\mathrm r}[\tau]_{\mathrm r},
 s[\rho^{\lift\lift}]_{\mathrm r}
   [\tau^{\lift\lift}]_{\mathrm r},
 k[\rho]_{\mathrm r}[\tau]_{\mathrm r})\\
&=\NatElim_\ell(
 M[(\rho;_{\Ren}\tau)^\lift]_{\mathrm r},
 z[\rho;_{\Ren}\tau]_{\mathrm r},
 s[(\rho;_{\Ren}\tau)^{\lift\lift}]_{\mathrm r},
 k[\rho;_{\Ren}\tau]_{\mathrm r}),
\end{aligned}
\]
and
\[
\begin{aligned}
&\EqJ(M,r,b,p)[\rho]_{\mathrm r}[\tau]_{\mathrm r}\\
&=\EqJ(
 M[\rho^{\lift\lift}]_{\mathrm r}
   [\tau^{\lift\lift}]_{\mathrm r},
 r[\rho]_{\mathrm r}[\tau]_{\mathrm r},
 b[\rho]_{\mathrm r}[\tau]_{\mathrm r},
 p[\rho]_{\mathrm r}[\tau]_{\mathrm r})\\
&=\EqJ(
 M[(\rho;_{\Ren}\tau)^{\lift\lift}]_{\mathrm r},
 r[\rho;_{\Ren}\tau]_{\mathrm r},
 b[\rho;_{\Ren}\tau]_{\mathrm r},
 p[\rho;_{\Ren}\tau]_{\mathrm r}).
\end{aligned}
\]
The identity proof is the same induction with
\((\idren^{\Ren})^\lift=\idren^{\Ren}\), used twice in these final two
cases.  These are all constructors of \(\Expr\).
\end{proof}

\subsection{Simultaneous substitutions}

Set
\[
       \Sub(n,m)\coloneqq\Fin(n)\to\Expr_m.
\]
The embedding of renamings as variable-only substitutions is
\[
 \mathsf{ofRen}_{n,m}:\Ren(n,m)\to\Sub(n,m),\qquad
 \mathsf{ofRen}(\rho)(i)=\Var(\rho(i)).
\tag{4.2}\label{eq:of-ren}
\]
The substitution identity is not the renaming identity:
\[
       \idren^{\Sub}_n\coloneqq
       \mathsf{ofRen}(\idren^{\Ren}_n)\in\Sub(n,n).
\]
For \(\sigma\in\Sub(n,m)\), its lift lies in
\(\Sub(n+1,m+1)\) and is
\[
\sigma^\lift(0)=\Var(0),\qquad
\sigma^\lift(i+1)=\sigma(i)[\wk_m]_{\mathrm r}.
\tag{4.3}\label{eq:sub-lift}
\]
Capture-avoiding substitution \(t[\sigma]\) is again structural recursion,
using the same one- and two-lift pattern as renaming.  Composition is
\[
 \sigma;_{\Sub}\tau\in\Sub(n,k),\qquad
 (\sigma;_{\Sub}\tau)(i)=\sigma(i)[\tau],
\tag{4.4}\label{eq:sub-composition}
\]
for \(\sigma\in\Sub(n,m)\) and \(\tau\in\Sub(m,k)\).

The distinction between pre- and post-renaming is useful in the proof.
Writing \(\rho^\ast\sigma=\sigma\circ\rho\) and
\(\rho_\ast\sigma=(i\mapsto\sigma(i)[\rho]_{\mathrm r})\), one obtains the
fusion laws
\[
\begin{aligned}
t[\sigma][\rho]_{\mathrm r}&=t[\rho_\ast\sigma],\\
t[\rho]_{\mathrm r}[\sigma]&=t[\rho^\ast\sigma].
\end{aligned}
\tag{4.5}\label{eq:fusion}
\]
Their binder cases depend on the naturality square in
Proposition~\ref{prop:renaming}.

\begin{theorem}[Substitution algebra]
\label{thm:substitution-algebra}
For \(t\in\Expr_n\), \(\sigma\in\Sub(n,m)\), and
\(\tau\in\Sub(m,k)\), and \(\upsilon\in\Sub(k,l)\),
\[
\begin{aligned}
t[\idren^{\Sub}_n]&=t,\\
t[\sigma][\tau]&=t[\sigma;_{\Sub}\tau],\\
(\idren^{\Sub}_n)^\lift&=\idren^{\Sub}_{n+1},\\
(\sigma;_{\Sub}\tau)^\lift&=\sigma^\lift;_{\Sub}\tau^\lift.
\end{aligned}
\tag{4.6}\label{eq:sub-algebra}
\]
The substitution environments themselves satisfy
\[
\begin{aligned}
\idren^{\Sub}_n;_{\Sub}\sigma&=\sigma,&
\sigma;_{\Sub}\idren^{\Sub}_m&=\sigma,\\
(\sigma;_{\Sub}\tau);_{\Sub}\upsilon
 &=\sigma;_{\Sub}(\tau;_{\Sub}\upsilon).
\end{aligned}
\tag{4.7}\label{eq:sub-environment-algebra}
\]
Moreover
\[
                 t[\mathsf{ofRen}(\rho)]=t[\rho]_{\mathrm r},
\tag{4.8}\label{eq:ofren-subst}
\]
and the fusion equations~\eqref{eq:fusion} hold with their displayed scope
types.
\end{theorem}

\begin{proof}
We establish the lift and fusion equations before expression composition.
First,
\[
 \mathsf{ofRen}(\rho)^\lift=\mathsf{ofRen}(\rho^\lift)
\tag{4.9}\label{eq:ofren-lift}
\]
pointwise: both sides return \(\Var(0)\) at \(0\), and
\(\Var(\rho(i)+1)\) at \(i+1\).  Structural induction on \(t\), using
\eqref{eq:ofren-lift} once or twice under binders, proves
\eqref{eq:ofren-subst}.

Next prove the two fusion equations~\eqref{eq:fusion} by structural induction
on \(t\).  Their variable cases are the definitions of
\(\rho_\ast\sigma\) and \(\rho^\ast\sigma\).  Under one binder the required
environment equalities are, pointwise,
\[
\begin{aligned}
(\rho_\ast\sigma)^\lift(0)&=\Var(0),\\
(\rho_\ast\sigma)^\lift(i+1)
 &=\sigma(i)[\rho]_{\mathrm r}[\wk]_{\mathrm r}
  =\sigma(i)[\wk]_{\mathrm r}[\rho^\lift]_{\mathrm r},\\
(\rho^\ast\sigma)^\lift(0)&=\Var(0),\\
(\rho^\ast\sigma)^\lift(i+1)
 &=\sigma(\rho(i))[\wk]_{\mathrm r}
  =\sigma^\lift(\rho^\lift(i+1)).
\end{aligned}
\]
The middle equality in the first pair is the weakening naturality equation
of Proposition~\ref{prop:renaming}.  The two-binder constructors apply these
equalities twice.  All nonbinding constructors follow componentwise.

The identity lift is also pointwise:
\[
(\idren^{\Sub}_n)^\lift(0)=\Var(0),\qquad
(\idren^{\Sub}_n)^\lift(i+1)=\Var(i+1),
\]
so it is \(\idren^{\Sub}_{n+1}\).
For lift composition, both sides return \(\Var(0)\) at \(0\).  At \(i+1\),
\[
\begin{aligned}
(\sigma^\lift;_{\Sub}\tau^\lift)(i+1)
 &=\bigl(\sigma(i)[\wk_m]_{\mathrm r}\bigr)[\tau^\lift]\\
 &=\bigl(\sigma(i)[\tau]\bigr)[\wk_k]_{\mathrm r}\\
 &=(\sigma;_{\Sub}\tau)^\lift(i+1),
\end{aligned}
\tag{4.10}\label{eq:lift-composition-proof}
\]
where the middle step is the two fusion equations, equivalently the
imagewise square
\[
 \mathsf{ofRen}(\wk_m);_{\Sub}\tau^\lift
 =\tau;_{\Sub}\mathsf{ofRen}(\wk_k).
\]

We can now prove expression substitution.  Identity follows from
\eqref{eq:ofren-subst} and renaming identity.  For composition, induct
structurally on \(t\).  Variables give the definition of
\(\sigma;_{\Sub}\tau\); constants are fixed; all nonbinding constructors
use their induction hypotheses componentwise.  The one-binder cases use
\eqref{eq:lift-composition-proof}.  For the two-binder arguments,
\[
 M[\sigma^{\lift\lift}][\tau^{\lift\lift}]
  =M[\sigma^{\lift\lift};_{\Sub}\tau^{\lift\lift}]
  =M[(\sigma;_{\Sub}\tau)^{\lift\lift}],
\tag{4.11}\label{eq:eqj-double-lift-composition}
\]
and likewise for the successor branch \(s\) of \(\NatElim\).
Thus the \(\NatElim\) motive uses one lift, its successor branch two, and its
other arguments none; the \(\EqJ\) motive uses two lifts and its other
arguments none.  This exhausts the binding constructors.

Only after expression composition is established do we derive
\eqref{eq:sub-environment-algebra}.  At \(i\), left identity is
\(\Var(i)[\sigma]=\sigma(i)\), right identity is
\(\sigma(i)[\idren^{\Sub}_m]=\sigma(i)\), and associativity is
\[
 ((\sigma;_{\Sub}\tau);_{\Sub}\upsilon)(i)
 =(\sigma(i)[\tau])[\upsilon]
 =\sigma(i)[\tau;_{\Sub}\upsilon]
 =(\sigma;_{\Sub}(\tau;_{\Sub}\upsilon))(i).
\]
Function extensionality yields the environment equalities.  There is
therefore no use of environment associativity in the proof of expression
composition.
\end{proof}

\subsection{Weakening and instantiation}

For \(a\in\Expr_n\), let
\(\langle a\rangle\in\Sub(n+1,n)\) send \(0\) to \(a\) and \(i+1\) to
\(\Var(i)\).  One-variable instantiation is
\[
                     B\{a\}\coloneqq B[\langle a\rangle].
\]
For \(M\in\Expr_{n+2}\) and \(b,p\in\Expr_n\), let
\(\langle b,p\rangle\in\Sub(n+2,n)\) send \(0\) to \(p\), \(1\) to \(b\),
and \(i+2\) to \(\Var(i)\).  Then
\[
                     M\{b,p\}\coloneqq M[\langle b,p\rangle].
\]
The order reflects de Bruijn extension: the proof is newest.

\begin{proposition}[Instantiation laws]
\label{prop:instantiation}
Weakening is cancelled by one-variable instantiation,
\[
                   t[\wk]\{a\}=t,
\]
and by two-variable instantiation,
\[
              t[\wk][\wk]\{b,p\}=t.
\]
Renaming and substitution commute with both forms:
\[
\begin{aligned}
B\{a\}[\rho]_{\mathrm r}
 &=B[\rho^\lift]_{\mathrm r}\{a[\rho]_{\mathrm r}\},\\
B\{a\}[\sigma]
 &=B[\sigma^\lift]\{a[\sigma]\},\\
M\{b,p\}[\rho]_{\mathrm r}
 &=M[\rho^{\lift\lift}]_{\mathrm r}
       \{b[\rho]_{\mathrm r},p[\rho]_{\mathrm r}\},\\
M\{b,p\}[\sigma]
 &=M[\sigma^{\lift\lift}]\{b[\sigma],p[\sigma]\}.
\end{aligned}
\tag{4.12}\label{eq:instantiation-laws}
\]
\end{proposition}

\begin{proof}
Expand instantiation as simultaneous substitution and use expression
composition from Theorem~\ref{thm:substitution-algebra}.  For the
one-variable substitution law, the two induced substitutions in
\(\Sub(n+1,m)\) agree at \(0\), where both return \(a[\sigma]\), and at
\(i+1\), where both return \(\sigma(i)\).  Function extensionality and
expression composition give the displayed equality.

For the two-variable substitution law, both sides act on \(M\) by substitutions in
\(\Sub(n+2,m)\).  At index \(0\) both return \(p[\sigma]\); at index \(1\)
both return \(b[\sigma]\); at \(i+2\), one side gives
\(\Var(i)[\sigma]=\sigma(i)\), while the other gives
\(\Var(i+2)[\sigma^{\lift\lift}]\) and then instantiates, again reducing to
\(\sigma(i)\).  The renaming proof has the identical three cases with
\(\rho\) in place of \(\sigma\).  For weakening cancellation,
\(\mathsf{ofRen}(\wk_n);_{\Sub}\langle a\rangle\) is
\(\idren^{\Sub}_n\) pointwise, and the two-fold version satisfies
\[
\mathsf{ofRen}(\wk_n);_{\Sub}
\mathsf{ofRen}(\wk_{n+1});_{\Sub}\langle b,p\rangle
 =\idren^{\Sub}_n.
\]
Expression identity completes both equations.
\end{proof}

\section{Typing, computation, and primitive-redex preservation}
\label{sec:judgments}

\subsection{Typing rules}

The judgment \(\Gamma\vdash t:A\) is an inductive family, not a predicate
obtained from target semantics.  Variables and universes satisfy
\[
 \frac{i\in\Fin(n)}{\Gamma\vdash\Var(i):\Gamma(i)}
 \qquad
 \frac{}{\Gamma\vdash\Sort_\ell:\Sort_{\ell+1}}.
\]
The dependent product rules are
\[
\frac{\Gamma\vdash A:\Sort_\ell\qquad
      \Ext{\Gamma}{A}\vdash B:\Sort_\ell}
     {\Gamma\vdash\PiTy(A,B):\Sort_\ell}.
\]
\[
\frac{\Gamma\vdash A:\Sort_\ell\quad
      \Ext{\Gamma}{A}\vdash B:\Sort_\ell\quad
      \Ext{\Gamma}{A}\vdash t:B}
     {\Gamma\vdash\Lam(t):\PiTy(A,B)},
\]
\[
\frac{\Gamma\vdash f:\PiTy(A,B)\qquad\Gamma\vdash a:A}
     {\Gamma\vdash\App(f,a):B\{a\}}.
\]
The sum rules form \(\SigmaTy(A,B)\), introduce
\(\Pair(a,b)\) from \(a:A\) and \(b:B\{a\}\), and type the projections as
\[
  \Gamma\vdash\Fst(p):A,\qquad
  \Gamma\vdash\Snd(p):B\{\Fst(p)\}.
\]

For each \(\ell\), the natural-number rules are
\[
\Gamma\vdash\NatTy_\ell:\Sort_\ell,\quad
\Gamma\vdash\Zero_\ell:\NatTy_\ell,\quad
\frac{\Gamma\vdash k:\NatTy_\ell}
     {\Gamma\vdash\Succ_\ell(k):\NatTy_\ell}.
\]
If \(M\) is a type in \(\Ext{\Gamma}{\NatTy_\ell}\), the zero case has type
\(M\{\Zero_\ell\}\), and the successor case is checked in
\(\Ext{\Ext{\Gamma}{\NatTy_\ell}}{M}\) against the explicit substitution of
\(\Succ_\ell(\Var(1))\) for the natural argument, then the recursor has type
\(M\{k\}\).  Appendix~\ref{app:grammar} records the complete rule.

The universe-code rules include
\[
\frac{}{\Gamma\vdash\CodeNat_\ell:\Sort_\ell},
\qquad
\frac{\Gamma\vdash A:\Sort_\ell\quad
      \Ext{\Gamma}{\El(A)}\vdash B:\Sort_\ell}
     {\Gamma\vdash\CodePi(A,B):\Sort_\ell},
\]
with an analogous sum rule, and
\[
\frac{\Gamma\vdash A:\Sort_\ell\quad
      \Gamma\vdash a:\El(A)\quad\Gamma\vdash b:\El(A)}
     {\Gamma\vdash\CodeId(A,a,b):\Sort_\ell}.
\]
Any \(c:\Sort_\ell\) decodes to \(\El(c):\Sort_\ell\).

Identity formation and reflexivity are standard for the bounded grammar:
\[
\frac{\Gamma\vdash A:\Sort_\ell\quad\Gamma\vdash a:A\quad
      \Gamma\vdash b:A}
     {\Gamma\vdash\IdTy(A,a,b):\Sort_\ell},
\qquad
\frac{\Gamma\vdash a:A}
     {\Gamma\vdash\Refl(a):\IdTy(A,a,a)}.
\]
For raw \(\EqJ\), if \(M\) is a type expression in the motive
assignment~\eqref{eq:motive-context}, then
\[
\frac{
 \begin{array}{c}
 \Gamma\vdash A:\Sort_\ell\quad\Gamma\vdash a:A\\
 \Ext{\Ext{\Gamma}{A}}{\mathsf{IdProof}_n(A,a)}
      \vdash M:\Sort_j\\
 \Gamma\vdash r:M\{a,\Refl(a)\}\quad
 \Gamma\vdash b:A\quad
 \Gamma\vdash p:\IdTy(A,a,b)
 \end{array}}
 {\Gamma\vdash\EqJ(M,r,b,p):M\{b,p\}}.
\tag{5.1}\label{eq:eqj-rule}
\]

\begin{warning}[No conversion rule]
The raw judgment has no general rule converting a term from \(A\) to a
definitionally equal \(B\).  Nor is there a typed judgment
\(\Gamma\vdash A\defeq B\).  The preservation theorem below therefore
returns an explicit target type expression \(B\) and an \emph{untyped}
derivation \(B\defeq A\), rather than silently applying conversion.
\end{warning}

\subsection{Primitive computation and definitional equality}

The directed computation judgment \(r\red s\) has exactly the following
schemes:
\begin{equation}
\begin{aligned}
\App(\Lam(t),a)&\red t\{a\},\\
\Fst(\Pair(a,b))&\red a,
&
\Snd(\Pair(a,b))&\red b,\\
\NatElim_\ell(M,z,s,\Zero_\ell)&\red z,\\
\NatElim_\ell(M,z,s,\Succ_\ell(k))
 &\red s\{k,\NatElim_\ell(M,z,s,k)\},\\
\EqJ(M,r,a,\Refl(a))&\red r,\\
\El(\CodeNat_\ell)&\red\NatTy_\ell,\\
\El(\CodePi(A,B))&\red\PiTy(\El(A),\El(B)),\\
\El(\CodeSigma(A,B))&\red\SigmaTy(\El(A),\El(B)),\\
\El(\CodeId(A,a,b))&\red\IdTy(\El(A),a,b).
\end{aligned}
\tag{5.2}\label{eq:computations}
\end{equation}
Only raw expressions occur in these rules.  These are exactly ten
constructors of a top-level relation.  In particular, \(r\red s\) does not
mean reduction under an expression context, and it has no reflexive,
transitive, or multi-step constructor.  Contextual closure appears only in
the separate relation \(\defeq\).

Source definitional equality \(\defeq\) is generated by primitive
computation, reflexivity, symmetry, transitivity, and a constructor-by-
constructor compatible closure.  The closure is explicit: it covers both
arguments of products and sums, lambda bodies, function and argument
positions of application, pair components, projections, successor, all four
recursor positions, code arguments, decoding, the three identity positions,
reflexivity, and all four \(\EqJ\) positions.  This enumeration is important:
there is no clause saying that an arbitrary target trace creates source
definitional equality.

\begin{theorem}[Fundamental typing substitution]
\label{thm:fundamental-substitution}
Let \(\Gamma\in\Ctx(n)\), \(\Delta\in\Ctx(m)\), and
\(\sigma\in\Sub(n,m)\).  Suppose that \(\Gamma\vdash t:A\), and that
\(\sigma\) is typed from \(\Gamma\) to \(\Delta\), meaning
\[
      \Delta\vdash\sigma(i):\Gamma(i)[\sigma]
                 \quad\text{for every }i\in\Fin(n).
\tag{5.3}\label{eq:typed-substitution}
\]
Then
\[
                   \Delta\vdash t[\sigma]:A[\sigma].
\]
\end{theorem}

\begin{proof}
Induct on the typing derivation.  The variable case is
\eqref{eq:typed-substitution}.  The binder lemma says that a typed
\(\sigma\in\Sub(n,m)\) induces a typed
\(\sigma^\lift\in\Sub(n+1,m+1)\) from \(\Ext{\Gamma}{A}\) to
\(\Ext{\Delta}{A[\sigma]}\).  At index \(0\), the image is \(\Var(0)\), whose
declaration is \(A[\sigma][\wk_m]_{\mathrm r}\); this is also
\((A[\wk_n]_{\mathrm r})[\sigma^\lift]\) by the weakening square.  At
\(i+1\), the image \(\sigma(i)[\wk_m]_{\mathrm r}\) is typed by the renaming
admissibility theorem applied to
\(\Delta\vdash\sigma(i):\Gamma(i)[\sigma]\).  The type is rewritten by the
same square to
\((\Gamma(i)[\wk_n]_{\mathrm r})[\sigma^\lift]\).

We now inspect every typing constructor.  Universes, natural formation,
zero, and the natural code are unchanged.  Product and sum formation, lambda
introduction, and product/sum code formation apply the induction hypotheses
under \(\sigma^\lift\) using the binder lemma.  Application uses the
induction hypotheses for \(f\) and \(a\), then the one-variable equation
\[
 (B\{a\})[\sigma]=B[\sigma^\lift]\{a[\sigma]\}.
\]
Pairing uses the same equation for the second component.  First and second
projection, successor, decoding, identity/code-identity formation, and
reflexivity use their induction hypotheses componentwise; the second
projection again rewrites its displayed type by the one-variable equation.
This accounts for every constructor except \(\NatElim\) and \(\EqJ\).

For the natural successor branch, apply the binder lemma twice to
obtain a typed
\(\sigma^{\lift\lift}\in\Sub(n+2,m+2)\).  The induction hypothesis types the
substituted branch at
\(T_{\mathrm{succ}}(M)[\sigma^{\lift\lift}]\); the pointwise calculation
\[
 T_{\mathrm{succ}}(M)[\sigma^{\lift\lift}]
   =T_{\mathrm{succ}}(M[\sigma^\lift])
\tag{5.4}\label{eq:nat-target-substitution}
\]
has two cases on an index of \(M\): its natural argument \(0\), which is sent
to \(\Succ_\ell(\Var(1))\), and an inherited variable \(i+1\), shifted by
two.  The motive premise uses \(\sigma^\lift\), the zero case and scrutinee
use \(\sigma\), and the successor branch uses
\(\sigma^{\lift\lift}\).  The conclusion type is put in the required form by
\[
 M\{k\}[\sigma]=M[\sigma^\lift]\{k[\sigma]\}.
\]

The raw \(\EqJ\) motive uses the same double lift.  The identity-proof
declaration satisfies
\[
 \mathsf{IdProof}_n(A,a)[\sigma^\lift]
   =\mathsf{IdProof}_m(A[\sigma],a[\sigma]),
\tag{5.5}\label{eq:id-proof-substitution}
\]
because the weakened \(A\) and \(a\) commute with lift and \(\Var(0)\) is
fixed.  Thus the twice-lifted environment is typed for the substituted motive
assignment.  In the \(\EqJ\) case the induction hypotheses give, separately,
the substituted formation of \(A\), typing of \(a\), typing of
\(M[\sigma^{\lift\lift}]\) in that assignment, and typing of
\(r[\sigma]\), \(b[\sigma]\), and \(p[\sigma]\).  The proof premise has type
\[
 \IdTy(A[\sigma],a[\sigma],b[\sigma])
\]
by structural substitution.  The reflexivity-case type and conclusion type
are rewritten by the same two-variable law:
\[
 \begin{aligned}
 M\{a,\Refl(a)\}[\sigma]
  &=M[\sigma^{\lift\lift}]
       \{a[\sigma],\Refl(a[\sigma])\},\\
 M\{b,p\}[\sigma]
  =M[\sigma^{\lift\lift}]\{b[\sigma],p[\sigma]\},
 \end{aligned}
\]
the three-index calculation of Proposition~\ref{prop:instantiation}.  These
are exactly the premise and conclusion types of the substituted raw
\(\EqJ\) rule.
No quotient or target trace is used.
\end{proof}

\begin{corollary}[Typed instantiation]
\label{cor:typed-instantiation}
If \(\Ext{\Gamma}{A}\vdash t:T\) and \(\Gamma\vdash a:A\), then
\[
             \Gamma\vdash t\{a\}:T\{a\}.
\]
If \(\Ext{\Ext{\Gamma}{A}}{B}\vdash t:T\), \(\Gamma\vdash a:A\), and
\(\Gamma\vdash p:B\{a\}\), then
\[
             \Gamma\vdash t\{a,p\}:T\{a,p\}.
\]
\end{corollary}

\begin{theorem}[Computation and definitional equality under substitution]
\label{thm:computation-substitution}
For \(r,s\in\Expr_n\) and every \(\sigma\in\Sub(n,m)\),
\[
 r\red s\Longrightarrow r[\sigma]\red s[\sigma],
 \qquad
 r\defeq s\Longrightarrow r[\sigma]\defeq s[\sigma].
\]
\end{theorem}

\begin{proof}
Inspect the ten computation constructors.  Product beta is preserved because
\[
\App(\Lam(t),a)[\sigma]
 =\App(\Lam(t[\sigma^\lift]),a[\sigma])
 \red t[\sigma^\lift]\{a[\sigma]\}
 =(t\{a\})[\sigma].
\]
The two projection redexes become
\[
\Fst(\Pair(a[\sigma],b[\sigma]))\red a[\sigma],
\qquad
\Snd(\Pair(a[\sigma],b[\sigma]))\red b[\sigma].
\]
Natural-zero beta becomes
\[
\NatElim_\ell(M[\sigma^\lift],z[\sigma],
 s[\sigma^{\lift\lift}],\Zero_\ell)\red z[\sigma].
\]
Natural-successor beta is
\[
\begin{aligned}
&\NatElim_\ell(M[\sigma^\lift],z[\sigma],
 s[\sigma^{\lift\lift}],\Succ_\ell(k[\sigma]))\\
&\quad\red
s[\sigma^{\lift\lift}]
 \{k[\sigma],
   \NatElim_\ell(M[\sigma^\lift],z[\sigma],
       s[\sigma^{\lift\lift}],k[\sigma])\}\\
&\quad=
\bigl(s\{k,\NatElim_\ell(M,z,s,k)\}\bigr)[\sigma],
\end{aligned}
\]
by the two-variable instantiation equation.  Raw \(J\)-beta becomes
\[
\begin{aligned}
&\EqJ(M,r,a,\Refl(a))[\sigma]\\
&\quad=
\EqJ(M[\sigma^{\lift\lift}],r[\sigma],a[\sigma],
     \Refl(a[\sigma]))
\red r[\sigma],
\end{aligned}
\]
and the four decoding redexes become, respectively,
\[
\begin{gathered}
\El(\CodeNat_\ell)\red\NatTy_\ell,\\
\El(\CodePi(A[\sigma],B[\sigma^\lift]))
 \red\PiTy(\El(A[\sigma]),\El(B[\sigma^\lift])),\\
\El(\CodeSigma(A[\sigma],B[\sigma^\lift]))
 \red\SigmaTy(\El(A[\sigma]),\El(B[\sigma^\lift])),\\
\El(\CodeId(A[\sigma],a[\sigma],b[\sigma]))
 \red\IdTy(\El(A[\sigma]),a[\sigma],b[\sigma]).
\end{gathered}
\]
This exhausts the ten constructors.

For \(\defeq\), induct on its derivation.  Reflexivity, symmetry, and
transitivity use the corresponding closure constructors; a beta leaf uses
the preceding ten-case argument.  Product/sum domain congruence applies the
induction hypothesis under \(\sigma\), while codomain congruence applies it
under \(\sigma^\lift\); lambda and product/sum-code binders likewise use one
lift.  In \(\NatElim\), the motive uses \(\sigma^\lift\), the successor
branch uses \(\sigma^{\lift\lift}\), and the other two positions use
\(\sigma\).  In \(\EqJ\), the motive uses two lifts and the other three
positions use none.  Application, pairing, projections, successor, identity,
code identity, decoding, and reflexivity use the induction hypotheses in
their same-scope positions.  Reapplying the matching congruence constructor
proves each case, with no quotient assumption.
\end{proof}

\begin{theorem}[Primitive-redex preservation modulo type \(\defeq\)]
\label{thm:redex-preservation}
For \(\Gamma\in\Ctx(n)\) and \(r,s,A\in\Expr_n\), if
\(\Gamma\vdash r:A\) and \(r\red s\), then there is \(B\in\Expr_n\)
such that
\[
                  \Gamma\vdash s:B
                  \qquad\text{and}\qquad B\defeq A.
\tag{5.6}\label{eq:redex-preservation}
\]
Here \(B\defeq A\) is the untyped source relation applied to expressions
being used as types.  The theorem covers exactly the ten top-level schemes
in~\eqref{eq:computations}; it does not cover contextual or multi-step
reduction.  We reserve ``subject reduction'' for that stronger future
theorem.
\end{theorem}

\begin{proof}
Invert the typing derivation of the redex and inspect which of the ten
computation constructors was supplied.
The complete case analysis is summarized here; ``same'' means that the
target derivation has exactly the displayed source type.
\[
\begin{array}{c|c|c}
\text{redex}&\text{target term and chosen target type}&
  \text{type comparison}\\ \hline
\Pi\text{-}\beta&t\{a\}:B\{a\}&\text{same}\\
\Sigma\text{-fst}&a:A&\text{same}\\
\Sigma\text{-snd}&b:B\{a\}&
 B\{a\}\defeq B\{\Fst(\Pair(a,b))\}\\
\NatTy\text{-zero}&z:M\{\Zero_\ell\}&\text{same}\\
\NatTy\text{-succ}&s\{k,R_k\}:M\{\Succ_\ell(k)\}&\text{same}\\
\EqJ\text{-}\beta&r:M\{a,\Refl(a)\}&\text{same}\\
\El\text{-Nat}&\NatTy_\ell:\Sort_\ell&\text{same}\\
\El\text{-}\Pi&\PiTy(\El(A),\El(B)):\Sort_\ell&\text{same}\\
\El\text{-}\Sigma&\SigmaTy(\El(A),\El(B)):\Sort_\ell&\text{same}\\
\El\text{-Id}&\IdTy(\El(A),a,b):\Sort_\ell&\text{same}
\end{array}
\tag{5.7}\label{eq:redex-case-table}
\]

For product beta, inversion produces
\(\Ext{\Gamma}{A}\vdash t:B\) and \(\Gamma\vdash a:A\);
Corollary~\ref{cor:typed-instantiation} gives
\(\Gamma\vdash t\{a\}:B\{a\}\).
For first projection, inversion reaches the pair premise
\(\Gamma\vdash a:A\).  For natural-zero beta, inversion reaches the stored
zero-case premise \(\Gamma\vdash z:M\{\Zero_\ell\}\).

For the dependent second projection, inversion has the exact shape
\[
 \frac{\Gamma\vdash a:A\qquad\Gamma\vdash b:B\{a\}}
      {\Gamma\vdash\Snd(\Pair(a,b)):
         B\{\Fst(\Pair(a,b))\}}.
\]
The target \(b\) is already typed at \(B\{a\}\).  Primitive first-projection
beta gives
\(\Fst(\Pair(a,b))\defeq a\); symmetry gives
\(a\defeq\Fst(\Pair(a,b))\).  Compare the two substitutions in
\(\Sub(n+1,n)\):
\[
\begin{array}{lll}
\langle a\rangle(0)=a,
&\quad&
\langle\Fst(\Pair(a,b))\rangle(0)=\Fst(\Pair(a,b)),\\
\langle a\rangle(i+1)=\Var(i),
&&
\langle\Fst(\Pair(a,b))\rangle(i+1)=\Var(i).
\end{array}
\]
They are pointwise \(\defeq\)-equal by the symmetric beta step at \(0\) and
reflexivity at \(i+1\).  Applying substitution-environment congruence
(Proposition~\ref{prop:environment-congruence}) to the body \(B\) gives
\[
       B\{a\}\defeq B\{\Fst(\Pair(a,b))\}.
\]
This is why the theorem returns a possibly different type expression instead
of pretending that a conversion rule exists.

For natural successor beta, write
\[
 R_k=\NatElim_\ell(M,z,s,k).
\]
Inversion gives \(\Gamma\vdash k:\NatTy_\ell\), hence the recursor rule gives
\(\Gamma\vdash R_k:M\{k\}\).  The successor branch is typed in
\(\Ext{\Ext{\Gamma}{\NatTy_\ell}}{M}\) at
\(T_{\mathrm{succ}}(M)\).  Two-variable typed
instantiation with \(k\) and \(R_k\) yields
\[
 \Gamma\vdash s\{k,R_k\}:
     T_{\mathrm{succ}}(M)\{k,R_k\}.
\]
The pointwise substitution calculation
\[
 T_{\mathrm{succ}}(M)\{k,R_k\}=M\{\Succ_\ell(k)\}
\tag{5.8}\label{eq:nat-target-instantiation}
\]
is exact: the branch proof variable does not occur in
\(T_{\mathrm{succ}}(M)\), the predecessor becomes \(k\), and the motive
variable becomes \(\Succ_\ell(k)\).  This is the displayed source type of the
successor redex.

For raw \(J\)-beta, inversion reaches the stored premise
\(\Gamma\vdash r:M\{a,\Refl(a)\}\), exactly the redex type.

For \(\El(\CodeNat_\ell)\), inversion first yields
\(\Gamma\vdash\CodeNat_\ell:\Sort_\ell\); natural formation gives
\(\Gamma\vdash\NatTy_\ell:\Sort_\ell\).
For a product code, inversion yields
\[
\Gamma\vdash A:\Sort_\ell,\qquad
\Ext{\Gamma}{\El(A)}\vdash B:\Sort_\ell.
\]
Decoding formation gives the domain
\(\Gamma\vdash\El(A):\Sort_\ell\) and the codomain
\(\Ext{\Gamma}{\El(A)}\vdash\El(B):\Sort_\ell\); product formation therefore
types \(\PiTy(\El(A),\El(B))\) at \(\Sort_\ell\).
The sum-code case is identical with sum formation.  For an identity code,
inversion gives
\[
\Gamma\vdash A:\Sort_\ell,\quad
\Gamma\vdash a:\El(A),\quad
\Gamma\vdash b:\El(A);
\]
decoding formation and identity formation type
\(\IdTy(\El(A),a,b)\) at \(\Sort_\ell\).
All four source decoding redexes were displayed at \(\Sort_\ell\), so their
type-coherence proof is reflexivity.  This completes all ten cases without a
target-typing premise or a conversion rule.
\end{proof}

\section{The syntactic quotient and compositional substitution}
\label{sec:syntactic-quotient}

\subsection{The quotient carrier}

For each scope \(n\), source definitional equality is a setoid on
\(\Expr_n\).  Define
\[
              \Q_n=\Expr_n/{\defeq},
       \qquad \eta_n(t)=\den t.
\tag{6.1}\label{eq:syntactic-quotient}
\]
This is only a syntactic quotient at a fixed scope.  It carries neither a
context-formation judgment nor the structure of a CwF, and no initiality
property is asserted.  No normal forms are chosen.

For \(\sigma\in\Sub(n,m)\), put
\[
       \Q(\sigma)(\den t)=\den{t[\sigma]}.
\tag{6.2}\label{eq:quotient-substitution}
\]
Theorem~\ref{thm:computation-substitution} proves that this definition is
independent of the representative \(t\).  This order matters logically:
substitution closure is a theorem about the source relation, proved before
the quotient action is defined.

\begin{proposition}[Substitution environment congruence]
\label{prop:environment-congruence}
Suppose \(\sigma,\tau\in\Sub(n,m)\) and
\(\sigma(i)\defeq\tau(i)\) for every \(i\in\Fin(n)\).  Then
\[
                     t[\sigma]\defeq t[\tau]
                     \quad\text{for every }t\in\Expr_n.
\tag{6.3}\label{eq:subst-congruence}
\]
Consequently \(\Q(\sigma)=\Q(\tau)\) pointwise.
\end{proposition}

\begin{proof}
Induct on \(t\).  The variable case is the assumption.  First prove the lift
lemma pointwise:
\[
 \sigma\equiv_{\mathrm{env}}\tau
 \Longrightarrow
 \sigma^\lift\equiv_{\mathrm{env}}\tau^\lift.
\tag{6.4}\label{eq:lift-env-congruence}
\]
At index \(0\), both sides are \(\Var(0)\).  At \(i+1\), apply
Theorem~\ref{thm:computation-substitution} to
\(\sigma(i)\defeq\tau(i)\) with
\(\mathsf{ofRen}(\wk_m)\); by~\eqref{eq:ofren-subst} this is precisely
\(\sigma(i)[\wk_m]_{\mathrm r}\defeq
  \tau(i)[\wk_m]_{\mathrm r}\).  Applying the lemma twice gives pointwise
\(\sigma^{\lift\lift}\equiv_{\mathrm{env}}
 \tau^{\lift\lift}\).

For \(\PiTy,\SigmaTy,\Lam,\CodePi,\CodeSigma\), use
\eqref{eq:lift-env-congruence} in the bound argument and the original
assumption in the unbound argument.  Application, pairing, projections,
successor, decoding, identity, code identity, and reflexivity use the
induction hypotheses in their same-scope positions, chaining one-position
congruence constructors when there is more than one argument.  Constants
are reflexive.  For \(\NatElim_\ell(M,z,s,k)\), the four induction hypotheses use,
respectively, \(\sigma^\lift\), \(\sigma\),
\(\sigma^{\lift\lift}\), and \(\sigma\); transitivity chains the four
one-position congruence constructors.  Concretely, it compares motives,
zero branches, successor branches, and scrutinees in that order:
\[
\begin{aligned}
&\NatElim_\ell(M[\sigma^\lift],z[\sigma],
 s[\sigma^{\lift\lift}],k[\sigma])\\
\defeq{}&
\NatElim_\ell(M[\tau^\lift],z[\sigma],
 s[\sigma^{\lift\lift}],k[\sigma])\\
\defeq{}&
\NatElim_\ell(M[\tau^\lift],z[\tau],
 s[\sigma^{\lift\lift}],k[\sigma])\\
\defeq{}&
\NatElim_\ell(M[\tau^\lift],z[\tau],
 s[\tau^{\lift\lift}],k[\sigma])\\
\defeq{}&
\NatElim_\ell(M[\tau^\lift],z[\tau],
 s[\tau^{\lift\lift}],k[\tau]).
\end{aligned}
\]
Each step changes exactly one displayed argument using the corresponding
congruence constructor.  Raw \(\EqJ(M,r,b,p)\) similarly uses
\(\sigma^{\lift\lift}\) for \(M\) and \(\sigma\) for the remaining three
arguments:
\[
\begin{aligned}
&\EqJ(M[\sigma^{\lift\lift}],r[\sigma],b[\sigma],p[\sigma])\\
\defeq{}&
\EqJ(M[\tau^{\lift\lift}],r[\sigma],b[\sigma],p[\sigma])\\
\defeq{}&
\EqJ(M[\tau^{\lift\lift}],r[\tau],b[\sigma],p[\sigma])\\
\defeq{}&
\EqJ(M[\tau^{\lift\lift}],r[\tau],b[\tau],p[\sigma])\\
\defeq{}&
\EqJ(M[\tau^{\lift\lift}],r[\tau],b[\tau],p[\tau]).
\end{aligned}
\]
Thus every constructor, including the two-binder cases, is covered by the
explicit compatible closure of \(\defeq\).  Quotient soundness gives
\(\Q(\sigma)(\den t)=\Q(\tau)(\den t)\); quotient induction extends this to
arbitrary \(x:\Q_n\), and function extensionality identifies the two maps if
map equality rather than pointwise equality is desired.
\end{proof}

\begin{theorem}[Compositionality of quotient substitution]
\label{thm:quotient-compositionality}
For every \(x\in\Q_n\) and composable substitutions
\(\sigma\in\Sub(n,m)\), \(\tau\in\Sub(m,k)\),
\[
\begin{aligned}
\Q(\idren^{\Sub}_n)(x)&=x,\\
\Q(\tau)(\Q(\sigma)(x))&=\Q(\sigma;_{\Sub}\tau)(x).
\end{aligned}
\tag{6.5}\label{eq:quotient-compositionality}
\]
The equations hold for arbitrary quotient values, not merely for a chosen
notation of the form \(\den t\).  If \(\sigma\) and \(\tau\) are pointwise
definitionally equal, their induced maps agree.
\end{theorem}

\begin{proof}
Define \(\Q(\sigma)\) by quotient elimination from
\[
 t\longmapsto\den{t[\sigma]}.
\]
Its representative-independence obligation is: from \(t\defeq u\), prove
\(\den{t[\sigma]}=\den{u[\sigma]}\).  Theorem~\ref{thm:computation-substitution}
gives \(t[\sigma]\defeq u[\sigma]\), and quotient soundness supplies exactly
that equality.  Thus the map is well defined.

For identity and composition, use quotient induction to reduce an arbitrary
\(x\) to \(\den t\).  Equations~\eqref{eq:sub-algebra} give
\(\den{t[\idren^{\Sub}_n]}=\den t\) and
\[
 \begin{aligned}
 \Q(\idren^{\Sub}_n)(\den t)
   &=\den{t[\idren^{\Sub}_n]}=\den t,\\
 \Q(\tau)(\Q(\sigma)(\den t))
   &=\den{t[\sigma][\tau]}
     =\den{t[\sigma;_{\Sub}\tau]}
     =\Q(\sigma;_{\Sub}\tau)(\den t).
 \end{aligned}
\]
Proposition~\ref{prop:environment-congruence} proves that pointwise
\(\defeq\)-equal environments induce pointwise equal quotient maps; function
extensionality gives equality of maps.  Hence the construction is independent
both of expression representatives and of pointwise equivalent environment
representatives, and the two displayed equations hold on every quotient
value.
\end{proof}

Theorem~\ref{thm:quotient-compositionality} asserts only these identity and
composition equations.  We have not defined a category of well-formed
contexts and substitutions, so no categorical structure is asserted.

\subsection{Action on quoted expressions}

For \(t\in\Expr_n\) and \(\sigma\in\Sub(n,m)\), abbreviate
\[
                       \den t_\sigma\coloneqq
                       \Q(\sigma)(\den t)=\den{t[\sigma]}.
\tag{6.6}\label{eq:environment-interpretation}
\]
Then
\[
          \den{t[\sigma]}_\tau
            =\den t_{\sigma;_{\Sub}\tau}.
\tag{6.7}\label{eq:interpretation-composition}
\]
The quotient equality in~\eqref{eq:interpretation-composition} may be
decorated by a singleton unlabelled quotient trace.  That decoration does
not remember the substitution proof and plays no role in establishing the
equation.

Typing can be saturated by quotient representatives without calling the
quotient a typed model.

\begin{definition}[Quotient-saturated typing]
\label{def:quotient-typing}
For \(x,X\in\Q_n\), write
\(\Gamma\models_{\Q}x:X\) when there exist \(t,A\in\Expr_n\) such that
\[
             \Gamma\vdash t:A,\qquad x=\den t,\qquad X=\den A.
\]
\end{definition}

\begin{proposition}[Substitution for quotient-saturated typing]
\label{prop:quotient-typing-substitution}
If \(\Gamma\models_{\Q}x:X\) and \(\sigma\) is a typed substitution from
\(\Gamma\) to \(\Delta\), then
\[
        \Delta\models_{\Q}\Q(\sigma)(x):\Q(\sigma)(X).
\]
\end{proposition}

\begin{proof}
Choose the representatives in Definition~\ref{def:quotient-typing} and
apply Theorem~\ref{thm:fundamental-substitution}.  The equalities identifying
\(x\) and \(X\) with their representatives are preserved by the function
\(\Q(\sigma)\).
\end{proof}

This relation merely packages a raw representative and two quotient
equalities.  It supplies no context validity, conversion, or initial CwF.

\section{Source identity programs and unlabelled quotient traces}
\label{sec:identity-programs}

\subsection{Finite same-scope frames and source programs}

For \(t\in\Expr_n\), let \(C[t]\) denote plugging \(t\) into the one-hole
frame \(C\).  The finite frame grammar used here is
\[
\begin{array}{llll}
\App([-],a),&\App(f,[-]),&
\Pair([-],b),&\Pair(a,[-]),\\
\Fst([-]),&\Snd([-]),&\Succ_\ell([-]),&\El([-]),\\
\Refl([-]),&
\IdTy([-],a,b),&\IdTy(A,[-],b),&\IdTy(A,a,[-]),\\
\CodeId([-],a,b),&\CodeId(A,[-],b),&\CodeId(A,a,[-]).
\end{array}
\tag{7.1}\label{eq:frames}
\]
These are same-scope frames; they do not include binder contexts.

\begin{lemma}[Frame congruence]
\label{lem:frame-congruence}
If \(t\defeq u\), then \(C[t]\defeq C[u]\) for every
\(C\) in~\eqref{eq:frames}.  Hence
\[
 \overline C:\Q_n\to\Q_n,\qquad
 \overline C(\den t)=\den{C[t]}
\]
is representative-independent.
\end{lemma}

\begin{proof}
Each case is the corresponding one-position congruence constructor of
\(\defeq\).  Quotient soundness proves representative independence.
\end{proof}

\begin{definition}[Source identity programs]
\label{def:identity-programs}
For \(t,u\in\Expr_n\), the family \(\IProg_n(t,u)\) is generated by
\[
\frac{h:t\defeq u}{\mathsf{atom}(h):\IProg_n(t,u)},\qquad
\mathbf{1}_t:\IProg_n(t,t),
\]
\[
\frac{p:\IProg_n(t,u)}{p^{-1}:\IProg_n(u,t)},\qquad
\frac{p:\IProg_n(t,u)\quad q:\IProg_n(u,v)}
     {p\cdot q:\IProg_n(t,v)}.
\]
\[
\frac{p:\IProg_n(t,u)}
     {C[p]:\IProg_n(C[t],C[u])}.
\]
\end{definition}

The atom accepts an untyped source \(\defeq\)-derivation.  It cannot accept a
target trace step or target trace equivalence.  The family itself has no
typing judgment for its endpoints.

\begin{proposition}[Endpoint triviality]
\label{prop:iprogram-endpoints}
For \(t,u\in\Expr_n\),
\[
 \operatorname{Nonempty}\bigl(\IProg_n(t,u)\bigr)
 \quad\Longleftrightarrow\quad t\defeq u.
\tag{7.2}\label{eq:iprogram-endpoint-iff}
\]
\end{proposition}

\begin{proof}
For the forward implication, choose \(p:\IProg_n(t,u)\) and induct on
\(p\).  An atom contains the required derivation.  Unit uses reflexivity;
inverse uses symmetry; composition uses transitivity; and \(C[p]\) uses
Lemma~\ref{lem:frame-congruence}.  Conversely, from \(h:t\defeq u\) take
\(\mathsf{atom}(h):\IProg_n(t,u)\).
\end{proof}

Thus identity-program syntax adds compositional trace structure only after
endpoint reachability has already been supplied by the untyped relation
\(\defeq\).  It cannot discover a new pair of reachable endpoints, and its
terms need not relate expressions having a common raw type.

\begin{proposition}[Structural-recursion interpretation]
\label{prop:iprogram-recursion}
Fix \(n\).  Let \(G(t,u)\) be any family indexed by
\(t,u\in\Expr_n\), equipped with:
\[
\begin{array}{ll}
a_h:G(t,u)&\text{for every }h:t\defeq u,\\
e_t:G(t,t),&\\
i_{t,u}:G(t,u)\to G(u,t),&\\
m_{t,u,v}:G(t,u)\to G(u,v)\to G(t,v),&\\
F_C:G(t,u)\to G(C[t],C[u])&\text{for every }C\in\Frame_n.
\end{array}
\tag{7.3}\label{eq:target-signature}
\]
There is a unique pointwise family of maps
\[
        \Phi_{t,u}:\IProg_n(t,u)\to G(t,u)
\tag{7.4}\label{eq:iprogram-fold}
\]
such that
\[
\begin{aligned}
\Phi(\mathsf{atom}(h))&=a_h,&
\Phi(\mathbf1_t)&=e_t,\\
\Phi(p^{-1})&=i(\Phi p),&
\Phi(p\cdot q)&=m(\Phi p)(\Phi q),\\
\Phi(C[p])&=F_C(\Phi p).
\end{aligned}
\tag{7.5}\label{eq:iprogram-fold-equations}
\]
\end{proposition}

\begin{proof}
Existence is structural recursion on the five constructors of
\(\IProg\).  If \(\Psi\) satisfies the same equations, induction on
\(p:\IProg_n(t,u)\) gives \(\Phi(p)=\Psi(p)\) in each constructor case.
This is pointwise uniqueness; function extensionality converts it to equality
of the map families if desired.
\end{proof}

Proposition~\ref{prop:iprogram-recursion} is a free-\emph{syntax} property
only.  It imposes no groupoid equations on \(G\) and says nothing about the
quotient of \(\IProg\) by source rewrites.  A universal property for a free
groupoid quotient would require proving that all source relations are
respected and is not claimed here.

\subsection{The unlabelled quotient-trace record}

\begin{definition}[Trace steps and trace records]
\label{def:trace-record}
For a metatheoretic carrier \(X\), define
\[
 \mathsf{QStep}(X)\coloneqq\sum_{x,y:X}(x=y),
 \qquad
 \Tr_X(a,b)\coloneqq
   \operatorname{List}(\mathsf{QStep}(X))\times(a=b).
\tag{7.6}\label{eq:trace-record}
\]
If \(S=(L,e):\Tr_X(a,b)\), write
\(\operatorname{list}(S)=L\) and
\(\operatorname{end}(S)=e\).
\end{definition}

The basic operations are fully determined on both fields.  Put
\[
\begin{aligned}
\mathbf r_a&=([],\mathsf{refl}_a),\\
(L,e)^{-1}
  &=\bigl(\operatorname{map}(\mathsf{inv},
       \operatorname{reverse}(L)),e^{-1}\bigr),\\
(L,e)\cdot(M,d)&=(L{+\!\!+}M,e\mathbin{\cdot_{=}}d),
\end{aligned}
\tag{7.7}\label{eq:trace-operations}
\]
where
\(\mathsf{inv}(x,y,h)=(y,x,h^{-1})\), \(+\!\!+\) is list
concatenation, and \(\cdot_{=}\) is transitivity of ambient equality.
For \(f:X\to Y\),
\[
 f_\ast(L,e)=
 \bigl(\operatorname{map}(x,y,h\mapsto
       (fx,fy,\mathsf{ap}_f(h)),L),\mathsf{ap}_f(e)\bigr).
\tag{7.8}\label{eq:trace-map}
\]
Thus \(f_\ast\mathbf r_a=\mathbf r_{fa}\),
\(f_\ast(S^{-1})=(f_\ast S)^{-1}\), and
\(f_\ast(S\cdot T)=f_\ast S\cdot f_\ast T\).

For a completely arbitrary record, Definition~\ref{def:trace-record} imposes
no adjacency relation between successive list entries and no relation between
the list and the outer endpoint field.  This fact is not used as a positive
invariant.  The soundness theorem quantifies only over expressions of the
form \(\ev(p)\).

When \(X=\Q_n\), the record is \emph{unlabelled}.  Given
\(h:t\defeq u\), quotient soundness produces
\(\bar h:\den t=\den u\), and the atom is evaluated by
\[
 \langle\bar h\rangle
 =\bigl([(\den t,\den u,\bar h)],\bar h\bigr).
\]
The derivation \(h\) is absent from this record.  In particular, two
\(\defeq\)-derivations with the same endpoints yield equal quotient-equality
proofs by proof irrelevance.  This differs from the label-sensitive calculus
of~\cite{deQueirozEtAl2016,RamosEtAl2017,RamosEtAl2018,RamosEtAl2026}.
Because \(\den t=\den u\) already in \(\Q_n\), an evaluated identity program
is best regarded as a trace-decorated loop over that quotient-collapsed
endpoint (although the displayed record retains the two representatives).

Evaluation is structural:
\[
\begin{aligned}
\ev(\mathsf{atom}(h))&=\langle\bar h\rangle,&
\ev(\mathbf{1}_t)&=\mathbf r_{\den t},\\
\ev(p^{-1})&=\ev(p)^{-1},&
\ev(p\cdot q)&=\ev(p)\cdot\ev(q),\\
\ev(C[p])&=\overline C_\ast(\ev(p)).
\end{aligned}
\tag{7.9}\label{eq:evaluation}
\]
This is the instance of Proposition~\ref{prop:iprogram-recursion} with
\[
 G(t,u)=\Tr_{\Q_n}(\den t,\den u),\quad
 a_h=\langle\bar h\rangle,\quad e_t=\mathbf r_{\den t},
\]
and with inverse, composition, and frame actions supplied by
\eqref{eq:trace-operations} and~\eqref{eq:trace-map}.
If desired, the evaluated sublanguage can be named without asserting an
unproved trace invariant:
\[
 \mathsf{EvalTr}_n(t,u)
   =\{\,T:\Tr_{\Q_n}(\den t,\den u)
        \mid \exists p:\IProg_n(t,u),\ T=\ev(p)\,\}.
\tag{7.10}\label{eq:evaluated-image}
\]
We use only the witness \(p\) and the defining equations~\eqref{eq:evaluation}.

\subsection{The target trace rewrite relation}

We now give the entire target relation needed below.  At fixed endpoints,
the one-step relation \(S\TrStep T\) has exactly eight schemes:
\[
\begin{array}{rclcrcl}
(\mathbf r_a)^{-1}&\TrStep&\mathbf r_a,
&& (S^{-1})^{-1}&\TrStep&S,\\
\mathbf r_a\cdot S&\TrStep&S,
&&S\cdot\mathbf r_b&\TrStep&S,\\
S\cdot S^{-1}&\TrStep&\mathbf r_a,
&&S^{-1}\cdot S&\TrStep&\mathbf r_b,\\
(S\cdot T)^{-1}&\TrStep&T^{-1}\cdot S^{-1},
&&(S\cdot T)\cdot U&\TrStep&S\cdot(T\cdot U).
\end{array}
\tag{7.11}\label{eq:target-one-step}
\]
Here \(S:\Tr_X(a,b)\), \(T:\Tr_X(b,c)\), and
\(U:\Tr_X(c,d)\) where required.  These rules are equations of trace
expressions oriented from left to right; they do not inspect source syntax.

\begin{definition}[Target trace equivalence]
\label{def:target-trace-equivalence}
\(\TrEq\) is the least endpoint-indexed family generated by
\[
\frac{}{S\TrEq S},\qquad
\frac{S\TrStep T}{S\TrEq T},\qquad
\frac{S\TrEq T}{T\TrEq S},\qquad
\frac{S\TrEq T\quad T\TrEq U}{S\TrEq U},
\tag{7.12}\label{eq:target-equivalence-closure}
\]
and the congruence rules
\[
\frac{S\TrEq T}{S^{-1}\TrEq T^{-1}},\qquad
\frac{S\TrEq T}{S\cdot U\TrEq T\cdot U},\qquad
\frac{T\TrEq U}{S\cdot T\TrEq S\cdot U},\qquad
\frac{S\TrEq T}{f_\ast S\TrEq f_\ast T}.
\tag{7.13}\label{eq:target-congruence}
\]
\end{definition}

Equations~\eqref{eq:target-one-step}--\eqref{eq:target-congruence} are a
complete mathematical presentation of the target fragment used in the
soundness proof.  They are the groupoid core and congruence fragment of the
larger computational-path rewrite system, not a presentation of all rules in
the monograph.

\subsection{Source rewrites and soundness}

The source one-step relation \(p\srw q\) has the eight schemes obtained from
\eqref{eq:target-one-step} by replacing \(S,T,U,\mathbf r\) with
\(p,q,r,\mathbf1\).  It also has the congruence rules
\[
\frac{p\srw q}{p^{-1}\srw q^{-1}},\quad
\frac{p\srw q}{p\cdot r\srw q\cdot r},\quad
\frac{q\srw r}{p\cdot q\srw p\cdot r},\quad
\frac{p\srw q}{C[p]\srw C[q]},
\tag{7.14}\label{eq:source-congruence-rules}
\]
and the frame-unit rule
\[
                  C[\mathbf1_t]\srw\mathbf1_{C[t]}.
\tag{7.15}\label{eq:frame-unit}
\]
Let \(\srweq\) be the reflexive, symmetric, transitive closure of
\(\srw\).

\begin{theorem}[Source identity rewrite soundness]
\label{thm:identity-soundness}
For \(p,q:\IProg_n(t,u)\),
\[
          p\srweq q
          \quad\Longrightarrow\quad
          \ev(p)\TrEq\ev(q).
\tag{7.16}\label{eq:rewrite-soundness}
\]
\end{theorem}

\begin{proof}
Induct first on a source one-step derivation.  The eight groupoid cases are
the eight target rules~\eqref{eq:target-one-step} after unfolding
\eqref{eq:evaluation}.  Inverse and the two composition congruences use the
first three rules in~\eqref{eq:target-congruence}.  Frame congruence uses the
last rule with \(f=\overline C\).  The frame-unit case is reflexivity because
\(\overline C_\ast\mathbf r_{\den t}=\mathbf r_{\den{C[t]}}\) by
\eqref{eq:trace-map}.  Induct next on the reflexive, symmetric, transitive
closure of the source relation and use the first, third, and fourth rules in
\eqref{eq:target-equivalence-closure}.  No target derivation is an input to a
source constructor.
\end{proof}

\begin{example}[Multi-step soundness]
\label{ex:two-step-coherence}
For \(p:\IProg_n(t,u)\) and \(q:\IProg_n(u,v)\), source associativity
followed by right-unit congruence gives
\[
       (p\cdot q)\cdot\mathbf{1}_v
          \srweq p\cdot q.
\]
The soundness derivation first applies target associativity and then removes
the right unit inside the right factor.  This is a two-step derivation of
trace equivalence; no equality between derivations is asserted.
\end{example}

\section{Examples: beta, dependency, and two-binder instantiation}
\label{sec:example}

\subsection{Lambda beta across the quotient layers}

Fix a level \(\ell\) and work in the empty scope.  In one bound variable let
\(x=\Var(0)\), and consider
\[
       r=\App(\Lam(x),\Zero_\ell),
       \qquad s=\Zero_\ell.
\tag{8.1}\label{eq:running-redex}
\]
The variable rule gives
\(\Ext{\EmptyCtx_0}{\NatTy_\ell}\vdash
x:\NatTy_\ell[\wk_0]_{\mathrm r}\); the latter expression is
definitionally the constant \(\NatTy_\ell\).  Product formation and lambda
introduction therefore derive
\[
  \EmptyCtx_0\vdash
     \Lam(x):\PiTy(\NatTy_\ell,\NatTy_\ell),
\]
where the codomain notation suppresses its one-variable scope.  Together
with
\(\EmptyCtx_0\vdash\Zero_\ell:\NatTy_\ell\), application gives
\[
                  \EmptyCtx_0\vdash r:\NatTy_\ell.
\]

At the operational layer, product beta gives
\[
          r\red x\{\Zero_\ell\}=\Zero_\ell=s.
\tag{8.2}\label{eq:running-beta}
\]
The final equality is the variable clause of simultaneous substitution.
Embedding primitive computation into the untyped definitional relation gives
\[
                         r\defeq s.
\tag{8.3}\label{eq:running-defeq}
\]
Consequently quotient soundness identifies their classes:
\[
                         \den r=\den s
                          \quad\text{in }\Q_0.
\tag{8.4}\label{eq:running-quotient}
\]

Let \(\beta\) denote the derivation~\eqref{eq:running-defeq} and define the
source identity atom
\[
                   \alpha=\mathsf{atom}(\beta)
                     :\IProg_0(r,s).
\]
Its evaluation is the singleton unlabelled quotient trace
\[
 P_\beta=\ev(\alpha)
  =\Bigl(
      [\,(\den r,\den s,\overline\beta)\,],
      \overline\beta
    \Bigr)
 :\Tr_{\Q_0}(\den r,\den s).
\tag{8.5}\label{eq:running-path}
\]
It has one visible list entry, but that entry contains only the quotient
equality \(\overline\beta\), not the derivation \(\beta\).  Source
cancellation is sound as
\[
    \alpha\cdot\alpha^{-1}\srw\mathbf{1}_r
    \quad\longmapsto\quad
    P_\beta\cdot P_\beta^{-1}
        \TrEq\mathbf r_{\den r}.
\tag{8.6}\label{eq:running-rweq}
\]

Thus the example exhibits source beta, untyped source definitional equality,
quotient equality, an unlabelled trace decoration, and target trace
equivalence.  The quotient step erases the computational label; no arrow is
reversed to define a source judgment.

The same calculation is natural in environments.  If
\(\App(\Lam(t),a)\in\Expr_n\) and \(\sigma\in\Sub(n,m)\), then
\[
\App(\Lam(t),a)[\sigma]
   \red t[\sigma^\lift]\{a[\sigma]\}
   =t\{a\}[\sigma].
\]
The final equation is the one-variable case of
Proposition~\ref{prop:instantiation}.

\subsection{Dependent second projection}
\label{sec:example-snd}

Suppose
\[
\Gamma\vdash A:\Sort_\ell,\qquad
\Ext{\Gamma}{A}\vdash B:\Sort_\ell,\qquad
\Gamma\vdash a:A,\qquad
\Gamma\vdash b:B\{a\}.
\]
Then pair introduction and second-projection elimination derive
\[
 \Gamma\vdash
 \Snd(\Pair(a,b)):
 B\{\Fst(\Pair(a,b))\}.
\tag{8.7}\label{eq:snd-source-type}
\]
The primitive redex is
\[
                    \Snd(\Pair(a,b))\red b.
\tag{8.8}\label{eq:snd-redex}
\]
The target \(b\) is not presented at the expression in
\eqref{eq:snd-source-type}; its derivation has type \(B\{a\}\).  The required
coherence is a source calculation:
\[
\Fst(\Pair(a,b))\red a
\ \Longrightarrow\
a\defeq\Fst(\Pair(a,b))
\ \Longrightarrow\
B\{a\}\defeq B\{\Fst(\Pair(a,b))\}.
\tag{8.9}\label{eq:snd-type-coherence}
\]
The first implication uses symmetry in \(\defeq\); the second is
one-variable substitution-environment congruence for the body \(B\).
Therefore Theorem~\ref{thm:redex-preservation} returns \(B\{a\}\) as the
target type and~\eqref{eq:snd-type-coherence} as its relation to the source
type.  This example is the reason neither a conversion rule nor literal
equality of displayed types should be inferred from the theorem.

\subsection{Natural successor beta}
\label{sec:example-nat}

Assume the premises of the natural eliminator and
\(\Gamma\vdash k:\NatTy_\ell\).  Put
\[
             R_k=\NatElim_\ell(M,z,s,k).
\]
The successor redex and contractum are
\[
\NatElim_\ell(M,z,s,\Succ_\ell(k))
  \red s\{k,R_k\}.
\tag{8.10}\label{eq:nat-succ-example}
\]
The branch derivation lives under two binders:
\[
  \Ext{\Ext{\Gamma}{\NatTy_\ell}}{M}\vdash s:T_{\mathrm{succ}}(M).
\]
Typed two-variable instantiation first uses
\(\Gamma\vdash k:\NatTy_\ell\) and then
\(\Gamma\vdash R_k:M\{k\}\), producing
\[
 \Gamma\vdash s\{k,R_k\}:
   T_{\mathrm{succ}}(M)\{k,R_k\}.
\]
The explicit substitution defining \(T_{\mathrm{succ}}\) yields
\[
 T_{\mathrm{succ}}(M)\{k,R_k\}
   =M\{\Succ_\ell(k)\},
\]
which is exactly the type of the redex.  The equality is a three-index
calculation on the substitution into \(M\), not an appeal to target traces.

\subsection{\texorpdfstring{What raw \(J\)-beta says}{What raw J-beta says}}

If the six premises of~\eqref{eq:eqj-rule} specialize to endpoint \(a\) and
proof \(\Refl(a)\), then
\[
             \EqJ(M,r,a,\Refl(a))\red r
\]
and the source and target are typed at the same raw expression
\(M\{a,\Refl(a)\}\).  This is the entire raw result.  It does not say that
\(M\) factors through quotient equality and is not interpreted by the
external construction of Section~\ref{sec:j-boundary}.

\section{Summary of the raw structural theorems}
\label{sec:main-result}

\begin{theorem}[Raw scoped structural theorem]
\label{thm:raw-structural}
For the exact grammar and judgments of
Sections~\ref{sec:syntax}--\ref{sec:judgments}:
\begin{enumerate}
\item \(\Ren(n,m)\), \(\Sub(n,m)\), their lifts, the embedding
  \(\mathsf{ofRen}\), weakening, and one- and two-variable instantiation
  satisfy Propositions~\ref{prop:renaming} and~\ref{prop:instantiation} and
  Theorem~\ref{thm:substitution-algebra}.
\item Pointwise typed \(\sigma\in\Sub(n,m)\) preserves the raw typing family.
  Primitive computation and untyped \(\defeq\) are stable under every such
  raw substitution.
\item The ten top-level primitive redex constructors satisfy
  Theorem~\ref{thm:redex-preservation}.  No contextual or multi-step
  preservation statement is included.
\item The quotient action~\eqref{eq:quotient-substitution} is independent of
  expression representatives and pointwise \(\defeq\)-equal substitution
  environments.  It satisfies identity and composition on arbitrary
  quotient values, and the quotient-saturated typing relation is stable
  under pointwise typed substitutions.
\item Source identity programs are endpoint-trivial
  (Proposition~\ref{prop:iprogram-endpoints}) and satisfy the
  operation-preserving structural-recursion principle of
  Proposition~\ref{prop:iprogram-recursion}.  Its quotient-trace instance is
  evaluation~\eqref{eq:evaluation}, and
  \(p\srweq q\) implies \(\ev(p)\TrEq\ev(q)\).
\item The raw \(\EqJ\) constructor participates only through its raw typing,
  two-binder substitution, congruence, and top-level reflexivity beta rules.
  No external motive-factorization theorem is included.
\end{enumerate}
\end{theorem}

\begin{proof}
The clauses collect
Theorems~\ref{thm:substitution-algebra},
\ref{thm:fundamental-substitution},
\ref{thm:computation-substitution},
\ref{thm:redex-preservation},
\ref{thm:quotient-compositionality}, and
\ref{thm:identity-soundness}, with
Propositions~\ref{prop:environment-congruence} and
\ref{prop:quotient-typing-substitution},
\ref{prop:iprogram-endpoints}, and
\ref{prop:iprogram-recursion}.  Each constituent proof was given
before quotient traces were introduced, except for source identity rewrite
soundness, whose target relation was explicitly defined in
\eqref{eq:target-one-step}--\eqref{eq:target-congruence}.
\end{proof}

\begin{proposition}[Endpoint exactness of the syntactic quotient]
\label{prop:endpoint-exactness}
For \(t,u\in\Expr_n\),
\[
  \operatorname{Nonempty}
    \bigl(\Tr_{\Q_n}(\den t,\den u)\bigr)
       \quad\Longleftrightarrow\quad t\defeq u.
\tag{9.1}\label{eq:exact-endpoint}
\]
\end{proposition}

\begin{proof}
Given a trace record, discard its list and retain
\(\den t=\den u\); quotient exactness gives \(t\defeq u\).
Conversely, \(h:t\defeq u\) gives a quotient equality by quotient soundness,
and Definition~\ref{def:trace-record} accepts the corresponding singleton
list and endpoint proof.
\end{proof}

\begin{remark}
Proposition~\ref{prop:endpoint-exactness} is tautological after defining
\(\Q_n\) as the quotient by \(\defeq\).  It is not a completeness or
external-semantic theorem.  The substantive results are
the source structural inductions, primitive-redex preservation, quotient-map
well-definedness, and source-to-target rewrite soundness collected in
Theorem~\ref{thm:raw-structural}.
\end{remark}

\section{\texorpdfstring{The metadata-fiber criterion for \(J\)}
{The metadata-fiber criterion for J}}
\label{sec:j-boundary}

\subsection{\texorpdfstring{\(J\) and based total-space contractibility}
{J and based total-space contractibility}}

Let \(A\in\mathcal U\) be a small set and let
\[
                    R:A\to A\to\mathcal U
\]
be any reflexivity-equipped family, with a chosen
\(r_x:R(x,x)\) for every \(x:A\).  No composition, inverse, or equality
projection is assumed.  At a fixed \(a:A\), define the based total space
\[
                B_a\coloneqq\sum_{b:A}R(a,b),
       \qquad c_a\coloneqq(a,r_a):B_a.
\tag{10.1}\label{eq:based-total-space}
\]
Write
\[
 \mathsf{Contr}_{c}(X)\coloneqq\prod_{x:X}(c=x),
 \qquad
 \mathsf{isContr}(X)\coloneqq\sum_{c:X}\mathsf{Contr}_{c}(X).
\tag{10.2}\label{eq:contractibility}
\]
For a specified \(c:X\), \(\mathsf{Contr}_c(X)\) is equivalent to
\(\mathsf{isContr}(X)\).  One direction chooses \(c\).  Conversely, from
\((d,k):\mathsf{isContr}(X)\) obtain
\((k(c))^{-1}\cdot k(x):c=x\).

\begin{definition}[Unrestricted based elimination with propositional beta]
\label{def:unrestricted-j}
Write \(\mathsf{J}^{\beta}_R(a)\) for the proposition that there exist
eliminator data
\[
 \mathcal J:
 \prod_{C:B_a\to\mathcal U}
 C(c_a)\to\prod_{x:B_a}C(x)
\tag{10.3}\label{eq:unrestricted-j}
\]
together with the propositional beta family
\[
 \beta_{C,d}:\mathcal J(C,d,c_a)=d
 \qquad(C:B_a\to\mathcal U,\ d:C(c_a)).
\tag{10.4}\label{eq:unrestricted-j-beta}
\]
In endpoint notation,
\(\mathcal J(C,d,(b,p))\) is written \(\mathcal J(C,d,b,p)\), so beta is
explicitly
\[
             \mathcal J(C,d,a,r_a)=d.
\]
The eliminator data and the proposition
\(\mathsf{J}^{\beta}_R(a)\) lie in \(\mathcal V\); each beta equation is a
small proof-irrelevant proposition.  Beta is propositional only.  No
judgmental beta rule is claimed.
\end{definition}

\begin{theorem}[Based elimination with propositional beta classifies
total-space contractibility]
\label{thm:j-contracts-total}
There are equivalences of \(\mathcal V\)-propositions
\[
 \mathsf{J}^{\beta}_R(a)
 \quad\Longleftrightarrow\quad
 \mathsf{Contr}_{c_a}(B_a)
 \quad\Longleftrightarrow\quad
 \mathsf{isContr}(B_a).
\tag{10.5}\label{eq:j-total-equivalence}
\]
\end{theorem}

\begin{proof}
Suppose first that \((\mathcal J,\beta)\) witnesses
\(\mathsf{J}^{\beta}_R(a)\).  Choose the motive
\[
              C(b,p)\coloneqq\bigl(c_a=(b,p)\bigr).
\tag{10.6}\label{eq:contraction-motive}
\]
Its reflexivity case is
\(\mathsf{refl}_{c_a}:C(a,r_a)\).  Define, for every \(b:A\) and
\(p:R(a,b)\),
\[
 \kappa_{b,p}\coloneqq
 \mathcal J(C,\mathsf{refl}_{c_a},b,p):c_a=(b,p).
\tag{10.7}\label{eq:contraction-path}
\]
Every \(x:B_a\) is a pair \((b,p)\); dependent-sum elimination therefore
defines
\(\kappa:\prod_{x:B_a}c_a=x\) by
\(\kappa(b,p)=\kappa_{b,p}\).  This gives
\(\mathsf{Contr}_{c_a}(B_a)\).  The supplied beta law also records the
coherence
\[
             \kappa(c_a)=\mathsf{refl}_{c_a}.
\tag{10.8}\label{eq:contraction-center-beta}
\]

Conversely, suppose
\(\kappa:\mathsf{Contr}_{c_a}(B_a)\).  For a
\(\mathcal U\)-valued motive \(C\), a base element
\(d:C(a,r_a)\), and \(x=(b,p):B_a\), define
\[
 J_{\kappa}(C,d,b,p)
 \coloneqq
 \mathsf{transport}_{\widetilde C}(\kappa(b,p),d):C(b,p),
\tag{10.9}\label{eq:j-from-contraction}
\]
where \(\widetilde C(b,p)=C(b,p)\) is regarded as a family on \(B_a\).
At the center, proof irrelevance gives
\(\kappa(c_a)=\mathsf{refl}_{c_a}\).  Congruence of transport in its
equality argument and~\eqref{eq:transport-beta} give the propositional beta
equation
\[
 J_\kappa(C,d,a,r_a)
 =\mathsf{transport}_{\widetilde C}(\kappa(c_a),d)
 =d.
\tag{10.10}\label{eq:contraction-derived-beta}
\]
Thus \((J_\kappa,\beta^\kappa)\) witnesses
\(\mathsf{J}^{\beta}_R(a)\); no judgmental equality was used.  The final equivalence with
\(\mathsf{isContr}(B_a)\) is the specified-center argument following
\eqref{eq:contractibility}.
\end{proof}

\begin{corollary}[Observable-metadata obstruction]
\label{cor:observable-obstruction}
Each of the following shows that \(R\) does not admit an unrestricted based
eliminator satisfying propositional beta, i.e. refutes
\(\mathsf{J}^{\beta}_R(a)\):
\begin{enumerate}
\item an element \(x:B_a\) with \(c_a\neq x\);
\item in particular, \(p:R(a,a)\) with
  \((a,r_a)\neq(a,p)\);
\item a predicate \(O:B_a\to\mathsf{Prop}\) and \(x:B_a\) such that
  \(O(c_a)\) and \(\neg O(x)\).
\end{enumerate}
\end{corollary}

\begin{proof}
In the first two cases, Theorem~\ref{thm:j-contracts-total} supplies
\(c_a=x\), contradicting the hypothesis.  In the third case one may either
transport \(O(c_a)\) along that equality or argue directly: choose the motive
\(C(b,p)=O(b,p)\) for the eliminator~\eqref{eq:unrestricted-j}; its base case yields
\(O(x)\), contradicting \(\neg O(x)\).
\end{proof}

Theorem~\ref{thm:j-contracts-total} is standard background: it abstracts the
familiar contractibility proof for the based total identity space.  The next
step identifies exactly what it says for equality enriched by metadata.

\subsection{Equality plus metadata}
\label{sec:metadata-classification}

Fix \(a:A\).  Let
\[
 M:\prod_{b:A}(a=b)\to\mathcal U
\]
be a metadata family and choose
\(m_0:M(a,\mathsf{refl}_a)\).  Define
\[
 R_M(a,b)\coloneqq\sum_{h:a=b}M(b,h),
 \qquad
 r^M_a\coloneqq(\mathsf{refl}_a,m_0),
\tag{10.11}\label{eq:metadata-refinement}
\]
and
\[
 B_M\coloneqq
 \sum_{b:A}\sum_{h:a=b}M(b,h).
\tag{10.12}\label{eq:metadata-total-space}
\]
Only the row based at the fixed \(a\) is required by
Definition~\ref{def:unrestricted-j}; no refinement data at another base point
is being inferred.

\begin{lemma}[The based equality total space]
\label{lem:based-equality-total}
The small set
\[
 P_a\coloneqq\sum_{b:A}(a=b)
\]
is contractible with center
\(s_0=(a,\mathsf{refl}_a)\).
\end{lemma}

\begin{proof}
Equality elimination on \(h:a=b\) defines
\[
 \lambda(b,h).\,\kappa_{b,h}:
 \prod_{(b,h):P_a}s_0=(b,h),
\]
with
\(\kappa_{a,\mathsf{refl}_a}=\mathsf{refl}_{s_0}\).
This is a contraction centered at \(s_0\).
\end{proof}

\begin{theorem}[The metadata total space is its reflexivity fiber]
\label{thm:metadata-total-equivalence}
There is an equivalence
\[
       B_M\simeq M(a,\mathsf{refl}_a)
\tag{10.13}\label{eq:metadata-equivalence}
\]
that sends \((a,\mathsf{refl}_a,m)\) to \(m\).
\end{theorem}

\begin{proof}
Regard \(M\) as the family
\(\widetilde M:P_a\to\mathcal U\) given by
\(\widetilde M(b,h)=M(b,h)\), and use the contraction
\(\kappa_{b,h}:s_0=(b,h)\) above.  Define
\[
 \begin{aligned}
 F:B_M&\longrightarrow M(a,\mathsf{refl}_a),&
 F(b,h,m)&=
 \mathsf{transport}_{\widetilde M}(\kappa_{b,h}^{-1},m),\\
 G:M(a,\mathsf{refl}_a)&\longrightarrow B_M,&
 G(m)&=(a,\mathsf{refl}_a,m).
 \end{aligned}
\tag{10.14}\label{eq:metadata-equivalence-maps}
\]
Equality elimination makes
\(\kappa_{a,\mathsf{refl}_a}\) reflexivity, hence
\(F(G(m))=m\).  For \(G(F(b,h,m))=(b,h,m)\), eliminate on
\(h:a=b\).  The remaining case is
\[
 G(F(a,\mathsf{refl}_a,m))
 =(a,\mathsf{refl}_a,m),
\]
which is reflexivity.  Proof irrelevance ensures there is no further
equality-proof case.  Thus \(F\) and \(G\) are inverse.
\end{proof}

\begin{lemma}[Contractibility is invariant under equivalence]
\label{lem:contractibility-equivalence}
If \(F:X\simeq Y\), then
\(\mathsf{isContr}(X)\) iff \(\mathsf{isContr}(Y)\).
\end{lemma}

\begin{proof}
Let \(G\) be an inverse.  If \((c,k)\) contracts \(X\), then \(F(c)\)
contracts \(Y\): for \(y:Y\), apply \(F\) to \(k(Gy):c=Gy\) and compose
with the inverse law \(F(Gy)=y\).  The converse exchanges \(F\) and \(G\).
\end{proof}

\begin{theorem}[Metadata-fiber criterion for \(J\)]
\label{thm:metadata-fiber-criterion}
The refinement~\eqref{eq:metadata-refinement} admits an unrestricted based
eliminator satisfying propositional beta exactly when its reflexivity
metadata fiber is contractible:
\[
 \mathsf{J}^{\beta}_{R_M}(a)
 \quad\Longleftrightarrow\quad
 \mathsf{isContr}\bigl(M(a,\mathsf{refl}_a)\bigr).
\tag{10.15}\label{eq:metadata-fiber-criterion}
\]
The equivalence is read in \(\mathcal V\); the contractibility witness itself
is \(\mathcal U\)-small.
\end{theorem}

\begin{proof}
Theorem~\ref{thm:j-contracts-total} identifies the left side with
\(\mathsf{isContr}(B_M)\).
Theorem~\ref{thm:metadata-total-equivalence} and
Lemma~\ref{lem:contractibility-equivalence} identify that proposition with
contractibility of the reflexivity metadata fiber.
\end{proof}

\subsection{The available equality-component eliminator}

The metadata refinement has the projection
\[
 q_{a,b}:R_M(a,b)\to(a=b),\qquad q(h,m)=h,
\tag{10.16}\label{eq:equality-projection}
\]
with \(q(r^M_a)=\mathsf{refl}_a\).  For a family \(D(b,h)\) depending only
on the equality component, ambient equality elimination defines
\[
 J_q(D,d,b,p):D(b,q(p))
\tag{10.17}\label{eq:general-equality-component-j}
\]
from \(d:D(a,\mathsf{refl}_a)\), with propositional beta
\[
                J_q(D,d,a,r^M_a)=d.
\]
This is not unrestricted
\(\mathsf{J}^{\beta}_{R_M}(a)\): the motive is forced to factor through
\(q\), although it has its own propositional reflexivity equation.
For the trace record, this projection is
\[
             \pi:\Tr_X(a,b)\longrightarrow(a=b),
             \qquad \pi(L,e)=e.
\tag{10.18}\label{eq:forget-trace}
\]
Let \(C(b,e)\) be a family depending on an endpoint \(b:X\) and an ambient
equality \(e:a=b\), but not on a trace list.  Given
\(c:C(a,\mathsf{refl}_a)\), ordinary elimination on the equality component
defines
\[
  J_{=}(C,c,P):C(b,\pi(P))
             \qquad(P:\Tr_X(a,b)).
\tag{10.19}\label{eq:proof-insensitive-j}
\]
It satisfies the reflexivity equation at the empty reflexive trace:
\[
             J_{=}(C,c,\mathbf r_a)=c.
\tag{10.20}\label{eq:j-beta}
\]
This beta equation is propositional, not judgmental.
Moreover, for any \(P,Q:\Tr_X(a,b)\), proof irrelevance of the equality
components gives
\[
             J_{=}(C,c,P)=J_{=}(C,c,Q).
\tag{10.21}\label{eq:j-trace-irrelevance}
\]
This is an \emph{available equality-component eliminator with propositional
beta}.  We do not claim
that the syntactic motive \(M\) of raw \(\EqJ\) is interpreted by \(C\), or
that raw beta is proved through this construction.

\subsection{Trace metadata}

Let
\[
 \mathbf s_a=
 \Bigl([\, (a,a,\mathsf{refl}_a)\,],\mathsf{refl}_a\Bigr)
 :\Tr_X(a,a)
\tag{10.22}\label{eq:traced-refl}
\]
be the explicit one-step reflexive trace.  It is well formed even under the
usual intrinsic chain conditions: its sole step starts and ends at \(a\),
its outer endpoints are \(a,a\), and its composite equality is reflexivity.
It differs from \(\mathbf r_a\) only by observable, noncanonical metadata.

\begin{theorem}[Trace instance of the metadata-fiber criterion]
\label{thm:no-trace-j}
For every \(X\) and \(a:X\), \(\Tr_X\) with reflexivity
\(\mathbf r\) does not admit an unrestricted based eliminator satisfying
propositional beta.
\end{theorem}

\begin{proof}
Define the metadata family
\[
 M_X(b,h)\coloneqq
   \operatorname{List}(\mathsf{QStep}(X)).
\]
The trace record is equivalent to the associated refinement by the explicit
maps
\[
\begin{aligned}
 S_{a,b}:\Tr_X(a,b)&\longrightarrow\sum_{h:a=b}M_X(b,h),
 &S_{a,b}(L,e)&=(e,L),\\
 T_{a,b}:\sum_{h:a=b}M_X(b,h)&\longrightarrow\Tr_X(a,b),
 &T_{a,b}(h,L)&=(L,h).
\end{aligned}
\]
For every \((h,L)\),
\[
 S_{a,b}(T_{a,b}(h,L))=(h,L),
\]
and for every \((L,e)\),
\[
 T_{a,b}(S_{a,b}(L,e))=(L,e).
\]
Both equalities are reflexivity of ordered pairs, so \(S_{a,b}\) and
\(T_{a,b}\) exhibit the required equivalence in the sense fixed in
Section~\ref{sec:ambient}.
The reflexivity metadata is the empty list.  Its fiber
\(M_X(a,\mathsf{refl}_a)\) also contains the singleton
\([\, (a,a,\mathsf{refl}_a)\,]\), distinct from the empty list by
no-confusion.  A contractible set has all elements equal, so
Theorem~\ref{thm:metadata-fiber-criterion} refutes the existence of an
unrestricted based eliminator satisfying propositional beta.
\end{proof}

Intrinsic composability does not repair the obstruction, because
\(\mathbf s_a\) is already a well-formed composable trace with reflexive
composite.  Any intrinsic metadata family that still contains distinct empty
and singleton reflexive values has a noncontractible reflexivity fiber and
fails the same criterion.  An unrestricted based eliminator satisfying
propositional beta requires that fiber to become contractible, for example
by hiding or quotienting metadata.  Alternatively,
restricting motives to factor through~\eqref{eq:forget-trace} retains the
available equality-component eliminator.

\section{An independent fixed-universe family algebra}
\label{sec:family-semantics}

This section records an independent algebraic example.  It is not a
semantics of the raw calculus and is not used in
Theorem~\ref{thm:raw-structural}.  Use the fixed universe
\(\mathcal U\) of Section~\ref{sec:ambient}.  Take a context-like object to be a small set
\(\Gamma:\mathcal U\), a substitution \(\Delta\to\Gamma\) to be a function,
a family over \(\Gamma\) to be \(A:\Gamma\to\mathcal U\), and a section to
be
\[
                  a:\prod_{\gamma:\Gamma}A(\gamma).
\]
Extension is the explicitly defined dependent sum
\(\Ext{\Gamma}{A}=\sum_{\gamma:\Gamma}A(\gamma)\).  Substitution is
precomposition:
\[
\begin{aligned}
A[\sigma](\delta)&=A(\sigma\delta),\\
a[\sigma](\delta)&=a(\sigma\delta).
\end{aligned}
\tag{11.1}\label{eq:family-substitution}
\]
Consequently identity and composition of substitution on contexts, types,
and terms hold definitionally.

Products and sums are interpreted pointwise:
\[
\begin{aligned}
(\Pi_A B)(\gamma)&=\prod_{x:A(\gamma)}B(\gamma,x),\\
(\Sigma_A B)(\gamma)&=\sum_{x:A(\gamma)}B(\gamma,x).
\end{aligned}
\]
Lambda, application, pairing, and projection are the corresponding
metatheoretic operations.  Application/lambda beta and both projection/pair
beta equations are definitional.  Eta is different: equality of sections
such as
\[
 \lambda\gamma.\,(\pi_1 t(\gamma),\pi_2 t(\gamma))=t
 \quad\text{and}\quad
 \lambda\gamma\,x.\,f(\gamma)(x)=f
\]
is propositional/extensional in the ambient metatheory, using pair equality
and function extensionality unless a chosen metatheory happens to make a
particular eta law definitional.  Eta is not included among the definitional
claims below.

Natural numbers are represented by a universe lift of the metatheoretic
natural numbers.  The motive and branches are defined pointwise, so zero,
successor, and substitution equations of the recursor are definitional.  A
small code in this fixed algebra is simply a record containing a carrier in
\(\mathcal U\).  Independent code operations for lifted naturals, dependent
products, dependent sums, and trace records decode definitionally.

For sections \(a,b:\prod_{\gamma:\Gamma}A(\gamma)\), define
\[
              \mathsf{Id}^{\mathrm{tr}}_A(a,b)(\gamma)
                             =\Tr_{A(\gamma)}(a(\gamma),b(\gamma)).
\tag{11.2}\label{eq:family-path}
\]
Reflexivity is the empty trace.  Substitution stability follows
definitionally from precomposition.  The equality-component eliminator is
available pointwise with propositional beta by
Section~\ref{sec:j-boundary}; unrestricted trace-sensitive elimination with
propositional beta is not.

Definitionally, the operations satisfy substitution identity and
composition; substitution stability of \(\Pi,\Sigma\), lifted naturals,
decoded codes, and~\eqref{eq:family-path}; product/sum beta; natural-number
beta; and the listed decoding equations.

There is no map interpreting raw \(\Sort_\ell\), raw codes, contexts, or
derivations into this algebra, and no relation between raw level \(\ell\) and
its single universe lift.  A comparison theorem would have to cover
assignments, levels, codes, typing, all ten redexes, and de Bruijn
substitution.  None is supplied here.

\section{Boundaries and possible strengthenings}
\label{sec:limitations}

\paragraph{Source and reduction.}
The source has only the constructors in Section~\ref{sec:syntax}.  Contexts
are same-scope assignments, with no context-formation judgment, typed
\(\defeq\), or conversion rule.  The paper proves substitution and
preservation for ten top-level redexes, not contextual or multi-step subject
reduction, normalization, confluence, canonicity, or decidable type checking.

\paragraph{Identity programs and endpoints.}
\(\IProg\) is untyped and endpoint-trivial.  We prove only the homomorphic
simulation \(\srweq\Rightarrow\TrEq\), with no converse, completeness, or
claim that every trace is evaluated.  Distinct empty and singleton metadata
does not refute UIP for ambient equality; it shows that the metadata fiber is
noncontractible.  The singleton is intrinsically composable, so
composability alone cannot produce an unrestricted based eliminator
satisfying propositional beta.

\paragraph{Quotient and external algebra.}
\(\Tr_X(a,b)\) presupposes \(a=b\), and endpoint exactness is immediate from
the chosen quotient.  The fixed-universe algebra is not a raw
interpretation.  Such an interpretation would have to cover assignments,
levels, codes, typing, and all redexes.  Univalence, higher inductive types,
and a pentagon equality between rewrite derivations are not claimed.

\section{Comparison with prior computational-path results}
\label{sec:comparison}

Table~\ref{tab:comparison} compares statements, not implementation size or
historical importance.  ``Not stated in this form'' is deliberately weaker
than a priority claim: the table records how the cited theorems are used and
how the present statements differ.

The comparison uses identifiable portions of the sources.  In the 2016
article, Section~2 treats normal forms for equality proofs, Section~3 gives
labelled deduction, termination, and rewrite equality, and Section~4 treats
groupoid structure and UIP
\cite{deQueirozEtAl2016}.  In the 2017 article, Sections~2--3 introduce
computational paths and the path-based identity construction, while
Section~4 establishes groupoid laws and sketches higher structure
\cite{RamosEtAl2017}.  The 2018 article reviews paths and identity in
Sections~2--3, presents its path term-rewriting system in Section~4, and
develops type-theoretic and homotopical constructions in Section~5
\cite{RamosEtAl2018}.  In the 2026 monograph, Chapter~2 develops labelled
natural deduction, Chapter~3 computational paths, Chapter~4 the path rewrite
system, Chapters~6--8 the groupoid, higher, and weak
\(\omega\)-groupoid structures, Chapter~9 \(J\), UIP, and transport, and
Chapter~10 applications
\cite{RamosEtAl2026}.  Those are label-sensitive developments.  This note
instead uses an unlabelled quotient-trace target as a source-separation
experiment.  We do not claim that those works contain no substitution
results; the claim is only that the same-scope raw interface assembled here
is isolated and proved as the particular theorem package of
Theorem~\ref{thm:raw-structural}.

\begin{table}[ht]
\centering
\scriptsize
\renewcommand{\arraystretch}{1.18}
\begin{tabular}{>{\raggedright\arraybackslash}p{0.20\textwidth}
                >{\raggedright\arraybackslash}p{0.33\textwidth}
                >{\raggedright\arraybackslash}p{0.38\textwidth}}
\toprule
Present statement & Prior labelled-path results & Exact comparison\\
\midrule
Scope-indexed \(\Ren/\Sub\) algebra
& The 2016 and 2017 papers develop labelled equality reasons and their
  identity-type reading; the 2018 paper makes path operations explicit
  \cite{deQueirozEtAl2016,RamosEtAl2017,RamosEtAl2018}.
& No absence claim is made about substitution results in those works.
  This paper isolates and proves the specific same-scope de Bruijn
  \(\Ren(n,m)\), \(\Sub(n,m)\), double-lift, and instantiation interface used
  by its theorem package.\\

Ten top-level primitive-redex cases
& The prior calculus studies equality rewrites and computational-path
  reductions; the monograph gives a comprehensive labelled calculus
  \cite{RamosEtAl2026}.
& Theorem~\ref{thm:redex-preservation} is narrower: ten raw constructors,
  no contextual closure, no multi-step theorem, and only untyped
  \(B\defeq A\) coherence between displayed types.\\

Source identity rewrite soundness
& Groupoid and rewrite laws for label-sensitive computational paths are
  central in the earlier papers and book
  \cite{RamosEtAl2017,RamosEtAl2018,deVerasEtAl2025,RamosEtAl2026}.
& Theorem~\ref{thm:identity-soundness} is a separation result for the
  present source grammar.  Its target is only the eight-rule core plus
  congruence on unlabelled quotient traces.\\

Endpoint exactness
& Earlier work relates labelled paths and equality/identity judgments
  \cite{deQueirozEtAl2016,RamosEtAl2017}.
& Proposition~\ref{prop:endpoint-exactness} is not a strengthening of those
  results.  It is tautological exactness for the quotient by \(\defeq\).\\

Metadata-fiber criterion and trace obstruction
& Earlier identity-type discussions concern the labelled computational-path
  programme; the book develops the broader calculus
  \cite{RamosEtAl2017,RamosEtAl2026}.
& The standard background total-space contraction is refined by
  Theorem~\ref{thm:metadata-fiber-criterion}:
  an unrestricted based eliminator satisfying propositional beta for \(R_M\)
  exists exactly when
  \(M(a,\mathsf{refl})\) is contractible.
  Theorem~\ref{thm:no-trace-j} applies the criterion to visible lists; it
  does not preclude an unrestricted based eliminator with propositional beta
  for a quotient or hidden-label identity type.\\

Information carried by a path
& A labelled path retains a computational reason and rewrites that reason
  \cite{RamosEtAl2018,RamosEtAl2026}.
& A trace atom here stores only
  \((\den t,\den u,\bar h)\); the source derivation \(h\) is erased.
  Evaluated programs decorate already quotient-collapsed endpoints.\\
\bottomrule
\end{tabular}
\caption{Statement-by-statement comparison with the labelled
computational-path literature.}
\label{tab:comparison}
\end{table}

\section{Related work}
\label{sec:related}

\subsection{Syntax and categorical semantics of dependent type theory}

The scope representation follows de Bruijn's nameless notation
\cite{deBruijn1972}; the judgmental organization follows the intensional
tradition of Martin-L\"of~\cite{MartinLof1984}.  Contextual categories and generalized
algebraic theories provide a categorical account of contexts and
substitution~\cite{Cartmell1986}.  Dybjer's internal type theory
\cite{Dybjer1996} and Hofmann's syntax--semantics account
\cite{Hofmann1997} are particularly relevant to the separation between raw
substitution and semantic substitution.  Uemura's general framework
\cite{Uemura2023} gives a much broader organization of theories and their
models.  We do not establish the initiality or universal-model properties
studied in that line of work; \(\Q_n\) is only an explicitly constructed
syntactic quotient.

\subsection{Identity types and homotopical semantics}

The Hofmann--Streicher groupoid interpretation
\cite{HofmannStreicher1998} demonstrated that intensional identity types need
not collapse to proof-irrelevant equality.  Homotopy-theoretic models
\cite{AwodeyWarren2009} and the identity-type weak factorization system
\cite{GambinoGarner2008} place identity elimination in categorical and
homotopical structure.  Lumsdaine~\cite{Lumsdaine2010} and van den Berg and
Garner~\cite{vanDenBergGarner2011} establish weak higher-groupoid structure
generated by intensional identity types.  The HoTT Book
\cite{HoTTBook2013} develops this perspective together with univalence and
higher inductive types.

Theorem~\ref{thm:j-contracts-total} abstracts the familiar path-induction
proof that the based total identity space
\(\sum_{b:A}(a=b)\) is contractible
\cite{HoTTBook2013}.  The present note then identifies the total space of an
equality-plus-metadata refinement with its reflexivity metadata fiber and
applies that criterion to quotient-trace records.

Our target is different in a decisive respect: it is a record consisting of
proof-irrelevant equality and an observable trace list.  The no-go theorem
in Section~\ref{sec:j-boundary} explains why the higher-groupoid results for
ordinary identity types cannot simply be transferred to this record.  The
sound groupoid rewrites of Section~\ref{sec:identity-programs} concern
evaluations of source identity programs, not a derivation of an
identity-type eliminator.

\subsection{Strict and two-level equality}

Two-level systems distinguish a strict equality suitable for coherence from
a fibrant or homotopical equality.  Capriotti's models
\cite{Capriotti2016}, the strict-equality extension of Altenkirch, Capriotti,
and Kraus~\cite{AltenkirchCapriottiKraus2016}, and the systematic treatment
by Annenkov, Capriotti, Kraus, and Sattler~\cite{AnnenkovEtAl2023} provide
the relevant comparison.  Our layered table is not itself a two-level type
theory: ambient equality, source definitional equality, source identity
programs, and target trace rewrites are metatheoretic layers, not two
internal equality formers with a fibrancy discipline.  Nevertheless, the
two-level literature reinforces the methodological lesson that equality
notions and their elimination rules must be kept explicit.

\subsection{Computational paths}

The 2016 and 2017 articles
\cite{deQueirozEtAl2016,RamosEtAl2017} connect labelled equality reasons
with identity types; the 2018 treatment makes computational paths explicit
\cite{RamosEtAl2018}; and later work develops weak groupoid structure
\cite{deVerasEtAl2025}.  The monograph
\cite{RamosEtAl2026} is the comprehensive background source for the
label-sensitive calculus, its history, rewriting operations, and
applications.  The present target deliberately erases those labels at
quotient soundness.  The structural results and this change of target are
summarized in Table~\ref{tab:comparison}; no claim of absolute priority is
intended.

\section{Reproducibility}
\label{sec:reproducibility}

The mathematical proofs were checked against a machine-verifiable
development in Lean~4~\cite{deMouraUllrich2021,ScopedTraceSoftware2026}.
The artifact covers the scoped syntax, structural laws, raw judgments,
primitive-redex preservation, quotient substitution, source identity rewrite
soundness, endpoint exactness, the independent fixed-universe algebra, and
the concrete trace-metadata obstruction.  It does not currently formalize
the abstract total-space equivalence or metadata-fiber classification
(Theorems~\ref{thm:j-contracts-total} and
\ref{thm:metadata-fiber-criterion}), whose complete mathematical proofs are
given in Section~\ref{sec:j-boundary}.

\medskip
\noindent
\fbox{\parbox{0.94\linewidth}{
\textbf{Software snapshot:}
repository snapshot at commit \texttt{70ee53ee}
(full commit
\texttt{70ee53eeac604d7430aaa4348abc1ff53936b0cf}).
A future Zenodo archival DOI may supplement this stable commit reference in
the final version.}}

\medskip
The software is evidence for the listed inductions; it does not supply the
missing context-formation, contextual reduction, or external interpretation
theorems.

\section{Conclusion}
\label{sec:conclusion}

The paper establishes a scoped syntax and its binder algebra, typed
substitution, preservation for ten top-level primitive redexes, compositional
substitution on a syntactic quotient, and sound evaluation of source identity
rewrites into an explicitly presented unlabelled trace relation.  None of
these statements requires calling the quotient a model of MLTT.

Endpoint exactness is immediate from quotient soundness and exactness.  Its
role is bookkeeping, not semantic completeness.  More structurally
informative is the separation of labels: the quotient trace records the
existence and arrangement of equality steps but erases the source
\(\defeq\)-derivations that produced them.  It is therefore different from
the label-sensitive computational paths of the earlier literature.

Raw \(\EqJ\)-beta and the external equality-component eliminator are also
separate results.  More generally, for
\(R_M(a,b)=\sum_{h:a=b}M(b,h)\), an unrestricted based eliminator satisfying
propositional beta exists exactly when the reflexivity metadata fiber
\(M(a,\mathsf{refl})\) is contractible.
The empty/singleton trace fiber is one noncontractible instance, while
equality-component motives remain available through projection.  Future work
may construct a genuine raw-to-CwF
interpretation and contextual subject reduction, or may hide or quotient
trace metadata.  Each extension must state which equality layer and which
elimination principle it uses.

\appendix
\small

\section{Raw grammar and judgment summary}
\label{app:grammar}

\subsection{Scope-indexed constructors}

The complete scope information is as follows.  In each row, unmarked
arguments live in \(\Expr_n\).
\[
\begin{array}{rcl}
\Var &:& \Fin(n)\to\Expr_n,\\
\Sort_\ell&:&\Expr_n,\\
\PiTy,\SigmaTy&:&\Expr_n\times\Expr_{n+1}\to\Expr_n,\\
\Lam&:&\Expr_{n+1}\to\Expr_n,\\
\App,\Pair&:&\Expr_n\times\Expr_n\to\Expr_n,\\
\Fst,\Snd&:&\Expr_n\to\Expr_n,\\
\NatTy_\ell,\Zero_\ell&:&\Expr_n,\\
\Succ_\ell&:&\Expr_n\to\Expr_n,\\
\NatElim_\ell&:&
  \Expr_{n+1}\times\Expr_n\times\Expr_{n+2}\times\Expr_n
  \to\Expr_n,\\
\CodeNat_\ell&:&\Expr_n,\\
\CodePi,\CodeSigma&:&\Expr_n\times\Expr_{n+1}\to\Expr_n,\\
\CodeId&:&\Expr_n^3\to\Expr_n,\\
\El&:&\Expr_n\to\Expr_n,\\
\IdTy&:&\Expr_n^3\to\Expr_n,\\
\Refl&:&\Expr_n\to\Expr_n,\\
\EqJ&:&\Expr_{n+2}\times\Expr_n^3\to\Expr_n.
\end{array}
\tag{A.1}\label{eq:full-grammar}
\]

\subsection{Renaming and substitution clauses under binders}

The nontrivial recursion clauses are
\[
\begin{aligned}
\PiTy(A,B)[\rho]_{\mathrm r}
 &=\PiTy(A[\rho]_{\mathrm r},B[\rho^\lift]_{\mathrm r}),\\
\Lam(t)[\rho]_{\mathrm r}
 &=\Lam(t[\rho^\lift]_{\mathrm r}),\\
\NatElim_\ell(M,z,s,k)[\rho]_{\mathrm r}
 &=\NatElim_\ell(M[\rho^\lift]_{\mathrm r},
    z[\rho]_{\mathrm r},s[\rho^{\lift\lift}]_{\mathrm r},
    k[\rho]_{\mathrm r}),\\
\EqJ(M,r,b,p)[\rho]_{\mathrm r}
 &=\EqJ(M[\rho^{\lift\lift}]_{\mathrm r},
    r[\rho]_{\mathrm r},b[\rho]_{\mathrm r},p[\rho]_{\mathrm r}),
\end{aligned}
\]
with identical lift placement for substitution:
\[
\begin{aligned}
\PiTy(A,B)[\sigma]
 &=\PiTy(A[\sigma],B[\sigma^\lift]),\\
\Lam(t)[\sigma]&=\Lam(t[\sigma^\lift]),\\
\NatElim_\ell(M,z,s,k)[\sigma]
 &=\NatElim_\ell(M[\sigma^\lift],z[\sigma],
     s[\sigma^{\lift\lift}],k[\sigma]),\\
\EqJ(M,r,b,p)[\sigma]
 &=\EqJ(M[\sigma^{\lift\lift}],r[\sigma],
        b[\sigma],p[\sigma]).
\end{aligned}
\tag{A.2}\label{eq:binder-clauses}
\]
Product and sum codes use one lift in their second argument.

\subsection{Typing rules omitted from the main display}

The complete sum fragment is
\[
\frac{\Gamma\vdash A:\Sort_\ell\quad
      \Ext{\Gamma}{A}\vdash B:\Sort_\ell}
     {\Gamma\vdash\SigmaTy(A,B):\Sort_\ell},
\qquad
\frac{\Gamma\vdash a:A\quad\Gamma\vdash b:B\{a\}}
     {\Gamma\vdash\Pair(a,b):\SigmaTy(A,B)},
\]
\[
\frac{\Gamma\vdash p:\SigmaTy(A,B)}
     {\Gamma\vdash\Fst(p):A},
\qquad
\frac{\Gamma\vdash p:\SigmaTy(A,B)}
     {\Gamma\vdash\Snd(p):B\{\Fst(p)\}}.
\tag{A.3}\label{eq:sigma-rules}
\]

For natural elimination, define the successor-branch context
\[
 \Gamma_{\mathrm{succ}}
   =\Ext{\Ext{\Gamma}{\NatTy_\ell}}{M}.
\]
In its scope, let
\[
 T_{\mathrm{succ}}(M)
   =M[
      0\mapsto\Succ_\ell(\Var(1)),\
      i+1\mapsto\Var(i+2)
     ].
\tag{A.4}\label{eq:nat-step-target}
\]
If
\[
\Ext{\Gamma}{\NatTy_\ell}\vdash M:\Sort_j,\quad
\Gamma\vdash z:M\{\Zero_\ell\},\quad
\Gamma_{\mathrm{succ}}\vdash s:T_{\mathrm{succ}}(M),\quad
\Gamma\vdash k:\NatTy_\ell,
\]
then
\[
\Gamma\vdash\NatElim_\ell(M,z,s,k):M\{k\}.
\tag{A.5}\label{eq:nat-elim-full}
\]
Two-variable instantiation of~\eqref{eq:nat-step-target} by \(k\) and an
induction hypothesis is \(M\{\Succ_\ell(k)\}\), the key equation in the
successor primitive-redex preservation case.

Code formation and decoding are
\[
\frac{}{\Gamma\vdash\CodeNat_\ell:\Sort_\ell},
\quad
\frac{\Gamma\vdash A:\Sort_\ell\quad
      \Ext{\Gamma}{\El(A)}\vdash B:\Sort_\ell}
     {\Gamma\vdash\CodePi(A,B):\Sort_\ell},
\]
\[
\frac{\Gamma\vdash A:\Sort_\ell\quad
      \Ext{\Gamma}{\El(A)}\vdash B:\Sort_\ell}
     {\Gamma\vdash\CodeSigma(A,B):\Sort_\ell},
\]
\[
\frac{\Gamma\vdash A:\Sort_\ell\quad
      \Gamma\vdash a:\El(A)\quad\Gamma\vdash b:\El(A)}
     {\Gamma\vdash\CodeId(A,a,b):\Sort_\ell},
\qquad
\frac{\Gamma\vdash c:\Sort_\ell}
     {\Gamma\vdash\El(c):\Sort_\ell}.
\tag{A.6}\label{eq:code-rules}
\]
The identity rules and raw two-binder eliminator are those in
\eqref{eq:eqj-rule}.

\subsection{Definitional-equality compatible closure}

Besides equivalence and the beta inclusions, \(\defeq\) is congruent in the
following positions:
\[
\begin{array}{ll}
\PiTy,\SigmaTy:&\text{domain and codomain},\\
\Lam:&\text{body},\\
\App:&\text{function and argument},\\
\Pair:&\text{first and second component},\\
\Fst,\Snd,\Succ_\ell,\El,\Refl:&\text{their unique argument},\\
\NatElim_\ell:&\text{motive, zero case, successor case, scrutinee},\\
\CodePi,\CodeSigma:&\text{domain and codomain},\\
\CodeId,\IdTy:&\text{type, left endpoint, right endpoint},\\
\EqJ:&\text{motive, reflexivity case, endpoint, proof}.
\end{array}
\tag{A.7}\label{eq:defeq-congruence}
\]
Binder positions live in their larger scopes.  Substitution closure applies
\(\sigma^\lift\) or \(\sigma^{\lift\lift}\) in precisely those positions.

\section{Source identity rewrite rules}
\label{app:rewrites}

\subsection{Source one-step judgment}

The judgment \(p\srw q\) is indexed by common source endpoints:
\[
 p,q:\IProg_n(t,u)\quad\Longrightarrow\quad
 (p\srw q)\ \mathsf{judgment}.
\]
For \(p:\IProg_n(t,u)\), \(q:\IProg_n(u,v)\), and
\(r:\IProg_n(v,w)\), its eight base constructors are
\[
\frac{}{(\mathbf1_t)^{-1}\srw\mathbf1_t}
\quad
\frac{}{(p^{-1})^{-1}\srw p}
\quad
\frac{}{\mathbf1_t\cdot p\srw p}
\quad
\frac{}{p\cdot\mathbf1_u\srw p},
\tag{B.1}\label{eq:app-source-units}
\]
\[
\frac{}{p\cdot p^{-1}\srw\mathbf1_t}
\quad
\frac{}{p^{-1}\cdot p\srw\mathbf1_u}
\quad
\frac{}{(p\cdot q)^{-1}\srw q^{-1}\cdot p^{-1}}
\quad
\frac{}{(p\cdot q)\cdot r\srw p\cdot(q\cdot r)}.
\tag{B.2}\label{eq:app-source-groupoid}
\]
The compatible constructors are
\[
\frac{p\srw p'}{p^{-1}\srw(p')^{-1}},
\qquad
\frac{p\srw p'}{p\cdot q\srw p'\cdot q},
\qquad
\frac{q\srw q'}{p\cdot q\srw p\cdot q'},
\tag{B.3}\label{eq:app-source-composition-congruence}
\]
\[
\frac{p\srw p'}{C[p]\srw C[p']},
\qquad
\frac{}{C[\mathbf1_t]\srw\mathbf1_{C[t]}}
\quad(C\in\Frame_n).
\tag{B.4}\label{eq:app-source-frame-congruence}
\]
No other source one-step constructor is implicit.

\subsection{Source rewrite equivalence}

At fixed \(t,u\), the judgment \(p\srweq q\) is inductively generated by
\[
\frac{}{p\srweq p},
\qquad
\frac{p\srw q}{p\srweq q},
\qquad
\frac{p\srweq q}{q\srweq p},
\qquad
\frac{p\srweq q\quad q\srweq r}{p\srweq r}.
\tag{B.5}\label{eq:app-source-equivalence}
\]
Thus \(\srweq\) is a derivation-carrying symmetric, reflexive, transitive
closure, not an equality between derivations.

\subsection{Target trace rules}

For \(S:\Tr_X(a,b)\), \(T:\Tr_X(b,c)\), and
\(U:\Tr_X(c,d)\), target one-step rewriting has exactly
\[
\frac{}{(\mathbf r_a)^{-1}\TrStep\mathbf r_a}
\quad
\frac{}{(S^{-1})^{-1}\TrStep S}
\quad
\frac{}{\mathbf r_a\cdot S\TrStep S}
\quad
\frac{}{S\cdot\mathbf r_b\TrStep S},
\tag{B.6}\label{eq:app-target-units}
\]
\[
\frac{}{S\cdot S^{-1}\TrStep\mathbf r_a}
\quad
\frac{}{S^{-1}\cdot S\TrStep\mathbf r_b}.
\]
\[
\frac{}{(S\cdot T)^{-1}\TrStep T^{-1}\cdot S^{-1}}
\quad
\frac{}{(S\cdot T)\cdot U\TrStep S\cdot(T\cdot U)}.
\tag{B.7}\label{eq:app-target-groupoid}
\]
Target equivalence \(S\TrEq T\) is generated by
\[
\frac{}{S\TrEq S},
\quad
\frac{S\TrStep T}{S\TrEq T},
\quad
\frac{S\TrEq T}{T\TrEq S},
\quad
\frac{S\TrEq T\quad T\TrEq U}{S\TrEq U},
\tag{B.8}\label{eq:app-target-equivalence}
\]
together with all four congruence rules
\[
\frac{S\TrEq T}{S^{-1}\TrEq T^{-1}},
\qquad
\frac{S\TrEq T}{S\cdot U\TrEq T\cdot U},
\qquad
\frac{T\TrEq U}{S\cdot T\TrEq S\cdot U},
\qquad
\frac{S\TrEq T}{f_\ast S\TrEq f_\ast T}.
\tag{B.9}\label{eq:app-target-congruence}
\]
The map operation \(f_\ast\) is
Definition~\ref{def:trace-record} and equation~\eqref{eq:trace-map};
in particular \(f_\ast\mathbf r_a=\mathbf r_{fa}\) by definition.

\subsection{Rule-by-rule soundness map}

Evaluation sends the four rules in~\eqref{eq:app-source-units} to
\eqref{eq:app-target-units}, the four rules
in~\eqref{eq:app-source-groupoid} to
\eqref{eq:app-target-groupoid}, and the first three rules
in~\eqref{eq:app-source-composition-congruence} to the first three congruence
rules in~\eqref{eq:app-target-congruence}.  Frame congruence uses the last
target congruence rule with \(f=\overline C\); frame-unit soundness is target
reflexivity after the definitional equation
\(\overline C_\ast\mathbf r_{\den t}=\mathbf r_{\den{C[t]}}\).
Finally, the four constructors in~\eqref{eq:app-source-equivalence} map to
the four constructors in~\eqref{eq:app-target-equivalence}.  This is the
complete induction for Theorem~\ref{thm:identity-soundness}.

\begingroup
\small
\bibliographystyle{amsplain}
\bibliography{refs}
\endgroup

\end{document}
