\documentclass[10pt,reqno]{amsart}

\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{booktabs}
\usepackage{array}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{url}
\def\UrlBreaks{\do\_\do\/\do\-}

\hypersetup{
  hidelinks,
  pdftitle={Equality with Observable Metadata: Diagnosis and Universal Quotient Repair},
  pdfauthor={Arthur Freitas Ramos, Ruy J. G. B. de Queiroz, Anjolina Grisi de Oliveira, and Dov M. Gabbay}
}
\allowdisplaybreaks
\emergencystretch=3em
\setlength{\abovedisplayskip}{7pt plus 2pt minus 3pt}
\setlength{\belowdisplayskip}{7pt plus 2pt minus 3pt}
\setlength{\abovedisplayshortskip}{4pt plus 2pt}
\setlength{\belowdisplayshortskip}{5pt plus 2pt minus 2pt}

\title[Diagnosis and Universal Quotient Repair]
  {Equality with Observable Metadata:\\
   Diagnosis, Universal Quotient Repair, and the
   \(\mathsf{PathRwQuot}\) Boundary}

\author{Arthur Freitas Ramos}
\author{Ruy J. G. B. de Queiroz}
\author{Anjolina Grisi de Oliveira}
\author{Dov M. Gabbay}
\renewcommand{\shortauthors}{Ramos et al.}

\subjclass[2020]{03B15, 03B38, 03B70, 18N45}
\keywords{identity elimination, path induction, observable metadata,
setoid quotient, contractibility, axiom K, computational paths, equality
traces}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{warning}[theorem]{Warning}

\newcommand{\U}{\mathcal U}
\newcommand{\V}{\mathcal V}
\newcommand{\isContr}{\mathsf{isContr}}
\newcommand{\Contr}{\mathsf{Contr}}
\newcommand{\Eqv}{\simeq}
\newcommand{\Jbeta}{\mathsf J^\beta}
\newcommand{\Meta}{\mathsf M}
\newcommand{\Tot}{\mathsf T}
\newcommand{\Tr}{\mathsf{Tr}}
\newcommand{\Step}{\mathsf{Step}}
\newcommand{\PathTy}{\mathsf{Path}}
\newcommand{\refl}{\mathsf{refl}}
\newcommand{\transport}{\mathsf{transport}}
\newcommand{\List}{\mathsf{List}}
\newcommand{\proj}{\pi_{=}}
\newcommand{\Unit}{\mathbf 1}
\newcommand{\Empty}{[\,]}
\newcommand{\SetTot}{\mathsf{SetoidTotal}}
\newcommand{\Ker}{\mathsf{Ker}}
\newcommand{\PRW}{\mathsf{PathRwQuot}}
\newcommand{\LocalK}{\mathsf K_{\mathrm{loc}}}
\newcommand{\Q}{\mathsf Q}

\begin{document}

\begin{abstract}
Suppose that equality from a fixed point \(a:A\) to \(b:A\) is refined by
observable metadata:
\[
  R_M(a,b)=\sum_{h:a=b}M(b,h),
  \qquad r^M_a=(\refl_a,m_0).
\]
We first diagnose exactly when this refinement supports unrestricted based
elimination into arbitrary motives with a propositional beta law:
\[
 \Jbeta_{R_M}(a,r^M_a)
 \quad\Longleftrightarrow\quad
 \isContr\bigl(M(a,\refl_a)\bigr).
\]
We then classify every ordinary setoid quotient repair.  If \(S_{b,h}\) is a
setoid on \(M(b,h)\) and \(m_0:M(a,\refl_a)\), then the quotient-repaired
refinement supports the same eliminator exactly when the reflexivity relation
\(S_{a,\refl_a}\) is total.  For an arbitrary carrier \(X\), the precise
empty-sensitive statement is
\[
 \isContr(X/S)\quad\Longleftrightarrow\quad
 \bigl(\mathsf{Inh}(X)\ \wedge\ \SetTot(S)\bigr).
\]
Thus every set-level repair supporting unrestricted \(J\) identifies all
reflexivity metadata; no visible distinction can survive there.  The
indiscrete relation is maximal as a relation, minimal in retained information,
and terminal---not initial---for quotient-map-preserving factorizations.

The classification extends beyond sigma-shaped refinements.  For a reflexive
projection \(q_b:R(b)\to(a=b)\), unrestricted elimination is equivalent to
contractibility of the reflexivity kernel
\(\Ker(q)=\{r:R(a)\mid q_a(r)=\refl_a\}\).  Applied to
\(\PRW\), this yields the exact local criterion: elimination exists if and only
if the genuine loop quotient is trivial, equivalently local quotient-level
axiom K holds.  The theorem says neither that rewrite quotienting always repairs
\(J\) nor that it always fails.

Finally we audit the computational-path examples.  Empty and singleton
reflexive raw paths are unequal records but are explicitly
\(\mathsf{RwEq}\)-equivalent.  The current ``circle'' is a one-constructor Lean
type and the current ``torus'' its product, so their genuine \(\PRW\)
fundamental groups are contractible.  The separately defined winding-expression
quotients are equivalent to \(\mathbb Z\) and \(\mathbb Z^2\), hence
noncontractible; consequently no equivalence can bridge the genuine and
synthetic objects under the present definitions.  A Lean development verifies
the quotient exactness, factorization, kernel, trace, contractibility, and
no-bridge results.  The scope remains proof-irrelevant ambient equality with
propositional beta, not full Martin-L\"of or homotopy type theory.
\end{abstract}

\maketitle

\section{Introduction}
\label{sec:introduction}

Equality evidence is often decorated.  A proof assistant may retain a
derivation history; an optimizer may attach a cost; a checker may keep a
certificate; a normalizer may record a log; and a rewriting calculus may expose
the sequence of elementary steps.  The decoration is useful precisely because
it is observable.  That same observability changes the elimination principle.

Fix \(a:A\).  Given a family
\[
        M:\prod_{b:A}(a=b)\longrightarrow\U
\]
and a chosen \(m_0:M(a,\refl_a)\), define
\[
 R_M(a,b)=\sum_{h:a=b}M(b,h),
 \qquad r^M_a=(\refl_a,m_0).
\tag{1.1}\label{eq:refinement}
\]
The projection
\[
       \proj:R_M(a,b)\longrightarrow(a=b),\qquad
       \proj(h,m)=h
\tag{1.2}\label{eq:projection}
\]
forgets the metadata.  Ordinary equality elimination can act through
\(\proj\).  The stronger question is whether \emph{every} motive on the
refined object can be eliminated from the single base case \(r^M_a\).

\begin{theorem}[Metadata-fiber criterion]
\label{thm:headline}
The refinement \(R_M\), based at \(r^M_a\), admits unrestricted based
elimination into all \(\U\)-valued motives with propositional beta if and only
if
\[
                        \isContr\bigl(M(a,\refl_a)\bigr).
\tag{1.3}\label{eq:headline}
\]
\end{theorem}

Theorem~\ref{thm:headline} is the organizing result of the paper.  Its two
directions are constructive at the level used here.  An eliminator contracts
the based total space by choosing the motive ``the base point equals the
current point.''  Conversely, a contraction transports the base datum to every
point.  The new classification step is that the total space of
\eqref{eq:refinement} is equivalent to the single reflexivity metadata fiber.
Consequently all endpoint and equality coordinates disappear, but observable
metadata over reflexivity remains.

This distinction matters.  The familiar total identity space
\[
                         \sum_{b:A}(a=b)
\tag{1.4}\label{eq:identity-total}
\]
is contractible by equality elimination.  That standard argument explains
ordinary based path induction.  It does not say that a larger space obtained
by retaining an arbitrary extra coordinate is contractible.  Theorem
\ref{thm:headline} measures exactly that extra coordinate.

Our main case study is the record
\[
 \PathTy_A(a,b)
   =\bigl\{\,L:\List(\Step(A)),\quad h:a=b\,\bigr\},
\tag{1.5}\label{eq:path-record}
\]
where \(L\) is a visible trace.  Reordering its fields gives
\[
 \PathTy_A(a,b)\Eqv
 \sum_{h:a=b}\List(\Step(A)).
\tag{1.6}\label{eq:path-refinement-intro}
\]
At reflexivity, the empty list and the singleton step
\([\,a\xrightarrow{\refl_a}a\,]\) differ.  The fiber is not contractible, so
unrestricted elimination for motives observing \(L\) cannot exist.  The
singleton is a valid chain from \(a\) to \(a\), hence imposing adjacency and
endpoint conditions does not repair the problem.

The diagnosis leads to two principled remedies.
\begin{enumerate}[label=(\roman*)]
\item Restrict motives to those factoring through
  \(\proj\).  These motives cannot observe the metadata and inherit equality
  elimination.
\item Quotient the metadata by explicit setoids until its reflexivity quotient
  is contractible.  The exact criterion developed below says that, for an
  inhabited reflexivity fiber, this happens precisely when the setoid is total.
\end{enumerate}
The first preserves the data but limits observation; the second changes the
identity-like object.

The second option has a sharp universal boundary.  Write
\(\Q_S(X)=X/S\) and
\[
 \SetTot(S)\quad:\!\!\Longleftrightarrow\quad
 \prod_{x,y:X}S(x,y).
\tag{1.7}\label{eq:setoid-total-intro}
\]
Quotient soundness proves that totality contracts \(\Q_S(X)\), while quotient
exactness recovers \(S(x,y)\) from equality of the quotient images.  Hence, for
a chosen \(x_0:X\),
\[
 \isContr(\Q_S(X))\quad\Longleftrightarrow\quad\SetTot(S).
\tag{1.8}\label{eq:quotient-intro}
\]
Without \(x_0\), inhabitance must appear as an additional conjunct.  Applied to
the reflexivity metadata fiber, this is a classification of \emph{all} ordinary
setoid repairs, not only a sufficient construction.

We also replace the sigma-specific viewpoint by a projection theorem.  A
reflexive map \(q_b:R(b)\to(a=b)\) determines
\[
 \Ker(q)=\sum_{r:R(a)}\bigl(q_a(r)=\refl_a\bigr),
\tag{1.9}\label{eq:kernel-intro}
\]
and equality induction gives
\(\sum_bR(b)\Eqv\Ker(q)\).  Consequently unrestricted elimination is
equivalent to contractibility of this kernel.  The \(\PRW\) application is
especially useful: its kernel is equivalent to the entire loop quotient, so
the exact condition is local quotient-level axiom K.

\paragraph{Contributions.}
The paper provides:
\begin{enumerate}[label=(\arabic*)]
\item an explicit equivalence between unrestricted based elimination with
  propositional beta and based total-space contractibility;
\item the total-space equivalence
  \(\sum_b\sum_hM(b,h)\Eqv M(a,\refl_a)\), yielding the
  metadata-fiber criterion;
\item the empty-sensitive theorem
  \(\isContr(X/S)\leftrightarrow(\mathsf{Inh}(X)\wedge\SetTot(S))\), and its
  specialization classifying every setoid repair of equality metadata;
\item the indiscrete repair together with the exact relation-inclusion
  criterion and unique terminal factorization direction;
\item the projection/kernel theorem and its
  \(\PRW\)/local-axiom-K specialization;
\item a factorization theorem for equality-projection motives, including
  motives fiberwise equivalent to strict pullbacks;
\item an explicit raw-versus-\(\mathsf{RwEq}\) reflexivity calculation; and
\item formal contractibility, noncontractibility, and no-bridge results
  separating the current genuine circle and torus loop quotients from their
  synthetic winding-expression presentations.
\end{enumerate}

\paragraph{Scope.}
We reason about proof-irrelevant ambient equality in an extensional
set/type-theoretic metatheory, matching the formal artifact.  We classify
\emph{propositional} beta only.  We do not present a full syntax or model of
Martin-L\"of type theory, do not assume or prove univalence or higher
inductive types, and do not identify \eqref{eq:path-record} with the HoTT
identity type.  The raw de Bruijn calculus previously developed alongside this
work is preserved as a separate companion manuscript
\cite{ScopedTraceSoftware2026}; Section~\ref{sec:formalization} describes the
division.

\paragraph{Organization.}
Section~\ref{sec:setting} fixes the metatheory.
Section~\ref{sec:total-space} proves the general total-space classification.
Section~\ref{sec:metadata} proves Theorem~\ref{thm:headline}.
Section~\ref{sec:setoid-repair} classifies setoid quotient repairs and proves
the indiscrete terminal factorization theorem.
Section~\ref{sec:projection-kernel} proves the projection/kernel theorem and
the exact \(\PRW\)/K criterion.
Section~\ref{sec:factorization} gives the factorizing-motive theorem.
Section~\ref{sec:traces} is the principal raw-trace case study, and
Section~\ref{sec:genuine-synthetic} audits the circle and torus definitions.
Section~\ref{sec:examples} classifies further metadata.  Sections
\ref{sec:comparison}--\ref{sec:conclusion} state boundaries, related work,
formal verification, and conclusions.

\section{Setting and definitions}
\label{sec:setting}

\subsection{Ambient equality and universes}

Fix a universe \(\U\) closed under dependent sums and products, equality,
finite lists, and the elementary constructions below.  A larger universe
\(\V\) contains collections quantifying over all \(\U\)-valued motives.
Thus an individual motive value and every equality proposition remain
\(\U\)-small, while the record of an operation accepting every such motive
may live in \(\V\).  No impredicative identification \(\U=\V\) is intended.

Ambient equality has elimination, transport, function extensionality, and
proof irrelevance.  For \(e:x=y\) and \(C:X\to\U\), write
\[
                  \transport_C(e,-):C(x)\longrightarrow C(y).
\]
Its reflexivity law is used propositionally:
\[
                  \transport_C(\refl_x,c)=c.
\tag{2.1}\label{eq:transport-beta}
\]
Proof irrelevance implies that any two witnesses of the same ambient equality
are equal.  This assumption is appropriate for Lean's \(\mathsf{Eq}\), which
lives in \(\mathsf{Prop}\).  It is also why the present theory must not be read
as a claim about arbitrary proof-relevant identity types.

An equivalence \(e:X\Eqv Y\) consists of maps
\[
 e:X\to Y,\qquad e^{-1}:Y\to X
\]
with specified inverse equations.  We use no univalence principle: an
equivalence is data, not an equality of types.

\begin{definition}[Contractibility]
For \(c:X\), define
\[
 \Contr_c(X)=\prod_{x:X}(c=x),\qquad
 \isContr(X)=\sum_{c:X}\Contr_c(X).
\tag{2.2}\label{eq:contractibility}
\]
\end{definition}

The orientation \(c=x\) is convenient for transport from the center.  If
\((c,\kappa):\isContr(X)\), then every two points are equal:
\[
             x=(\kappa x)^{-1}\cdot\kappa y=y.
\tag{2.3}\label{eq:all-equal}
\]
For a specified \(c_0:X\), the propositions
\(\Contr_{c_0}(X)\) and \(\isContr(X)\) are equivalent.  One direction chooses
\(c_0\); the other uses
\[
      c_0=(\kappa c_0)^{-1}\cdot\kappa x=x.
\tag{2.4}\label{eq:recenter}
\]

\subsection{Based relations and unrestricted elimination}

Let \(R:A\to\U\) be a relation row based at \(a\), and choose \(r_a:R(a)\).
Its based total space and center are
\[
             B_R=\sum_{b:A}R(b),\qquad c_R=(a,r_a).
\tag{2.5}\label{eq:based-total}
\]
Writing \(R(b)\) rather than \(R(a,b)\) keeps the fixed base point implicit.

\begin{definition}[Unrestricted based elimination with propositional beta]
\label{def:unrestricted}
The data \(\Jbeta_R(a,r_a)\) consist of an operation
\[
 \mathcal J:
   \prod_{C:B_R\to\U}
   C(c_R)\longrightarrow\prod_{x:B_R}C(x)
\tag{2.6}\label{eq:j-data}
\]
and equations
\[
       \beta_{C,d}:\mathcal J(C,d,c_R)=d
       \qquad(C:B_R\to\U,\ d:C(c_R)).
\tag{2.7}\label{eq:j-beta}
\]
\end{definition}

The adjective ``unrestricted'' refers only to the motive: \(C\) may inspect
the entire point \(x=(b,p)\), including every field of \(p\).  The beta law is
an ambient equality, not a judgmental reduction.  Definition
\ref{def:unrestricted} is based: \(a\) and \(r_a\) are fixed.

\section{Based elimination and total-space contractibility}
\label{sec:total-space}

\begin{theorem}[Total-space classification]
\label{thm:j-total}
For any based relation row \((R,r_a)\),
\[
        \Jbeta_R(a,r_a)
        \quad\Longleftrightarrow\quad
        \Contr_{c_R}(B_R)
        \quad\Longleftrightarrow\quad
        \isContr(B_R).
\tag{3.1}\label{eq:j-total}
\]
Here the first term means that the data of Definition
\ref{def:unrestricted} exist.
\end{theorem}

\begin{proof}
Assume \((\mathcal J,\beta)\).  Use the motive
\[
                         C(x)=(c_R=x).
\tag{3.2}\label{eq:contraction-motive}
\]
Its base datum is \(\refl_{c_R}:C(c_R)\).  Define
\[
       \kappa(x)=\mathcal J(C,\refl_{c_R},x):c_R=x.
\tag{3.3}\label{eq:kappa-from-j}
\]
This is an element of \(\Contr_{c_R}(B_R)\).  The supplied beta equation
also gives \(\kappa(c_R)=\refl_{c_R}\), although only the contraction itself
is needed for this direction.

Conversely, let \(\kappa:\Contr_{c_R}(B_R)\).  Given \(C\), \(d:C(c_R)\),
and \(x:B_R\), put
\[
        \mathcal J_\kappa(C,d,x)
        =\transport_C(\kappa(x),d):C(x).
\tag{3.4}\label{eq:j-from-kappa}
\]
At \(x=c_R\), proof irrelevance identifies
\(\kappa(c_R)\) with \(\refl_{c_R}\).  Congruence of transport in its equality
argument and~\eqref{eq:transport-beta} give
\[
 \mathcal J_\kappa(C,d,c_R)
 =\transport_C(\kappa(c_R),d)
 =\transport_C(\refl_{c_R},d)
 =d.
\tag{3.5}\label{eq:derived-beta}
\]
Thus \((\mathcal J_\kappa,\beta^\kappa)\) is the required data.
The final equivalence in~\eqref{eq:j-total} is the recentering argument
\eqref{eq:recenter}.
\end{proof}

\begin{remark}[What is standard]
\label{rem:standard-stage}
Theorem~\ref{thm:j-total} is the standard universal-property argument for a
pointed contractible space: sections of every family are determined by the
fiber over the center.  Applied to \(R(b)=(a=b)\), equality elimination proves
that~\eqref{eq:identity-total} is contractible.  Neither fact mentions
metadata.  The classification specific to decorated equality begins in
Section~\ref{sec:metadata}, where a second equivalence computes the enlarged
total space.
\end{remark}

\begin{corollary}[Observable-point obstruction]
\label{cor:observable-point}
If \(x:B_R\) satisfies \(x\ne c_R\), then \(\Jbeta_R(a,r_a)\) does not exist.
More generally, it is enough to give \(O:B_R\to\mathsf{Prop}\) with
\(O(c_R)\) and \(\neg O(x)\).
\end{corollary}

\begin{proof}
An eliminator gives \(c_R=x\) by~\eqref{eq:kappa-from-j}, contradicting the
first hypothesis.  For the predicate formulation, transport \(O(c_R)\) along
that equality.  Equivalently, use \(O\) itself as the motive in
\eqref{eq:j-data}.
\end{proof}

The corollary is useful for finding counterexamples, but it does not yet
classify a structured family.  For equality refinements, the entire
obstruction can be localized to one fiber.

\section{The metadata-fiber criterion}
\label{sec:metadata}

Return to the data of~\eqref{eq:refinement}.  The based total space is
\[
       T_M(a)
       =\sum_{b:A}R_M(a,b)
       =\sum_{b:A}\sum_{h:a=b}M(b,h),
\tag{4.1}\label{eq:metadata-total}
\]
with center
\[
                         c_M=(a,\refl_a,m_0).
\tag{4.2}\label{eq:metadata-center}
\]

\subsection{Collapsing the equality coordinates}

\begin{lemma}[Based equality total space]
\label{lem:equality-total}
The type \(P_a=\sum_{b:A}(a=b)\) is contractible with center
\((a,\refl_a)\).
\end{lemma}

\begin{proof}
For \((b,h):P_a\), equality elimination on \(h\) reduces the goal
\[
                    (a,\refl_a)=(b,h)
\]
to reflexivity at \((a,\refl_a)\).  This defines the contraction for every
\((b,h)\).
\end{proof}

\begin{theorem}[Total space--fiber equivalence]
\label{thm:total-fiber}
There is a specified equivalence
\[
             \Phi_M:T_M(a)\Eqv M(a,\refl_a)
\tag{4.3}\label{eq:total-fiber}
\]
whose forward map sends \((a,\refl_a,m)\) to \(m\), and whose inverse sends
\(m\) to \((a,\refl_a,m)\).
\end{theorem}

\begin{proof}
Define the inverse immediately:
\[
       G:M(a,\refl_a)\longrightarrow T_M(a),\qquad
       G(m)=(a,\refl_a,m).
\tag{4.4}\label{eq:fiber-inclusion}
\]
For the forward map, take \((b,h,m):T_M(a)\) and eliminate on \(h:a=b\).
In the reflexive case return \(m\):
\[
 F(b,h,m)=
 \mathsf J_{=}\bigl(
    b,h,m;\;m\mapsto m
 \bigr):M(a,\refl_a).
\tag{4.5}\label{eq:fiber-retraction}
\]
Equivalently, using the contraction
\(\gamma_{b,h}:(a,\refl_a)=(b,h)\) from
Lemma~\ref{lem:equality-total},
\[
 F(b,h,m)=
 \transport_{\widetilde M}(\gamma_{b,h}^{-1},m),
 \qquad \widetilde M(b,h)=M(b,h).
\tag{4.6}\label{eq:fiber-transport}
\]
At reflexivity, \(F(G(m))=m\).  For the other composite, eliminate on
\(h\).  The required equation becomes
\[
 G(F(a,\refl_a,m))=(a,\refl_a,m),
\]
which is reflexivity.  Thus \(F\) and \(G\) are inverse.
\end{proof}

The proof uses equality elimination; it does not choose representatives of
arbitrary paths, normalize metadata, or assume that all metadata values equal
\(m_0\).  It says that all metadata in the total space can be transported to
the reflexivity fiber, where it remains visible.

\begin{lemma}[Contractibility respects equivalence]
\label{lem:contr-equiv}
If \(e:X\Eqv Y\), then
\[
                         \isContr(X)\Longleftrightarrow\isContr(Y).
\tag{4.7}\label{eq:contr-equiv}
\]
\end{lemma}

\begin{proof}
If \((c,\kappa)\) contracts \(X\), use \(e(c)\) as center of \(Y\).  For
\(y:Y\), the composite
\[
 e(c)\xrightarrow{\,e(\kappa(e^{-1}y))\,}
 e(e^{-1}y)\xrightarrow{\,\varepsilon_y\,}y
\]
is the contraction path.  Apply the same construction to \(e^{-1}\) for the
converse.
\end{proof}

\begin{theorem}[Metadata-fiber criterion, full form]
\label{thm:metadata-fiber}
For \(R_M\) and \(r^M_a\) as in~\eqref{eq:refinement},
\[
 \Jbeta_{R_M}(a,r^M_a)
 \quad\Longleftrightarrow\quad
 \isContr\bigl(T_M(a)\bigr)
 \quad\Longleftrightarrow\quad
 \isContr\bigl(M(a,\refl_a)\bigr).
\tag{4.8}\label{eq:metadata-full}
\]
\end{theorem}

\begin{proof}
The first equivalence is Theorem~\ref{thm:j-total}, applied to the relation row
\(b\mapsto R_M(a,b)\) and center \(c_M\).  The second is
Lemma~\ref{lem:contr-equiv}, applied to the explicit equivalence
\(\Phi_M\) of Theorem~\ref{thm:total-fiber}.  Composing the equivalences proves
the claim.
\end{proof}

Theorem~\ref{thm:headline} is the first-to-last equivalence in
\eqref{eq:metadata-full}.  The chosen base metadata \(m_0\) does not by itself
need to equal every metadata value by definition; the criterion says exactly
that such equalities must be available as a contraction.

\subsection{Immediate classifications}

\begin{corollary}[Contractible decorations]
\label{cor:contractible-decoration}
If \(M(a,\refl_a)\) is contractible, then \(R_M\) admits unrestricted based
elimination with propositional beta.  In particular this holds when the
reflexivity fiber is:
\begin{enumerate}[label=(\roman*)]
\item the unit type;
\item an inhabited subsingleton;
\item a proof-irrelevant proposition with a witness; or
\item a type equipped with a quotient identifying every element with \(m_0\).
\end{enumerate}
\end{corollary}

\begin{corollary}[Visible alternatives]
\label{cor:visible-alternatives}
If \(m_1:M(a,\refl_a)\) and \(m_0\ne m_1\), unrestricted based elimination
with propositional beta does not exist.
\end{corollary}

\begin{proof}
A contraction would identify \(m_0\) and \(m_1\) by
\eqref{eq:all-equal}; apply Theorem~\ref{thm:metadata-fiber}.
\end{proof}

\begin{remark}[Only the reflexivity fiber]
\label{rem:only-fiber}
The criterion does not require a separate contractibility hypothesis for
every \(M(b,h)\).  Equality elimination transports the entire total space to
\(M(a,\refl_a)\).  Conversely, noncanonical metadata over reflexivity is
already fatal, even if every nonreflexive-looking presentation has a unique
decoration.
\end{remark}

\section{Universal setoid quotient repair}
\label{sec:setoid-repair}

The phrase ``quotient the metadata'' is too weak to be a theorem: many
relations can be quotiented, and a quotient can retain several reflexivity
classes.  We now classify exactly which ordinary setoid quotients repair
unrestricted elimination.  Quotient exactness is essential in the converse.

\subsection{The empty-sensitive quotient theorem}

\begin{definition}[Total setoid]
\label{def:setoid-total}
Let \(S\) be a setoid on a type \(X\).  Define
\[
       \SetTot(S)
       \quad:\!\!\Longleftrightarrow\quad
       \prod_{x,y:X} S(x,y).
\tag{5.1}\label{eq:setoid-total}
\]
Write \(\eta_S:X\to X/S\) for the canonical quotient map.
\end{definition}

Totality concerns the relation, not the number of raw terms.  A type may have
many inhabitants while a total setoid has one quotient class.  Conversely, on
the empty type the condition is vacuous.  This is why the following theorem
has an inhabitance conjunct.

\begin{lemma}[Quotient inhabitance]
\label{lem:quotient-inhabited}
For every setoid \(S\) on \(X\),
\[
                 \mathsf{Inh}(X/S)
                 \quad\Longleftrightarrow\quad
                 \mathsf{Inh}(X).
\tag{5.2}\label{eq:quotient-inhabited}
\]
\end{lemma}

\begin{proof}
An inhabitant \(x:X\) gives \(\eta_S(x):X/S\).  Conversely, quotient induction
on an inhabitant of \(X/S\) exposes a representative \(x:X\).  No choice of a
representative for \emph{every} quotient class is made.
\end{proof}

\begin{theorem}[Setoid quotient classification]
\label{thm:quotient-contractible}
For every setoid \(S\) on \(X\),
\[
 \isContr(X/S)
 \quad\Longleftrightarrow\quad
 \bigl(\mathsf{Inh}(X)\ \wedge\ \SetTot(S)\bigr).
\tag{5.3}\label{eq:quotient-contractible}
\]
\end{theorem}

\begin{proof}
Suppose first that \(X/S\) is contractible.  It is inhabited, so
Lemma~\ref{lem:quotient-inhabited} gives \(\mathsf{Inh}(X)\).  For arbitrary
\(x,y:X\), contractibility gives
\[
                         \eta_S(x)=\eta_S(y).
\tag{5.4}\label{eq:quotient-images-equal}
\]
Quotient exactness turns~\eqref{eq:quotient-images-equal} back into
\(S(x,y)\).  Thus \(S\) is total.

Conversely, choose \(x_0:X\) from the inhabitance hypothesis and use
\(\eta_S(x_0)\) as center.  Quotient induction reduces an arbitrary
\(\xi:X/S\) to \(\eta_S(x)\).  Totality gives \(S(x_0,x)\), and quotient
soundness gives
\[
                       \eta_S(x_0)=\eta_S(x).
\tag{5.5}\label{eq:quotient-contraction}
\]
These equalities define the contraction.
\end{proof}

\begin{corollary}[Chosen-center form]
\label{cor:quotient-total}
If \(x_0:X\), then
\[
                    \isContr(X/S)
                    \quad\Longleftrightarrow\quad
                    \SetTot(S).
\tag{5.6}\label{eq:quotient-total}
\]
\end{corollary}

\begin{warning}[The empty carrier]
\label{warn:empty-carrier}
If \(X=\varnothing\), then \(\SetTot(S)\) holds vacuously while \(X/S\) is
empty and hence not contractible.  Dropping inhabitance from
Theorem~\ref{thm:quotient-contractible} would therefore be false.  In the
metadata application below the chosen \(m_0\) supplies inhabitance; it must not
be silently inferred in unrelated quotient statements.
\end{warning}

\subsection{Classification of metadata repairs}

Let
\[
 S_{b,h}\ \text{ be a setoid on }\ M(b,h)
\tag{5.7}\label{eq:setoid-family}
\]
for every \(b:A\) and \(h:a=b\).  Define the quotient metadata family and its
refined equality by
\[
 \overline M_S(b,h)=M(b,h)/S_{b,h},
 \qquad
 \overline R_S(a,b)=\sum_{h:a=b}\overline M_S(b,h),
\tag{5.8}\label{eq:quotient-metadata}
\]
with base
\[
       \overline r^S_a=(\refl_a,\eta_{S_{a,\refl_a}}(m_0)).
\tag{5.9}\label{eq:quotient-metadata-base}
\]

\begin{theorem}[Universal setoid-repair criterion]
\label{thm:universal-repair}
The quotient-repaired refinement admits unrestricted based elimination with
propositional beta exactly when the setoid on the reflexivity metadata fiber is
total:
\[
 \Jbeta_{\overline R_S}(a,\overline r^S_a)
 \quad\Longleftrightarrow\quad
 \SetTot(S_{a,\refl_a}).
\tag{5.10}\label{eq:universal-repair}
\]
\end{theorem}

\begin{proof}
Apply the metadata-fiber criterion to \(\overline M_S\):
\[
 \Jbeta_{\overline R_S}(a,\overline r^S_a)
 \quad\Longleftrightarrow\quad
 \isContr\!\left(M(a,\refl_a)/S_{a,\refl_a}\right).
\tag{5.11}\label{eq:repair-first-step}
\]
The raw reflexivity fiber is inhabited by \(m_0\).  Corollary
\ref{cor:quotient-total} identifies the right-hand side with
\(\SetTot(S_{a,\refl_a})\).
\end{proof}

The theorem is both a construction and an impossibility result.  The forward
direction says that every successful ordinary setoid repair must relate
\emph{every} pair \(x,y:M(a,\refl_a)\).  By quotient soundness,
\[
             \eta_{S_{a,\refl_a}}(x)
             =
             \eta_{S_{a,\refl_a}}(y)
             \qquad\text{for all }x,y.
\tag{5.12}\label{eq:no-visible-reflexivity}
\]
Therefore:
\begin{quote}
\centering\bfseries
No set-level quotient supporting unrestricted \(J^\beta\) can retain a
visible reflexivity-metadata distinction.
\end{quote}
This does not prohibit retaining raw metadata behind an opaque interface or
restricting motives.  It says that the quotient \emph{as the identity-like
object} has only one reflexivity class.

\begin{proposition}[Family-wide transport]
\label{prop:family-total}
If \(S_{a,\refl_a}\) is total, then every \(S_{b,h}\) is total.  Moreover, the
chosen \(m_0\) transports to an inhabitant of every \(M(b,h)\), so every
quotient \(M(b,h)/S_{b,h}\) is contractible.
\end{proposition}

\begin{proof}
Fix \(b:A\) and \(h:a=b\).  Equality induction on \(h\) reduces both assertions
to the reflexive case.  Totality is then the hypothesis, and inhabitance is
provided by \(m_0\).  Apply Corollary~\ref{cor:quotient-total} to the resulting
fiber.
\end{proof}

This proposition explains the precise sense in which the reflexivity test is
family-wide.  It is not a pointwise cardinality assertion about arbitrary
indices.  Equality induction transports the setoid family itself.  Empty
fibers cannot appear in this equality-indexed family once \(m_0\) is given,
whereas Warning~\ref{warn:empty-carrier} still governs standalone quotient
claims.

\subsection{The indiscrete repair}

\begin{definition}[Indiscrete setoid]
\label{def:indiscrete}
The indiscrete setoid \(\nabla_X\) on \(X\) is
\[
                        \nabla_X(x,y)=\top
                        \qquad(x,y:X).
\tag{5.14}\label{eq:indiscrete}
\]
\(\top\) denotes the always-true proposition, matching the Prop-valued Lean
relation \texttt{fun \_ \_ => True}; it is not the data type \(\Unit\).
\end{definition}

The relation \(\nabla_X\) is total, hence its quotient is contractible whenever
\(X\) is inhabited.  Applied fiberwise, it is the canonical unconditional
repair supplied by the data \(m_0\).

Terminology matters here.
\begin{itemize}[leftmargin=1.4em]
\item \(\nabla_X\) is \emph{maximal as a relation}: every relation on \(X\) is
  included in it.
\item \(X/\nabla_X\) is \emph{minimal in retained information}: all
  representatives have the same image.
\item The corresponding quotient is \emph{terminal} for the factor direction
  defined below.  Calling it initial reverses the arrows and is false.
\end{itemize}

\subsection{Exact factorization direction}

Let \(S,T\) be setoids on the same carrier \(X\).  A
\emph{quotient-map-preserving factor} is a map
\[
 F:X/S\longrightarrow X/T
 \quad\text{such that}\quad
 F\circ\eta_S=\eta_T.
\tag{5.15}\label{eq:factor-map}
\]

\begin{theorem}[Relation inclusion versus quotient factorization]
\label{thm:factor-inclusion}
There exists a quotient-map-preserving factor \(X/S\to X/T\) exactly when
\[
                 S\subseteq T,
 \qquad
                 \prod_{x,y:X}S(x,y)\to T(x,y).
\tag{5.16}\label{eq:relation-inclusion}
\]
When it exists, the factor is unique as a function.
\end{theorem}

\begin{proof}
If \(S\subseteq T\), quotient recursion on representatives defines
\[
                         F(\eta_S(x))=\eta_T(x).
\tag{5.17}\label{eq:factor-on-reps}
\]
The inclusion is exactly the well-definedness obligation: \(S(x,y)\) implies
\(\eta_T(x)=\eta_T(y)\) by quotient soundness.

Conversely, assume \(F\) satisfies~\eqref{eq:factor-map} and let \(S(x,y)\).
Soundness gives \(\eta_S(x)=\eta_S(y)\); applying \(F\) and using the commuting
equations gives \(\eta_T(x)=\eta_T(y)\).  Exactness yields \(T(x,y)\).

For uniqueness, quotient induction reduces two candidate factors to a
representative \(\eta_S(x)\), where both values are forced to be
\(\eta_T(x)\).
\end{proof}

\begin{corollary}[Terminal indiscrete factorization]
\label{cor:indiscrete-terminal}
For every setoid \(S\) on \(X\), there is a unique quotient-map-preserving map
\[
                       X/S\longrightarrow X/\nabla_X.
\tag{5.18}\label{eq:terminal-factor}
\]
\end{corollary}

\begin{proof}
Every relation \(S\) is included in \(\nabla_X\); apply
Theorem~\ref{thm:factor-inclusion}.
\end{proof}

If \(S\) is itself total, then also \(\nabla_X\subseteq S\), so there is a
representative-preserving map in the reverse direction.  Quotient induction
shows that the two maps are inverse:
\[
                         X/S\Eqv X/\nabla_X.
\tag{5.19}\label{eq:total-indiscrete-equiv}
\]
Thus every successful setoid repair in
Theorem~\ref{thm:universal-repair} is, on the reflexivity fiber, canonically
equivalent to the indiscrete repair.  The theorem does not say that the raw
setoids are definitionally equal, nor that one can recover discarded metadata.

\begin{remark}[Why terminal, not initial]
\label{rem:terminal-not-initial}
We order quotient presentations by loss of information.  If \(S\subseteq T\),
then \(X/S\to X/T\): the target identifies at least as much as the source.
Since every \(S\) is included in \(\nabla_X\), arrows point
\emph{toward} the indiscrete quotient.  It is terminal in this
quotient-map-preserving category.  Relation-theoretic maximality and
information-theoretic minimality describe the same object from opposite
viewpoints; neither justifies an initiality claim.
\end{remark}

\begin{remark}[Scope of universality]
\label{rem:quotient-scope}
The necessity proof uses exactness:
\(\eta_S(x)=\eta_S(y)\Rightarrow S(x,y)\).  The result therefore classifies
ordinary exact setoid quotients in the present extensional metatheory.  It is
not asserted for arbitrary truncations, higher quotients, opaque
implementations without exactness, or homotopical localization.
\end{remark}

\section{Projection kernels and the
  \texorpdfstring{\(\mathsf{PathRwQuot}\)}{PathRwQuot} criterion}
\label{sec:projection-kernel}

The metadata presentation \(\sum_hM(b,h)\) is convenient but not fundamental.
What matters is a relation carrying a reflexive projection to equality.

\subsection{Reflexive equality projections}

\begin{definition}[Reflexive equality projection]
\label{def:reflexive-projection}
Fix \(a:A\).  A reflexive equality projection consists of
\[
 R:A\to\U,\qquad r_0:R(a),\qquad
 q_b:R(b)\to(a=b),
\tag{6.1}\label{eq:reflexive-projection}
\]
together with \(q_a(r_0)=\refl_a\).
\end{definition}

\begin{definition}[Reflexivity kernel]
\label{def:reflexivity-kernel}
The reflexivity kernel of \(q\) is
\[
       \Ker(q)
       =
       \sum_{r:R(a)}\bigl(q_a(r)=\refl_a\bigr),
\tag{6.2}\label{eq:reflexivity-kernel}
\]
with center \((r_0,q_a(r_0)=\refl_a)\).
\end{definition}

\begin{theorem}[Projection/kernel equivalence]
\label{thm:projection-kernel-equiv}
There is a specified equivalence
\[
                       \sum_{b:A}R(b)\Eqv\Ker(q).
\tag{6.3}\label{eq:projection-kernel-equiv}
\]
\end{theorem}

\begin{proof}
Given \((b,r):\sum_bR(b)\), eliminate on \(q_b(r):a=b\).  In the reflexive case
return \(r:R(a)\), paired with proof irrelevance
\(q_a(r)=\refl_a\).  Conversely, forget the kernel equation and send
\((r,\rho)\) to \((a,r)\).  Equality induction on \(q_b(r)\) proves the first
inverse law; subtype extensionality and proof irrelevance prove the second.
\end{proof}

\begin{theorem}[Projection/kernel classification]
\label{thm:projection-kernel}
The relation row \(R\), based at \(r_0\), admits unrestricted based elimination
with propositional beta exactly when its reflexivity kernel is contractible:
\[
                  \Jbeta_R(a,r_0)
                  \quad\Longleftrightarrow\quad
                  \isContr(\Ker(q)).
\tag{6.4}\label{eq:projection-kernel}
\]
\end{theorem}

\begin{proof}
The total-space classification gives
\(\Jbeta_R(a,r_0)\leftrightarrow\isContr(\sum_bR(b))\).  Transport
contractibility across Theorem~\ref{thm:projection-kernel-equiv}.
\end{proof}

The metadata theorem is recovered by taking
\[
 R(b)=\sum_{h:a=b}M(b,h),
 \qquad
 q_b(h,m)=h.
\tag{6.5}\label{eq:metadata-as-projection}
\]
The kernel is then equivalent to \(M(a,\refl_a)\).  The projection theorem is
strictly more reusable: \(R\) need not be presented as a sigma type, and \(q\)
may forget an implementation-specific representation.

\subsection{Application to rewrite-quotiented paths}

Let
\[
 R_{\mathsf{rw}}(b)=\PRW_A(a,b)
 =\PathTy_A(a,b)/\mathsf{RwEq},
\tag{6.6}\label{eq:pathrw-row}
\]
based at the class \([\mathbf r_a]\).  The quotient operation
\[
 \mathsf{toEq}_b:\PRW_A(a,b)\longrightarrow(a=b)
\tag{6.7}\label{eq:pathrw-toeq}
\]
forgets rewrite traces and is well-defined because every primitive rewrite
preserves the underlying equality proof.

The reflexivity kernel is
\[
 \Ker(\mathsf{toEq})
 =
 \sum_{x:\PRW_A(a,a)}
       \bigl(\mathsf{toEq}(x)=\refl_a\bigr).
\tag{6.8}\label{eq:pathrw-kernel}
\]
Ambient equality proofs live in \(\mathsf{Prop}\), so the second component is
automatic by proof irrelevance.  Projection to the first component and
reinsertion of that proof give
\[
              \Ker(\mathsf{toEq})\Eqv\PRW_A(a,a).
\tag{6.9}\label{eq:pathrw-kernel-loop}
\]

\begin{definition}[Local quotient-level axiom K]
\label{def:local-k}
Define
\[
 \LocalK(A,a)
 \quad:\!\!\Longleftrightarrow\quad
 \prod_{x:\PRW_A(a,a)}x=[\mathbf r_a].
\tag{6.10}\label{eq:local-k}
\]
\end{definition}

This is a statement about classes of computational paths.  It is not the
assertion that two raw `Path` records with different step lists are equal, and
it is not obtained merely by observing that their \(\mathsf{toEq}\) values are
proof-irrelevant.

\begin{theorem}[\(\mathsf{PathRwQuot}\)/K classification]
\label{thm:pathrw-k}
For every \(A\) and \(a:A\), the following are equivalent:
\[
\begin{split}
 &\text{\(\PRW_A(a,-)\), based at \([\mathbf r_a]\), admits
 unrestricted \(J^\beta\);}\\
 &\isContr\bigl(\PRW_A(a,a)\bigr);\\
 &\LocalK(A,a).
\end{split}
\tag{6.11}\label{eq:pathrw-k}
\]
\end{theorem}

\begin{proof}
The first equivalence is Theorem~\ref{thm:projection-kernel} followed by
\eqref{eq:pathrw-kernel-loop}.  For the second, a contraction identifies every
loop class with the reflexive class.  Conversely, the equations in
\(\LocalK(A,a)\), reversed in orientation, contract the loop quotient at
\([\mathbf r_a]\).
\end{proof}

\begin{corollary}[Global form]
\label{cor:pathrw-global-k}
The family \(\PRW_A(a,-)\) admits such an eliminator for every \(a:A\) exactly
when \(\prod_{a:A}\LocalK(A,a)\).
\end{corollary}

\begin{remark}[No unconditional slogan]
\label{rem:rweq-no-slogan}
Theorem~\ref{thm:pathrw-k} does not say that quotienting by
\(\mathsf{RwEq}\) always repairs unrestricted elimination.  That would require
local K for every carrier.  It also does not say that the quotient always
fails: when the loop quotient is trivial, the repair succeeds.  The theorem is
the exact criterion, and concrete spaces must be checked against it.
\end{remark}

\begin{remark}[Normalization is only one direction]
\label{rem:rweq-normalization}
In the artifact, \(\mathsf{RwEq}\) is the equivalence closure generated by
explicit primitive rewrites.  The current function
\(\mathsf{normalize}(p)=\mathsf{stepChain}(p.\mathsf{toEq})\) satisfies
\[
       p\mathrel{\mathsf{RwEq}}q
       \quad\Longrightarrow\quad
       \mathsf{normalize}(p)=\mathsf{normalize}(q).
\tag{6.12}\label{eq:rweq-normalization}
\]
The converse is not part of the bare definition and cannot replace an
\(\mathsf{RwEq}\) witness.  A normal-form characterization is available only
relative to a separately supplied complete step system.  This distinction is
crucial when claiming that a quotient collapses a particular raw trace.
\end{remark}

\section{The largest explicitly supported class of motives}
\label{sec:factorization}

Failure of unrestricted elimination does not mean that no useful elimination
principle remains.  The projection~\eqref{eq:projection} supports precisely the
class we can state and prove without inspecting metadata.

\begin{definition}[Strict equality-projection motive]
\label{def:strict-factor}
Given
\[
                  D:\prod_{b:A}(a=b)\longrightarrow\U,
\]
its pullback along \(\proj\) is the motive
\[
       C_D(b,h,m)=D(b,h)
       \qquad ((h,m):R_M(a,b)).
\tag{7.1}\label{eq:strict-factor}
\]
\end{definition}

\begin{theorem}[Motive factorization]
\label{thm:motive-factorization}
Every strict equality-projection motive \(C_D\) admits elimination
\[
  J_{\proj}:
  D(a,\refl_a)\longrightarrow
  \prod_{b:A}\prod_{p:R_M(a,b)}D(b,\proj p)
\tag{7.2}\label{eq:factored-j}
\]
with propositional beta at \(r^M_a\):
\[
                  J_{\proj}(d,a,r^M_a)=d.
\tag{7.3}\label{eq:factored-beta}
\]
More generally, the conclusion holds for any motive \(C\) equipped, for every
\(x=(b,h,m):T_M(a)\), with a specified fiberwise equivalence
\[
                         C(x)\Eqv D(b,h).
\tag{7.4}\label{eq:fiberwise-factor}
\]
\end{theorem}

\begin{proof}
For the strict case, ignore \(m\) and eliminate on \(h:a=b\):
\[
       J_{\proj}(d,b,(h,m))
       =\mathsf J_{=}(D,d,b,h):D(b,h).
\]
At \(h=\refl_a\), equality beta gives~\eqref{eq:factored-beta}.

For the fiberwise case, let \(e_x:C(x)\Eqv D(b,h)\) be the chosen equivalence.
Map the base datum \(c:C(c_M)\) forward along \(e_{c_M}\), apply the strict
construction in \(D\), and map the result back along \(e_x^{-1}\).  At the
center the equality eliminator has propositional beta, and
\(e_{c_M}^{-1}(e_{c_M}(c))=c\), proving the required beta equation.
\end{proof}

The theorem identifies the largest class \emph{explicitly supported here}:
motives supplied with a factorization through equality, strictly or up to
fiberwise equivalence.  We do not claim that this is a semantic maximality
theorem about isolated motives.  A particular metadata-sensitive motive might
have a section for accidental reasons.  What fails is one uniform operator
covering every motive from one base datum.

\begin{corollary}[Metadata irrelevance for factored elimination]
\label{cor:factored-irrelevance}
For \(p=(h,m)\) and \(q=(k,n)\) in \(R_M(a,b)\), if \(h=k\), then factored
elimination gives propositionally equal results after transport along that
equality.  In the proof-irrelevant ambient setting this applies to every
pair \(p,q:R_M(a,b)\).
\end{corollary}

\subsection{Two remedies}

\paragraph{Restrict motives.}
Keep the full record \(R_M\), but expose only motives of the form
\eqref{eq:strict-factor} or~\eqref{eq:fiberwise-factor}.  This retains
histories, costs, and traces for inspection by nondependent operations while
preventing a purported identity eliminator from distinguishing them.

\paragraph{Hide or quotient metadata.}
Replace \(M\) by a family \(\overline M\) whose reflexivity fiber is
contractible.  Examples include an opaque interface with a single observable
value, a propositional truncation when inhabited, or an explicit quotient
collapsing all reflexive histories.  Theorem~\ref{thm:metadata-fiber} then
provides unrestricted elimination.  This is not a proof that the original
metadata was irrelevant; it is a new identity-like object in which the
distinction has been removed.

\begin{warning}[Quotienting must match the intended observation]
Merely quotienting by administrative rewrites is insufficient if two
reflexive metadata classes remain.  The criterion asks for contractibility of
the resulting reflexivity fiber, not merely confluence, normalization, or
decidable equality.
\end{warning}

\section{Computational traces: the principal case study}
\label{sec:traces}

\subsection{The trace-carrying record}

For a carrier \(A\), let an elementary step be
\[
       \Step(A)=\sum_{x,y:A}(x=y).
\tag{8.1}\label{eq:step}
\]
A step stores source, target, and an ambient equality witness.  Define
\[
       \PathTy_A(a,b)
       =\List(\Step(A))\times(a=b),
\tag{8.2}\label{eq:path}
\]
written as a record \((L,h)\).  Its reflexive element is
\[
                         \mathbf r_a=(\Empty,\refl_a).
\tag{8.3}\label{eq:empty-refl}
\]
The list is observable data.  Concatenation and reversal may be used to define
composition and symmetry, but those operations are not needed for the
classification.

Set
\[
 M_{\mathrm{tr}}(b,h)=\List(\Step(A)).
\tag{8.4}\label{eq:trace-metadata}
\]
Reordering fields gives mutually inverse maps
\[
\begin{aligned}
 S:\PathTy_A(a,b)&\longrightarrow
      \sum_{h:a=b}M_{\mathrm{tr}}(b,h),
 & S(L,h)&=(h,L),\\
 T:\sum_{h:a=b}M_{\mathrm{tr}}(b,h)&\longrightarrow
      \PathTy_A(a,b),
 & T(h,L)&=(L,h).
\end{aligned}
\tag{8.5}\label{eq:path-refinement}
\]
Both inverse equations are reflexivity of records.  Thus this is not only an
analogy: \(\PathTy\) is the corresponding equality refinement.

\begin{definition}[Two reflexive traces]
\label{def:two-traces}
Besides \(\mathbf r_a\), define
\[
 \mathbf s_a=
  \bigl([\, (a,a,\refl_a)\,],\refl_a\bigr)
  :\PathTy_A(a,a).
\tag{8.6}\label{eq:singleton-refl}
\]
\end{definition}

The list no-confusion principle gives
\[
       \Empty\ne[\, (a,a,\refl_a)\,],
\qquad\text{hence}\qquad
       \mathbf r_a\ne\mathbf s_a.
\tag{8.7}\label{eq:trace-distinct}
\]

\begin{theorem}[No unrestricted trace-sensitive \(J\)]
\label{thm:no-trace-j}
For every \(A\) and \(a:A\), the family \(\PathTy_A(a,-)\), based at the empty
reflexive trace, has no unrestricted based eliminator satisfying
propositional beta.
\end{theorem}

\begin{proof}
By~\eqref{eq:path-refinement}, \(\PathTy_A(a,b)\) is the equality refinement
with metadata~\eqref{eq:trace-metadata}.  Its reflexivity metadata fiber is
\(\List(\Step(A))\).  That fiber contains the distinct values in
\eqref{eq:trace-distinct}, so it is not contractible.  Apply
Theorem~\ref{thm:metadata-fiber}.  This derives the result from the generic
criterion; no bespoke motive is needed.
\end{proof}

One may recover the familiar direct counter-motive by taking
\[
 C(b,L,h)=
 \begin{cases}
 \Unit,&L=\Empty,\\
 \varnothing,&L\ne\Empty.
 \end{cases}
\]
It has a base value but none at \(\mathbf s_a\).  This is a useful diagnostic,
not the classification proof.

\subsection{Composability is not the issue}

The raw list in~\eqref{eq:path} need not enforce adjacency.  It is tempting to
attribute Theorem~\ref{thm:no-trace-j} to that permissiveness.  The following
intrinsic version rules out the diagnosis.

\begin{definition}[Composable visible trace]
\label{def:composable}
A list \([s_1,\ldots,s_n]\) is composable from \(a\) to \(b\) when the source
of \(s_1\) is \(a\), the target of \(s_i\) is the source of \(s_{i+1}\), and
the target of \(s_n\) is \(b\).  The empty list is composable from \(a\) to
\(a\).  Write \(\mathsf{CTr}_A(a,b)\) for the subtype of lists satisfying this
condition.
\end{definition}

Both
\[
      \epsilon_a=(\Empty,\text{empty-chain proof})
      \quad\text{and}\quad
      \sigma_a=([\,a\xrightarrow{\refl_a}a\,],
                   \text{singleton-chain proof})
\tag{8.8}\label{eq:composable-two}
\]
belong to \(\mathsf{CTr}_A(a,a)\).  Equality of subtype values would imply
equality of their underlying lists, so \(\epsilon_a\ne\sigma_a\).

\begin{corollary}[Composable-trace obstruction]
\label{cor:composable}
The equality refinement
\[
   \sum_{h:a=b}\mathsf{CTr}_A(a,b),
\tag{8.9}\label{eq:composable-refinement}
\]
based at \((\refl_a,\epsilon_a)\), has no unrestricted based eliminator with
propositional beta.
\end{corollary}

\begin{proof}
The reflexivity metadata fiber
\(\mathsf{CTr}_A(a,a)\) contains the distinct values
\(\epsilon_a\) and \(\sigma_a\).  It is noncontractible, so
Theorem~\ref{thm:metadata-fiber} applies.
\end{proof}

The singleton step starts at \(a\), ends at \(a\), and has reflexive composite.
It survives any condition requiring adjacency, correct outer endpoints, and a
composite equality matching \(\refl_a\).  The obstruction is therefore:
\[
 \boxed{\text{visible noncanonical reflexive metadata, not
 non-composability.}}
\tag{8.10}\label{eq:boxed-obstruction}
\]

\subsection{What remains available}

The endpoint-equality projection
\[
       (L,h)\longmapsto h
\tag{8.11}\label{eq:trace-projection}
\]
is exactly~\eqref{eq:projection}.  If
\(D:\prod_{b:A}(a=b)\to\U\), Theorem
\ref{thm:motive-factorization} gives
\[
 D(a,\refl_a)\longrightarrow
 \prod_{b:A}\prod_{P:\PathTy_A(a,b)}D(b,\proj P)
\tag{8.12}\label{eq:trace-factored-j}
\]
with propositional beta at \(\mathbf r_a\).  The result is independent of the
list.  Thus trace algebra, including concatenation and rewrite equivalence,
can coexist with equality-component elimination; what cannot coexist is an
unrestricted eliminator whose motive is allowed to separate
\(\mathbf r_a\) from \(\mathbf s_a\).

\subsection{Computational paths and rewrite evidence}

In the accompanying library, \(\PathTy\) is the structure
\(\texttt{Path}\), with fields
\(\texttt{steps : List (Step A)}\) and
\(\texttt{proof : a = b}\).  Composition concatenates lists, and
\(\texttt{RwEq}\) records the LND\(_{\mathrm{EQ}}\)-TRS groupoid
coherences.  For example, if \(p\) is a two-step arithmetic path, the artifact
constructs
\[
       p\cdot p^{-1}\;\mathrel{\mathsf{RwEq}}\;\mathbf r
\tag{8.13}\label{eq:rweq-evidence}
\]
as an actual rewrite certificate.  Such a certificate organizes traces; it
does not equate every list as record data.  In particular it does not make the
raw reflexivity metadata fiber contractible unless one forms an appropriate
quotient and proves that quotient contractible.

\subsection{Empty and singleton after quotienting}
\label{subsec:raw-versus-rweq}

The two raw witnesses used in Theorem~\ref{thm:no-trace-j} behave differently
after quotienting.  In artifact notation, set
\[
 \epsilon_a=\texttt{Path.refl}\ a,
 \qquad
 \sigma_a=\texttt{Path.stepChain}\ \refl_a.
\tag{8.14}\label{eq:raw-empty-singleton}
\]
Their step lists have lengths zero and one, so
\(\epsilon_a\ne\sigma_a\) as `Path` records.  Nevertheless the primitive
transport beta rewrite, instantiated with a constant family, gives
\[
              \sigma_a
              \mathrel{\mathsf{RwEq}}
              \epsilon_a.
\tag{8.15}\label{eq:raw-empty-singleton-rweq}
\]
Symmetry gives the reverse orientation.  Quotient soundness therefore proves
\[
                    [\epsilon_a]=[\sigma_a]
                    \quad\text{in }\PRW_A(a,a).
\tag{8.16}\label{eq:raw-empty-singleton-quot}
\]

This example separates two observability questions.
\begin{enumerate}[label=(\roman*)]
\item A motive on raw `Path` may inspect the list and distinguish
  \(\epsilon_a\) from \(\sigma_a\).  The raw reflexivity fiber is therefore
  noncontractible.
\item A motive on \(\PRW\) sees only the quotient class.  This particular
  distinction has been removed by an actual rewrite certificate.
\end{enumerate}
Equality of the outputs of the current normalization function would not, by
itself, establish~\eqref{eq:raw-empty-singleton-rweq}; the formal development
uses the named primitive rewrite.  Nor does this one calculation prove that
the entire loop quotient is contractible.  The latter is exactly the local-K
question from Theorem~\ref{thm:pathrw-k}.

\section{Genuine and synthetic circle and torus quotients}
\label{sec:genuine-synthetic}

The repository contains two constructions that earlier names and comments
could invite a reader to conflate:
\begin{enumerate}[label=(\alph*)]
\item the genuine quotient
  \(\PRW_A(a,a)=\PathTy_A(a,a)/\mathsf{RwEq}\); and
\item a separate syntax of winding expressions, quotiented by equality of an
  integer-valued encoding.
\end{enumerate}
They have different carriers, different relations, and, for the present circle
and torus definitions, incompatible contractibility properties.

\subsection{A pointed-subsingleton calculation}

\begin{proposition}[Loop quotient of a pointed subsingleton]
\label{prop:subsingleton-loop-quotient}
Let \(a:A\) and suppose
\[
                         \prod_{x:A}(x=a).
\tag{9.1}\label{eq:all-points-base}
\]
Then every raw loop \(p:\PathTy_A(a,a)\) is
\(\mathsf{RwEq}\)-equivalent to \(\mathbf r_a\), and
\[
                         \isContr(\PRW_A(a,a)).
\tag{9.2}\label{eq:subsingleton-loop-contractible}
\]
\end{proposition}

\begin{proof}
Write the stored trace of \(p\) as a list of elementary steps.  By
\eqref{eq:all-points-base}, each stored source and target is \(a\).  Proof
irrelevance identifies the step's equality field with \(\refl_a\).  Induct on
the list.  The empty list is \(\mathbf r_a\).  In the successor case, split off
the leading singleton `stepChain`, rewrite that singleton to reflexivity by the
primitive transport-reflexivity beta rule, remove the resulting left unit, and
apply the induction hypothesis.  This constructs an explicit composite
\(\mathsf{RwEq}\) witness.

For a quotient class \([p]\), quotient soundness gives
\([p]=[\mathbf r_a]\).  Reversing the equation contracts the quotient at
\([\mathbf r_a]\).
\end{proof}

The proof is stronger than an equality-of-normalizations argument: it builds
the rewrite certificate needed by the defining quotient relation.

\subsection{The current genuine circle and torus}

The artifact currently defines
\[
 \begin{aligned}
  \mathsf{CircleCompPath}&::=\mathsf{base},\\
  \mathsf{Circle}&:=\mathsf{CircleCompPath},\\
  \mathsf{Torus}&:=\mathsf{Circle}\times\mathsf{Circle}.
 \end{aligned}
\tag{9.3}\label{eq:current-circle-torus}
\]
There is no higher-inductive loop constructor in the carrier
\(\mathsf{CircleCompPath}\).  Its separately declared loop-expression syntax
does not alter the equality type of the carrier.

Every element of \(\mathsf{Circle}\) is \(\mathsf{base}\) by ordinary
constructor elimination, and every element of \(\mathsf{Torus}\) is the pair
of base points.  Proposition~\ref{prop:subsingleton-loop-quotient} therefore
gives:

\begin{theorem}[Genuine circle and torus loop quotients]
\label{thm:genuine-circle-torus}
Under the current definitions,
\[
 \begin{aligned}
  \isContr\bigl(
    \PRW_{\mathsf{Circle}}(\mathsf{base},\mathsf{base})
  \bigr),\\
  \isContr\bigl(
    \PRW_{\mathsf{Torus}}(\mathsf{base},\mathsf{base})
  \bigr).
 \end{aligned}
\tag{9.4}\label{eq:genuine-contractible}
\]
Both genuine loop quotients satisfy local quotient-level axiom K and hence the
unrestricted elimination criterion of Theorem~\ref{thm:pathrw-k}.
\end{theorem}

The displayed notation abbreviates the loop fibers
\(\PRW\,\mathsf{Circle}\,\mathsf{base}\,\mathsf{base}\) and similarly for the
torus.  It does not assert that the raw `Path` type is a subsingleton: raw step
lists remain observable before quotienting.

\subsection{The synthetic winding presentations}

Independently, the file `CircleCompPath.lean` defines an indexed syntax
\[
\begin{aligned}
 e ::= {}&\mathsf{loop}
 \mid \mathsf{refl}
 \mid \mathsf{symm}(e)
 \mid \mathsf{trans}(e,e')
\end{aligned}
\tag{9.5}\label{eq:circle-expression}
\]
and an encoding \(w(e):\mathbb Z\) with the expected clauses
\[
\begin{aligned}
 w(\mathsf{loop})&=1,&
 w(\mathsf{refl})&=0,\\
 w(\mathsf{symm}(e))&=-w(e),&
 w(\mathsf{trans}(e,e'))&=w(e)+w(e').
\end{aligned}
\tag{9.6}\label{eq:winding-clauses}
\]
It then forms the setoid quotient
\[
 \mathsf{CircleWind}
 =
 \mathsf{Expr}/(e\sim e'\Longleftrightarrow w(e)=w(e')).
\tag{9.7}\label{eq:circle-wind}
\]
Encode/decode proves
\[
                  \mathsf{CircleWind}\Eqv\mathbb Z.
\tag{9.8}\label{eq:circle-wind-int}
\]
The legacy public alias `circlePiOne` denotes this synthetic quotient.

The synthetic torus presentation is the product
\[
             \mathsf{TorusWind}
             =
             \mathsf{CircleWind}\times\mathsf{CircleWind}
             \Eqv\mathbb Z\times\mathbb Z.
\tag{9.9}\label{eq:torus-wind}
\]
Its legacy alias is `torusPiOne`.  These encode/decode theorems are valid and
useful algebraic presentations.  The correction is solely that they are not
the genuine \(\PRW\) loop fibers of the carriers in
\eqref{eq:current-circle-torus}.

\begin{proposition}[Synthetic noncontractibility]
\label{prop:synthetic-noncontractible}
Both \(\mathsf{CircleWind}\) and \(\mathsf{TorusWind}\) are noncontractible.
\end{proposition}

\begin{proof}
Contractibility is invariant under equivalence.  The integers contain the
distinct points \(0\) and \(1\); integer pairs contain \((0,0)\) and \((1,0)\).
Thus neither target in~\eqref{eq:circle-wind-int} or~\eqref{eq:torus-wind} is
contractible, and neither source can be contractible.
\end{proof}

\subsection{No equivalence bridge}

\begin{theorem}[Circle and torus no-bridge results]
\label{thm:no-bridges}
There is no `SimpleEquiv` between the current genuine circle loop quotient and
\(\mathsf{CircleWind}\), and no `SimpleEquiv` between the current genuine torus
loop quotient and \(\mathsf{TorusWind}\):
\[
\begin{aligned}
 \neg\bigl(
   \PRW_{\mathsf{Circle}}(\mathsf{base},\mathsf{base})
   \Eqv\mathsf{CircleWind}
 \bigr),\\
 \neg\bigl(
   \PRW_{\mathsf{Torus}}(\mathsf{base},\mathsf{base})
   \Eqv\mathsf{TorusWind}
 \bigr).
\end{aligned}
\tag{9.10}\label{eq:no-bridges}
\]
\end{theorem}

\begin{proof}
An equivalence transports contractibility.  The source in each line is
contractible by Theorem~\ref{thm:genuine-circle-torus}; the target is
noncontractible by Proposition~\ref{prop:synthetic-noncontractible}.  Either
equivalence yields a contradiction.
\end{proof}

This settles the formerly suggested bridge under the current signatures: it is
not merely unformalized, but impossible.  The exact formal statement excludes
invertible `SimpleEquiv` bridges.  It does not claim that no function whatsoever
can be written between the types, nor that a future, genuinely higher-inductive
circle model could not have a nontrivial loop quotient.  Such a future model
would be a different carrier and would require a new theorem.

\begin{table}[t]
\centering
\small
\renewcommand{\arraystretch}{1.18}
\begin{tabular}{>{\raggedright\arraybackslash}p{0.29\textwidth}
                >{\raggedright\arraybackslash}p{0.28\textwidth}
                >{\raggedright\arraybackslash}p{0.31\textwidth}}
\toprule
Object & Defining quotient & Current result\\
\midrule
Genuine circle loop fiber
 & raw `Path` loops modulo `RwEq`
 & Contractible; local K holds.\\
Synthetic circle winding quotient
 & expressions modulo equal integer encoding
 & Equivalent to \(\mathbb Z\); noncontractible.\\
Genuine torus loop fiber
 & raw product-carrier loops modulo `RwEq`
 & Contractible; local K holds.\\
Synthetic torus winding quotient
 & product of two synthetic circle quotients
 & Equivalent to \(\mathbb Z^2\); noncontractible.\\
\bottomrule
\end{tabular}
\caption{The current genuine and synthetic objects have incompatible
contractibility and cannot be equivalent.}
\label{tab:genuine-synthetic}
\end{table}

\section{Beyond lists: a classification guide}
\label{sec:examples}

The criterion applies to any observable decoration.  The following examples
are not encodings of one another; they illustrate which property matters.

\begin{table}[t]
\centering
\small
\renewcommand{\arraystretch}{1.22}
\begin{tabular}{>{\raggedright\arraybackslash}p{0.20\textwidth}
                >{\raggedright\arraybackslash}p{0.36\textwidth}
                >{\raggedright\arraybackslash}p{0.34\textwidth}}
\toprule
Metadata & Reflexivity fiber & Unrestricted \(J^\beta\)\\
\midrule
Hidden token
 & A singleton or inhabited subsingleton
 & Yes: the fiber is contractible.\\
Derivation history
 & Empty history and a visible reflexive inference
 & No, unless histories are hidden or collapsed.\\
Cost
 & Distinct costs \(0\) and \(1\) for reflexive computations
 & No; uniqueness of a minimum is irrelevant.\\
Certificate
 & An inhabited proof-irrelevant proposition
 & Yes in the present metatheory.\\
Certificate as data
 & Two distinguishable certificate objects
 & No, even if both certify the same fact.\\
Normalization log
 & Empty log and a nonempty administrative loop
 & No; unique normal forms do not imply unique logs.\\
Proof label
 & Exactly one reflexivity label
 & Yes; two visible reflexivity labels make it fail.\\
\bottomrule
\end{tabular}
\caption{The reflexivity fiber, rather than the informal purpose of metadata,
determines unrestricted based elimination.}
\label{tab:examples}
\end{table}

\subsection{Derivation histories}

Let \(H(b,h)\) be a type of derivations whose conclusion is \(h:a=b\).
Suppose \(H(a,\refl_a)\) contains an empty derivation \(e\) and an explicit
reflexivity-rule derivation \(\rho\), with \(e\ne\rho\).
Then
\[
                  \sum_{h:a=b}H(b,h)
\]
fails unrestricted \(J^\beta\).  Confluence of derivations does not alter this
conclusion when histories remain visible as syntax.  A quotient by conversions
works only if the quotient reflexivity fiber is contractible.

\begin{example}[Administrative steps]
A derivation calculus may have a ``do nothing'' derivation and a derivation
that inserts then removes a coercion.  Even if both evaluate to the same
endpoint equality and are interconvertible, they remain an obstruction as raw
metadata.  Passing to conversion classes is a separate construction requiring
the contractibility check again.
\end{example}

\subsection{Costs and resource annotations}

Let \(M(b,h)=\mathbb N\), with base cost \(0\).  Since \(0\ne1\), the
reflexivity fiber is noncontractible and unrestricted elimination fails.
This remains true if composition adds costs, if cost comparison is decidable,
or if \(0\) is the unique optimal cost.  Optimization supplies an order and a
preferred element, not an equality between all costs.

A canonical-cost interface can recover elimination by exposing only the
proposition that the stored cost is optimal, provided that proposition is
inhabited and proof-irrelevant.  Alternatively, the cost can remain stored but
motives must factor through equality.

\subsection{Certificates}

The word ``certificate'' hides an important representation choice.  If
\(M(a,\refl_a)\) is an inhabited proposition, proof irrelevance makes it
contractible.  If certificates are records with serial numbers, proof trees,
or signatures, two records may certify the same equality while remaining
distinguishable.  The latter representation fails the criterion.

\begin{proposition}[Certificate dichotomy]
An inhabited proof-irrelevant certificate fiber supports unrestricted
elimination.  A certificate fiber with two provably distinct values does not.
\end{proposition}

\begin{proof}
Apply Corollaries~\ref{cor:contractible-decoration} and
\ref{cor:visible-alternatives}, respectively.
\end{proof}

\subsection{Normalization logs}

Suppose every term has a unique normal form, but normalization records the
sequence of reductions.  Unique endpoints do not imply unique logs.  An empty
reflexive log and a log containing a reversible administrative step are enough
to refute contractibility.  This separates three properties:
\[
\begin{array}{c}
\text{unique normal form},\\
\text{all logs connected by log rewrites},\\
\text{all observable reflexive log values equal}.
\end{array}
\]
Only the third is the metadata-fiber condition.  The second can imply the
third after quotienting, but not for raw logs.

\subsection{Proof labels}

Let labels depend on the endpoint and equality proof.  The criterion permits
arbitrarily elaborate indexing away from the chosen presentation; after the
total-space equivalence, only the transported reflexivity fiber remains.  A
single reflexivity label, or an inhabited subsingleton of labels, is harmless.
Two visible labels for reflexivity obstruct unrestricted elimination.

\begin{example}[Sparse metadata]
It can happen that a presentation assigns different-looking label sets to
different \(b,h\), yet equality elimination transports all of them into a
contractible reflexivity fiber.  Theorem~\ref{thm:total-fiber}, not a
pointwise cardinality argument, is the correct classification.
\end{example}

\section{Conceptual consequences and boundaries}
\label{sec:comparison}

\subsection{The standard identity argument versus the fiber criterion}

The standard argument has the shape
\[
  \text{equality elimination}
  \Longrightarrow
  \isContr\!\left(\sum_b(a=b)\right).
\tag{11.1}\label{eq:standard-chain}
\]
Our classification adds a genuinely separate coordinate calculation:
\[
\begin{aligned}
 \Jbeta_{R_M}
 &\Longleftrightarrow
   \isContr\!\left(\sum_b\sum_hM(b,h)\right)\\
 &\Longleftrightarrow
   \isContr\bigl(M(a,\refl_a)\bigr).
\end{aligned}
\tag{11.2}\label{eq:new-chain}
\]
The first line uses the universal property of a pointed contractible total
space.  The second uses Theorem~\ref{thm:total-fiber}.  Conflating these stages
would incorrectly suggest that every decoration of equality inherits path
induction.

The repair theorem adds a third, relation-theoretic stage:
\[
\begin{aligned}
 \Jbeta_{\overline R_S}
 &\Longleftrightarrow
 \isContr\bigl(M(a,\refl_a)/S_{a,\refl_a}\bigr)\\
 &\Longleftrightarrow
 \SetTot(S_{a,\refl_a}).
\end{aligned}
\tag{11.3}\label{eq:repair-chain}
\]
The first equivalence diagnoses the quotient metadata; the second uses
inhabitance and quotient exactness.  It is this last implication that turns
``collapse enough metadata'' into the exact statement ``relate every pair at
reflexivity.''

For a generic projection \(q:R\to{=}\), the corresponding summary is
\[
 \Jbeta_R
 \Longleftrightarrow
 \isContr(\Ker(q)).
\tag{11.4}\label{eq:kernel-chain}
\]
The \(\PRW\) case replaces \(\Ker(q)\) by the genuine loop quotient and hence
by local quotient-level axiom K.  These chains distinguish raw data, quotient
classes, and ambient equality rather than moving silently between them.

\subsection{What the theorem does not say}

\begin{enumerate}[label=(\roman*)]
\item It does not refute ordinary equality elimination.  The projection
  \(\proj\) still supports Theorem~\ref{thm:motive-factorization}.
\item It does not say metadata cannot be composed, inverted, normalized, or
  rewritten.  Those operations are orthogonal to raw fiber contractibility.
\item It does not establish a judgmental beta classification.  Every beta law
  in this paper is propositional.
\item It does not identify trace-carrying \(\PathTy\) with an intensional or
  homotopical identity type.
\item It does not model full MLTT or HoTT, and it proves nothing about
  univalence or higher inductive types.
\item It does not show that every individual metadata-sensitive motive lacks a
  section.  It rules out one unrestricted based eliminator for all motives.
\item The universal quotient theorem classifies exact ordinary setoid
  quotients.  It does not claim the same converse for higher quotients or
  quotient-like interfaces without exactness.
\item The \(\PRW\) theorem is conditional on local K; it does not label
  `RwEq` a universal success or a universal failure.
\item The no-bridge theorems concern the current one-constructor circle,
  product torus, and `SimpleEquiv`.  They do not preclude a future genuinely
  higher-inductive carrier with a different loop theory.
\end{enumerate}

\subsection{Relation to proof relevance}

Ambient equality in the present metatheory is proof-irrelevant.  The
proof-relevance in \(\PathTy\) comes from the \(\U\)-valued metadata field, not
from multiple inhabitants of the ambient equality proposition.  This
two-level distinction is essential:
\[
       (L,h)\ne(L',h)
       \quad\text{can hold because }L\ne L',
\]
even though any two witnesses \(h,k:a=b\) are equal.

In intensional homotopy type theory, identity itself may have higher
structure, and contractibility is formulated internally
\cite{HoTTBook2013}.  The total identity-space contraction remains a central
idea there.  Our theorem is deliberately about an external refinement of
proof-irrelevant equality.  It neither replaces the HoTT identity type nor
transfers the higher-groupoid interpretation to a trace record.

\subsection{Relation to computational paths}

Computational-path work treats equality reasons and their rewrites as
first-class syntax
\cite{deQueirozEtAl2016,RamosEtAl2017,RamosEtAl2018,deVerasEtAl2025,
RamosEtAl2026}.  The present result isolates one design boundary for an
implementation in which a \(\PathTy\) record contains both ambient equality
and visible steps.  Groupoid laws up to \(\mathsf{RwEq}\) organize the step
calculus; they do not automatically provide unrestricted dependent elimination
over raw records.  If one instead defines an identity type by a suitable
quotient or higher construction, the criterion must be applied to that new
fiber rather than to the raw list.  For ordinary setoid quotients,
Theorem~\ref{thm:universal-repair} performs that second check exactly.  For the
repository's `RwEq` quotient, Theorem~\ref{thm:pathrw-k} reduces the check to
local K.

\section{Related work}
\label{sec:related}

Martin-L\"of type theory makes identity elimination and its computation law
foundational \cite{MartinLof1984}.  The groupoid interpretation of
Hofmann and Streicher showed that intensional identity need not collapse to
proof-irrelevant equality \cite{HofmannStreicher1998}.  Homotopical and
weak-factorization semantics further explain identity elimination through
path-object structure \cite{AwodeyWarren2009,GambinoGarner2008}, while weak
higher-groupoid results expose the iterated coherence generated by identity
types \cite{Lumsdaine2010,vanDenBergGarner2011}.

The contraction of the based identity total space is standard in homotopy type
theory \cite{HoTTBook2013}.  Our contribution begins after that observation:
we calculate the total space of equality plus arbitrary observable metadata
and classify its elimination by one fiber.  This calculation is elementary,
but it is useful as a design theorem because it distinguishes a relation that
\emph{projects} to equality from one that itself supports unrestricted
identity elimination.

Two-level type theories carefully separate strict and fibrant equality
\cite{Capriotti2016,AltenkirchCapriottiKraus2016,AnnenkovEtAl2023}.
Our setting is not a two-level type theory: it has ambient equality and a data
record refining it, without a fibrancy judgment.  The methodological
similarity is only the insistence that the available eliminator be attached to
the correct equality notion.

The computational-path literature motivates the trace application and the
use of explicit rewrite histories
\cite{deQueirozEtAl2016,RamosEtAl2017,RamosEtAl2018,RamosEtAl2026}.
Earlier work studies a richer label-sensitive calculus.  The theorem here is
not a historical-priority claim about path induction; it is a classification
of a specific equality-plus-metadata interface.

\section{Formal verification and companion artifact}
\label{sec:formalization}

The mathematical development above is checked in Lean~4
\cite{deMouraUllrich2021,MetadataJSoftware2026}.  The public source
\path{ComputationalPaths/Path/TypeTheory/MetadataJ.lean} defines
\path{EqualityRefinement}, \path{MetadataTotal},
\path{UnrestrictedBasedEliminator}, and \path{IsContractible}.  Its main
theorems and constructions are:
\begin{itemize}[leftmargin=1.3em,itemsep=1pt,topsep=3pt]
\item \path{unrestricted_based_elimination_iff_contractible}, the equivalence
  \(\Jbeta\Eqv\isContr(\text{total})\);
\item \path{metadataTotalEquiv} and
  \path{contractible_iff_of_equiv}, giving
  \(T_M(a)\Eqv M(a,\refl_a)\) and invariance of contractibility;
\item \path{metadata_fiber_criterion} and
  \path{factorized_motive_eliminator}, formalizing
  Theorems~\ref{thm:metadata-fiber} and~\ref{thm:motive-factorization};
\item \path{pathRefinementEquiv} and
  \path{path_no_unrestricted_based_eliminator}, formalizing
  Theorem~\ref{thm:no-trace-j}; and
\item \path{composable_trace_no_unrestricted_based_eliminator}, formalizing
  Corollary~\ref{cor:composable}.
\end{itemize}
The same module contains a genuine length-two
\(\texttt{Path.trans}\) construction and the nontrivial
\(\texttt{RwEq}\) inverse-cancellation certificate
\texttt{metadataTwoStep\_cancel}.  There are no custom axioms or incomplete
proofs.

The repair module \path{MetadataRepair.lean}, in the same
\path{Path/TypeTheory} directory, contains the new repair and case-study
results:
\begingroup
\footnotesize
\begin{itemize}[leftmargin=1.3em,itemsep=1pt,topsep=3pt]
\item \path{quotient_contractible_iff_nonempty_and_setoidTotal} and
  \path{quotient_contractible_iff_setoidTotal}, using
  \path{Quotient.sound} and \path{Quotient.exact};
\item \path{metadata_quotient_repair_criterion},
  \path{metadata_setoid_total_all_fibers}, and
  \path{metadata_quotient_fibers_contractible};
\item \path{quotient_map_factor_iff_relationIncluded};\\
  \path{factorToIndiscrete_unique}; and\\
  \path{quotientEquivIndiscreteOfTotal};
\item \path{projection_kernel_criterion},
  \path{pathRwQuot_elimination_iff_loop_quotient_contractible}, and
  \path{pathRwQuot_elimination_iff_local_axiomK};
\item \path{raw_empty_ne_singleton_reflexive_path},
  \path{raw_empty_rweq_singleton_reflexive_path}, and
  \path{raw_empty_singleton_same_pathRwQuot_class}; and
\item \path{genuine_circle_}\allowbreak\path{piOne_contractible} and
  \path{genuine_torus_}\allowbreak\path{piOne_contractible};
  additionally,
  the two synthetic noncontractibility theorems, and
  \path{no_circle_genuine_}\allowbreak\path{synthetic_bridge} and
  \path{no_torus_genuine_}\allowbreak\path{synthetic_bridge}.
\end{itemize}
\endgroup
The pointed-subsingleton proof constructs a multi-step `RwEq` derivation by
induction on the stored trace; the raw singleton theorem uses the primitive
\path{Step.transport_refl_beta} rule.  The new file contains no custom axioms,
`sorry`, or `admit`.  Lean's axiom report for the principal quotient-repair and
no-bridge theorems contains only propositional extensionality and quotient
soundness; no classical choice is introduced by this development.

The release tag \texttt{metadata-quotient-repair-v1} names the stable
theorem-bearing commit cited in~\cite{MetadataJSoftware2026}; the tag is not a
placeholder DOI.  A future Zenodo archive may supplement that Git reference
when a DOI has actually been issued.

The earlier raw scope-indexed de Bruijn development is intentionally not
interleaved with the theory paper.  Its complete manuscript is preserved at
\[
                  \texttt{paper/adequacy/companion/main.tex}.
\]
It independently develops raw syntax, substitution, primitive-redex
preservation, syntactic quotients, and source identity programs.  The
companion builds separately and remains useful as a formal-interface study,
but none of those raw-calculus results is needed for
Theorem~\ref{thm:metadata-fiber}.  This separation keeps the present article a
mathematics/type-theory paper rather than a report on Lean or on a particular
de Bruijn implementation.

\section{Conclusion}
\label{sec:conclusion}

For equality decorated by observable metadata,
\[
                 R_M(a,b)=\sum_{h:a=b}M(b,h),
\]
unrestricted based elimination with propositional beta is controlled by one
object: \(M(a,\refl_a)\).  The proof separates a standard universal-property
step from a metadata-specific equivalence:
\[
 \Jbeta_{R_M}
 \Longleftrightarrow
 \isContr\!\left(\sum_bR_M(a,b)\right)
 \Longleftrightarrow
 \isContr\bigl(M(a,\refl_a)\bigr).
\]

For computational traces, empty and singleton reflexive lists are visible and
distinct.  The singleton is already well formed and composable, so the
failure is not caused by malformed chains.  It is caused by noncanonical
reflexive metadata.  Equality-projection motives remain eliminable, giving a
precise and useful boundary rather than a blanket impossibility result.

The diagnosis now has a universal ordinary-quotient repair.  Given setoids
\(S_{b,h}\), the repaired relation supports unrestricted \(J^\beta\) exactly
when \(S_{a,\refl_a}\) is total.  Every successful set-level repair therefore
identifies all reflexivity metadata.  The indiscrete relation is the canonical
instance: maximal as a relation, minimal in retained information, and terminal
for representative-preserving quotient factors.

The projection/kernel theorem removes the need for a sigma presentation.
For `PathRwQuot.toEq`, it yields the exact local-K criterion rather than a
blanket slogan about rewrite quotienting.  Empty and singleton reflexive paths
show why this precision matters: they are distinct raw records but explicitly
`RwEq`-equivalent and hence equal in the quotient.

Finally, the circle and torus audit separates constructions that current legacy
names had obscured.  The genuine loop quotients of the one-constructor circle
and its product torus are contractible.  The synthetic winding quotients are
equivalent to \(\mathbb Z\) and \(\mathbb Z^2\) and are noncontractible, so no
equivalence bridges the corresponding objects under the present definitions.
Histories, costs, certificates, normalization logs, and labels all follow the
same reflexivity-fiber test.  Any stronger claim---judgmental beta, full MLTT or
HoTT, univalence, a new higher-inductive circle, or a higher-quotient
universality theorem---requires additional theory and is outside the result
proved here.

\begingroup
\small
\bibliographystyle{amsplain}
\bibliography{refs}
\endgroup

\end{document}
