% ============================================================================
% \bibitem entries for the extended references (for manual bibliographies)
% Use these if your document does NOT use BibTeX.
% ============================================================================

\bibitem{Weibel13}
C.\,A.\ Weibel, \textit{The K-Book: An Introduction to Algebraic K-Theory},
Graduate Studies in Mathematics, vol.~145, American Mathematical Society, 2013.

\bibitem{Atiyah67}
M.\,F.\ Atiyah, \textit{K-Theory}, W.\,A.\ Benjamin, New York, 1967.

\bibitem{Bass68}
H.\ Bass, \textit{Algebraic K-Theory}, W.\,A.\ Benjamin, New York, 1968.

\bibitem{Lurie17}
J.\ Lurie, \textit{Higher Algebra}, preprint, 2017.
\url{https://www.math.ias.edu/~lurie/papers/HA.pdf}

\bibitem{Scholze19}
P.\ Scholze, \textit{Lectures on Condensed Mathematics}, lecture notes,
University of Bonn, 2019.

\bibitem{ClausenScholze22}
D.\ Clausen and P.\ Scholze, \textit{Condensed Mathematics and Complex Geometry},
lecture notes, 2022.

\bibitem{LodayVallette12}
J.-L.\ Loday and B.\ Vallette, \textit{Algebraic Operads},
Grundlehren der mathematischen Wissenschaften, vol.~346, Springer, 2012.

\bibitem{MilnorStasheff74}
J.\,W.\ Milnor and J.\,D.\ Stasheff, \textit{Characteristic Classes},
Annals of Mathematics Studies, vol.~76, Princeton University Press, 1974.

\bibitem{Milnor63}
J.\,W.\ Milnor, \textit{Morse Theory},
Annals of Mathematics Studies, vol.~51, Princeton University Press, 1963.

\bibitem{Schwarz93}
M.\ Schwarz, \textit{Morse Homology},
Progress in Mathematics, vol.~111, Birkh\"auser, 1993.

\bibitem{Floer88}
A.\ Floer, Morse theory for Lagrangian intersections,
\textit{J.\ Differential Geom.}\ \textbf{28} (1988), 513--547.

\bibitem{McDuffSalamon12}
D.\ McDuff and D.\ Salamon, \textit{J-Holomorphic Curves and Symplectic Topology},
2nd ed., Colloquium Publications, vol.~52, American Mathematical Society, 2012.

\bibitem{McDuffSalamon17}
D.\ McDuff and D.\ Salamon, \textit{Introduction to Symplectic Topology},
3rd ed., Oxford Graduate Texts in Mathematics, Oxford University Press, 2017.

\bibitem{Geiges08}
H.\ Geiges, \textit{An Introduction to Contact Topology},
Cambridge Studies in Advanced Mathematics, vol.~109, Cambridge University Press, 2008.

\bibitem{Mathlib}
The mathlib Community, The Lean Mathematical Library,
\textit{Proceedings of CPP 2020}, 367--381, ACM, 2020.

\bibitem{Connes94}
A.\ Connes, \textit{Noncommutative Geometry}, Academic Press, 1994.

\bibitem{Weibel94}
C.\,A.\ Weibel, \textit{An Introduction to Homological Algebra},
Cambridge Studies in Advanced Mathematics, vol.~38, Cambridge University Press, 1994.

\bibitem{GelfandManin03}
S.\,I.\ Gelfand and Yu.\,I.\ Manin, \textit{Methods of Homological Algebra},
2nd ed., Springer, 2003.

\bibitem{McCleary01}
J.\ McCleary, \textit{A User's Guide to Spectral Sequences},
2nd ed., Cambridge Studies in Advanced Mathematics, vol.~58,
Cambridge University Press, 2001.

\bibitem{Hartshorne77}
R.\ Hartshorne, \textit{Algebraic Geometry},
Graduate Texts in Mathematics, vol.~52, Springer, 1977.

\bibitem{Atiyah88}
M.\,F.\ Atiyah, Topological quantum field theories,
\textit{Publ.\ Math.\ IHES}\ \textbf{68} (1988), 175--186.

\bibitem{Kock04}
J.\ Kock, \textit{Frobenius Algebras and 2D Topological Quantum Field Theories},
London Mathematical Society Student Texts, vol.~59,
Cambridge University Press, 2004.

\bibitem{DonaldsonKronheimer90}
S.\,K.\ Donaldson and P.\,B.\ Kronheimer, \textit{The Geometry of Four-Manifolds},
Oxford Mathematical Monographs, Oxford University Press, 1990.

\bibitem{BridsonHaefliger99}
M.\,R.\ Bridson and A.\ Haefliger, \textit{Metric Spaces of Non-Positive Curvature},
Grundlehren der mathematischen Wissenschaften, vol.~319, Springer, 1999.

\bibitem{Serre80}
J.-P.\ Serre, \textit{Trees}, Springer, 1980.

\bibitem{HartshorneDeformation10}
R.\ Hartshorne, \textit{Deformation Theory},
Graduate Texts in Mathematics, vol.~257, Springer, 2010.

\bibitem{FontainePeriods82}
J.-M.\ Fontaine, Sur certains types de repr\'esentations $p$-adiques du groupe de Galois d'un corps local,
\textit{Ann.\ of Math.}\ \textbf{115} (1982), 529--577.

\bibitem{Voevodsky00}
V.\ Voevodsky, Triangulated categories of motives over a field,
in \textit{Cycles, Transfers, and Motivic Homology Theories},
Annals of Mathematics Studies, vol.~143, 188--238, 2000.

\bibitem{MorelVoevodsky99}
F.\ Morel and V.\ Voevodsky, $\mathbb{A}^1$-homotopy theory of schemes,
\textit{Publ.\ Math.\ IHES}\ \textbf{90} (1999), 45--143.

\bibitem{Conway90}
J.\,B.\ Conway, \textit{A Course in Functional Analysis},
2nd ed., Graduate Texts in Mathematics, vol.~96, Springer, 1990.

\bibitem{ReedSimon80}
M.\ Reed and B.\ Simon, \textit{Methods of Modern Mathematical Physics},
4 vols., Academic Press, 1972--1979.

\bibitem{Perelman02}
G.\ Perelman, The entropy formula for the Ricci flow and its geometric applications,
arXiv:math/0211159, 2002.

\bibitem{CoxKatz99}
D.\,A.\ Cox and S.\ Katz, \textit{Mirror Symmetry and Algebraic Geometry},
Mathematical Surveys and Monographs, vol.~68,
American Mathematical Society, 1999.
