% ============================================================================
% The Calculus of Computational Paths — Part I: Foundations
% ============================================================================
\documentclass[11pt,a4paper]{report}

% --- Encoding & Fonts ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}

% --- Mathematics ---
\usepackage{amsmath,amssymb,amsthm,mathtools}
\usepackage{stmaryrd}        % \llbracket, \rrbracket
\usepackage{bm}

% --- Cross-references & Hyperlinks ---
\usepackage[colorlinks=true,linkcolor=blue!60!black,citecolor=green!50!black,urlcolor=blue!70!black]{hyperref}
\usepackage[capitalize,noabbrev]{cleveref}

% --- Layout ---
\usepackage[margin=1in]{geometry}
\usepackage{enumitem}

% --- Diagrams ---
\usepackage{tikz-cd}

% ============================================================================
% Theorem-like environments
% ============================================================================
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{notation_def}[theorem]{Notation}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% ============================================================================
% Custom operators and notation
% ============================================================================
\DeclareMathOperator{\refl}{refl}
\DeclareMathOperator{\symop}{symm}
\DeclareMathOperator{\ofEq}{ofEq}
\DeclareMathOperator{\toEq}{toEq}
\DeclareMathOperator{\tr}{transport}
\DeclareMathOperator{\apd}{apd}
\DeclareMathOperator{\id}{id}

\DeclareMathOperator{\congrArgOp}{congrArg}
\DeclareMathOperator{\mapLeft}{mapLeft}
\DeclareMathOperator{\mapRight}{mapRight}
\DeclareMathOperator{\mapTwo}{map_2}
\DeclareMathOperator{\prodMk}{prodMk}
\DeclareMathOperator{\fst}{fst}
\DeclareMathOperator{\snd}{snd}
\DeclareMathOperator{\inlOp}{inl}
\DeclareMathOperator{\inrOp}{inr}
\DeclareMathOperator{\sigmaMk}{sigmaMk}
\DeclareMathOperator{\sigmaFst}{sigmaFst}
\DeclareMathOperator{\sigmaSnd}{sigmaSnd}
\DeclareMathOperator{\lamCongr}{lamCongr}
\DeclareMathOperator{\app}{app}

\DeclareMathOperator{\substL}{substL}
\DeclareMathOperator{\substR}{substR}
\DeclareMathOperator{\normalize}{normalize}

% Short-hand
\newcommand{\Path}{\operatorname{Path}}
\newcommand{\Step}{\operatorname{Step}}
\newcommand{\Rw}{\operatorname{Rw}}
\newcommand{\RwEq}{\operatorname{RwEq}}
\newcommand{\PathQuot}{\operatorname{PathQuot}}
\newcommand{\Context}{\operatorname{Context}}
\newcommand{\BiContext}{\operatorname{BiContext}}
\newcommand{\DepContext}{\operatorname{DepContext}}
\newcommand{\Derivation}{\operatorname{D}}

% Rewrite arrows
\newcommand{\rew}{\mathbin{\triangleright}}         % single step
\newcommand{\rews}{\mathbin{\triangleright^{*}}}    % multi-step
\newcommand{\rweq}{\mathbin{\approx}}               % rewrite equality

% Misc
\newcommand{\comp}{\mathbin{\cdot}}
\newcommand{\inv}[1]{{#1}^{-1}}
\newcommand{\ofEqfn}[1]{\ofEq(#1)}
\newcommand{\Nat}{\mathbb{N}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\Sort}{\mathsf{Sort}}
\newcommand{\Type}{\mathsf{Type}}
\newcommand{\Prop}{\mathsf{Prop}}
\newcommand{\List}{\operatorname{List}}
\newcommand{\Eq}{\operatorname{Eq}}
\newcommand{\IdA}{\operatorname{Id}}

% ============================================================================
\title{\textbf{The Calculus of Computational Paths}\\[6pt]
  \large Part I: Foundations}

\author{
  Arthur Ferreira Ramos\thanks{Centro de Informática, Universidade Federal de Pernambuco, Recife, Brazil.}
  \and
  Ruy J.\,G.\,B.\ de Queiroz\footnotemark[1]
  \and
  Anjolina G.\ de Oliveira\footnotemark[1]
}
\date{\today}

% ============================================================================
\begin{document}
\maketitle

\begin{abstract}
We develop a comprehensive mathematical theory of \emph{computational paths}---a
formalism in which propositional equalities between elements of a type carry
explicit rewrite traces recording the sequence of elementary steps by which they
were derived. A computational path from $a$ to~$b$ in a type~$A$ is a pair
$(s,\pi)$ where $\pi : a =_A b$ is a propositional equality and $s$ is a finite
list of elementary rewrite steps. The space of computational paths is equipped
with a confluent, terminating rewrite system whose 76~rules are organized into
eight groups covering path algebra, type-former $\beta/\eta$-rules, transport,
contexts, and structural closure. We prove that the quotient by rewrite equality
recovers the standard identity type as a strict groupoid, while the
non-quotiented path space forms a weak groupoid. Most significantly, we establish
that the tower of iterated derivation cells forms a weak $\omega$-groupoid in the
sense of Batanin--Leinster, with contractibility beginning at dimension~3---the
critical threshold that preserves non-trivial fundamental groups. The entire
development is formalized in Lean~4 (302~files, 60,860~lines, 1,710~definitions,
885~theorems).
\end{abstract}

\tableofcontents

% ============================================================================
\input{chapters/ch1_introduction}
\input{chapters/ch2_basic_constructions}
\input{chapters/ch3_rewrite_system}
\input{chapters/ch4_groupoid}
\input{chapters/ch5_higher_dimensional}
% ============================================================================

\bibliographystyle{alpha}
\begin{thebibliography}{DQOR18}

\bibitem[BG11]{vdBG11}
B.~van~den~Berg and R.~Garner.
\newblock Types are weak $\omega$-groupoids.
\newblock {\em Proc.\ London Math.\ Soc.}, 102(2):370--394, 2011.

\bibitem[BDJ95]{BaezDolan95}
J.~Baez and J.~Dolan.
\newblock Higher-dimensional algebra and topological quantum field theory.
\newblock {\em J.~Math.\ Phys.}, 36(11):6073--6105, 1995.

\bibitem[BCH14]{BeCH14}
M.~Bezem, T.~Coquand, and S.~Huber.
\newblock A model of type theory in cubical sets.
\newblock In {\em 19th International Conference on Types for Proofs and Programs (TYPES 2013)}, LIPIcs~26, pages 107--128, 2014.

\bibitem[DQOR16]{DQOR16}
R.\,J.\,G.\,B.~de Queiroz, A.\,G.~de Oliveira, and A.\,F.~Ramos.
\newblock Propositional equality, identity types, and direct computational paths.
\newblock {\em South Amer.\ J.\ Logic}, 2(2):245--296, 2016.

\bibitem[EH62]{EckHil62}
B.~Eckmann and P.\,J.~Hilton.
\newblock Group-like structures in general categories.\ {I}.
\newblock {\em Math.\ Annalen}, 145:227--255, 1962.

\bibitem[Hat02]{Hatcher02}
A.~Hatcher.
\newblock {\em Algebraic Topology}.
\newblock Cambridge University Press, 2002.

\bibitem[Lei04]{Leinster04}
T.~Leinster.
\newblock {\em Higher Operads, Higher Categories}.
\newblock London Math.\ Soc.\ Lecture Note Ser.~298. Cambridge University Press, 2004.

\bibitem[Lum10]{Lumsdaine10}
P.\,L.~Lumsdaine.
\newblock Weak $\omega$-categories from intensional type theory.
\newblock {\em Logical Methods in Computer Science}, 6(3):1--19, 2010.

\bibitem[May99]{May99}
J.\,P.~May.
\newblock {\em A Concise Course in Algebraic Topology}.
\newblock University of Chicago Press, 1999.

\bibitem[New42]{Newman42}
M.\,H.\,A.~Newman.
\newblock On theories with a combinatorial definition of ``equivalence''.
\newblock {\em Annals of Math.}, 43(2):223--243, 1942.

\bibitem[RDQO18]{RDQO18}
A.\,F.~Ramos, R.\,J.\,G.\,B.~de Queiroz, and A.\,G.~de Oliveira.
\newblock Explicit computational paths.
\newblock {\em South Amer.\ J.\ Logic}, 4(2):441--484, 2018.

\bibitem[UF13]{HoTTBook}
The {Univalent Foundations Program}.
\newblock {\em Homotopy Type Theory: Univalent Foundations of Mathematics}.
\newblock Institute for Advanced Study, 2013.

\end{thebibliography}

\end{document}
